# Pair Correlation Function

```{r include = FALSE}
library(spatstat)
```


## Motivation

In the previous section, we introduced the second-order product density
$\lambda^{(2)}(u, v)$, which describes the joint occurrences of points at two 
locations.

However, this quantity is difficult to interpret directly:

- its magnitude depends on the intensity $\ambda(u)$,
- it is not scale invariant
- and it does not isolate pure interaction between points.

To address this, we introduct a normalized version.

## Definition

The pair correlation function (PCF) is defined as

$$
g(u, v) = \frac{\lambda^{(2)}(u, v)}{\lambda(u) \lambda(v)}, \quad u \neq v
$$
## Interpretation

The PCF compares the observed joint occurrence of points at $u$ and $v$, to what
would be expected under independence.

In particular:

- $g(u, v) = 1 \Longrightarrow$ no interaction (Poisson-like)

- $g(u, v) > 1 \Longrightarrow$ clustering (points occur together more often 
than expected)

- $g(u, v) > 1 \Longrightarrow$ repulsion (points inhibit each other)

The PCF removes the effect of intensity. It isolates the pure spatial ineraction 
between points.

## Relation to the Poisson process

For a Poisson process, we have:

$$
\lambda^{(2)}(u, v) = \lambda(u) \lambda(v)
$$

and hence:

$$
g(u, v) = 1
$$

Thus, the PCF can be interpreted as a measure of deviation from complete spatial
randomness.

## Interpretation as a ratio

Another useful way to interpret the PCF is:

$$
g(u,v) \approx \frac{\text{probability of observing a pair at } (u,v)}
{\text{probability under independence}}
$$

So:

- $g(u,v) = 2$ means that pairs occur twice as often as expected

- $g(u, v) = 0.5$ means that pairs occur half as often as expected

## Distance based form

In many situations, interaction depends only on the distance between points.

In such cases, we write:

$$
r = \|u - v\|
$$

and express the PCF as a function of distance, $g(r)$.

This reduces the problem from studying interactions between all pairs of 
locations to understanding how interaction varies with separation distance.

This $g(r)$ form is most commonly used in practice.

## Examples (conceptual)

An example of a clustered process:

```{r}
par(mfrow = c(1, 2))
plot(longleaf)
plot(pcf(longleaf))
```

And an example of a regular process:

```{r}
par(mfrow = c(1, 2))
plot(hamster)
plot(pcf(hamster))
```

## Summary

- The PCF is defined as:
  $$
  g(u,v) = \frac{\lambda^{(2)}(u,v)}{\lambda(u)\lambda(v)}.
  $$
- It measures deviation from independence.
- Values of:
  - $g > 1$ indicate clustering,
  - $g < 1$ indicate repulsion.
- When expressed as $g(r)$, it describes how interaction varies with distance.

## Next Step

We now explore another second-order summary: the $K$-function, which 
aggregates interaction over distance and provides a complementary perspective.


















