```{r include = FALSE}
devtools::load_all()
library(spatstat)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(patchwork)
```

# Implications for Statistical Inference

## Overview

The previous sections showed that LGCP and CSCP models can produce nearly 
identical pair correlation functions when their parameters are suitably chosen.

This section investigates whether this similarity carries through to statistical 
inference.

In particular, we examine whether minimum contrast estimation based on the pair 
correlation function leads to similar conclusions under the two models.

For each simulated dataset, we fit both LGCP and CSCP models, and compare the 
resulting estimates and fitted second-order structure.

This allows us to assess whether the two models produce consistent descriptions of 
clustering when applied to the same data.

## Key quantities of interest

The LGCP and CSCP models use different functional forms for their pair 
correlation functions:

- CSCP:

$$
g_{CSCP}(r; \lambda, \phi, s_C) = 1 + \phi \exp\left(-\frac{2r}{s_C}\right),
$$

- LGCP:

$$
g_{LGCP}(r; \lambda, \phi, s_L) = (1 + \phi)^{\exp(-r/s_L)}.
$$

As a result, their scale parameters $s_C$ and $s_L$ are not directly comparable, 
even when the models produce nearly identical curves.

To enable meaningful comparison, we focus on two quantities:

**1. Clustering strength**

$$
\phi = g(0) - 1
$$

This is common to both models and directly comparable.

**2. Half-decay distance**

We define the half-decay distance $r_{1/2}$ as the distance at which the excess 
clustering halves:

$$
g(r_{1/2}) - 1 = \frac{\phi}{2}
$$

For the two models this gives:

- CSCP:

$$
r_{1/2}^{(C)} = \frac{s_C}{2}\log(2)
$$

- LGCP:

$$
r_{1/2}^{(L)} = s_L\log(\frac{\log(1 + \phi)}{\log(1 + \phi / 2)})
$$

This provides a model-independent measure of clustering scale, allowing 
meaningful comparison between models.

## Experimental setup

We consider the following procedure.

For each parameter setting $(\lambda, \phi, s)$:

1. Simulate $n_{\text{rep}} = 200$ point patterns from:
- a CSCP model
- an LGCP model

2. For each simulated dataset:
- estimate the pair correlation function $\hat{g}(r)$
- fit both models (LGCP and CSCP) using minimum contrast

3. For each fit, record:
- estimated clustering strength $\hat{\phi}$
- fitted scale parameter ($\hat{s}_C$ or $\hat{s}_L$)
- derived half-decay distance $\hat{r}_{1/2}$
- fitted pair correlation function

## Expected behaviour

If second-order structure alone is insufficient to distinguish between the models, we expect:

- both models to produce similar fitted PCFs on the same dataset
- both models to yield similar estimates of clustering strength
- raw scale parameters $\hat{s}_C$ and $\hat{s}_L$ to differ
- but derived scale summaries such as $\hat{r}_{1/2}$ to be more closely aligned

## Single dataset illustration

Some helpers for calculating $r_{1/2}$:

```{r}
#| code-fold: true
half_decay_cscp <- function(phi, s) {
  (s / 2) * log(2)
}

half_decay_lgcp <- function(phi, s) {
  s * log(log(1 + phi) / log(1 + phi/2))
}
```

And illustrating:

```{r echo = FALSE}
set.seed(1)

W <- spatstat.geom::owin(c(0, 1), c(0, 1))

true <- list(lambda = 300, phi = 2, s = 0.05)

fit_args <- list(
  start = list(phi = 0.8, s = 0.06),
  bw = 0.02,
  correction = "isotropic",
  divisor = "a",
  zerocor = "JonesFoster"
)

# simulate
X <- sim_cscp(W, lambda = true$lambda, phi = true$phi, s = true$s)$pp

# fit both
fit_c <- fit_cscp_mc(X, bw = fit_args$bw, correction = "isotropic", divisor = "a", zerocor = "JonesFoster")
fit_l <- fit_lgcp_mc(X, bw = fit_args$bw, correction = "isotropic", divisor = "a", zerocor = "JonesFoster")

# empirical PCF
pcf_obs <- spatstat.explore::pcf(X, correction = "isotropic", divisor = "a", zerocor = "JonesFoster", bw = fit_args$bw)

r <- pcf_obs$r

g_c_fit <- pcf_cscp_theoretical(r, lambda = true$lambda, phi = fit_c$par[["phi"]], s = fit_c$par[["s"]])
g_l_fit <- pcf_lgcp_theoretical(r, lambda = true$lambda, phi = fit_l$par[["phi"]], s = fit_l$par[["s"]])

plot(r, pcf_obs$iso, type = "l", lwd = 2,
     main = "CSCP data: both fits vs truth",
     ylab = "g(r)", xlab = "r", ylim = range(c(1, pcf_obs$iso, g_c_fit, g_l_fit), na.rm = TRUE))

lines(r, g_c_fit, col = "blue", lwd = 2)
lines(r, g_l_fit, col = "red", lwd = 2)
lines(r, pcf_cscp_theoretical(r, lambda = true$lambda, phi = true$phi, s = true$s), col = "darkgreen", lwd = 2)
abline(h = 1.0, lty = "dashed", lwd = 1)


legend("topright",
       legend = c("Empirical", "CSCP fit", "LGCP fit", "True PCF"),
       col = c("black", "blue", "red", "darkgreen"),
       lwd = 2)
```

:::callout-note
You can find my:

- Minimum contrast fitting methods [here](https://github.com/JosephReps/cscp/blob/main/R/fit.R)
- CSCP / LGCP simulation methods [here](https://github.com/JosephReps/cscp/blob/main/R/sim.R)
:::

## CSCP-generated data

We first consider datasets generated from the CSCP model.

For each parameter setting $(\lambda, \phi, s)$, we:

1. simulate $n_{\text{rep}}$ point patterns from a CSCP model;
2. estimate the empirical pair correlation function $\hat g(r)$;
3. fit both CSCP and LGCP models using minimum contrast;
4. record the estimated clustering strength $\hat\phi$, the fitted scale 
parameter, and the derived half-decay distance $\hat r_{1/2}$.

We then compare the resulting estimates across the two fitted models.

To examine performance across a range of clustering regimes, we fix 
$\lambda = 300$, and consider

$$
\phi \in \{0.75,\,1.5,\,2.0\}, \qquad s \in \{0.03,\,0.05,\,0.08\}.
$$
These values represent moderate to strong clustering (relative to the CSCP model) 
over short, intermediate, and longer spatial ranges.

:::callout-note
Originally I wanted to use $\phi \in \{0.25, 1, 2\}$, however for $\phi = 0.25$,
the estimation is very unstable so I decided to just use these new values.

Might've been better if I just used $\{1, 2\}$, but I wanted nice 3x3 panels.

Additionally, I chose $\lambda = 300$, as I wanted the non-parametric PCF 
estimator to perform well, so our results are not "muddied" by the poor empirical
fit.

Bandwidth chosen as a value which has worked well in my experience. I think the 
proper thing to do would be to run some type of CV to find the best value? And
importantly, we use the same bandwidth over each regime.

And the other settings are either matched to the ones you guys use in the paper
on the new PCF estimator, or the default spatstat.

Are these sensible choices?
:::

```{r echo = FALSE}
scenarios <- expand.grid(
  phi = c(0.75, 1.5, 2.0),
  s   = c(0.03, 0.05, 0.08)
)

r_grid <- seq(0, 0.25, length.out = 400)

pcf_true_df <- purrr::map_dfr(seq_len(nrow(scenarios)), function(i) {
  phi_i <- scenarios$phi[i]
  s_i   <- scenarios$s[i]

  data.frame(
    r = r_grid,
    phi = factor(phi_i, levels = c(0.75, 1.5, 2.0)),
    s = factor(s_i, levels = c(0.03, 0.05, 0.08)),
    g = pcf_cscp_theoretical(r_grid, lambda = 250, phi = phi_i, s = s_i)
  )
})

ggplot(pcf_true_df, aes(x = r, y = g)) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 0.3) +
  geom_line(linewidth = 0.9) +
  facet_grid(phi ~ s, labeller = label_both) +
  labs(
    title = "True CSCP pair correlation functions across simulation regimes",
    x = "Distance r",
    y = "g(r)"
  ) +
  theme_minimal()
```

These curves illustrate the range of second-order structures considered in the 
simulation study. In particular, increasing $\phi$ controls the *magnitude* of 
clustering, while increasing $s$ controls the spatial extent over which 
clustering persists (don't really think this comment is necessary, at this point
should be fairly clear the roles which $\phi$ and $s$ play).

```{r warning = FALSE, cache = TRUE}
#| code-fold: true
n_rep <- 20

results_list <- list()
counter <- 1

for (j in 1:nrow(scenarios)) {

  phi_true <- scenarios$phi[j]
  s_true   <- scenarios$s[j]

  for (i in 1:n_rep) {

    X <- sim_cscp(W, lambda = 300, phi = phi_true, s = s_true)$pp

    fit_c <- fit_cscp_mc(
      X,
      bw = fit_args$bw,
      correction = "isotropic",
      divisor = "a",
      zerocor = "JonesFoster"
    )

    fit_l <- fit_lgcp_mc(
      X,
      bw = fit_args$bw,
      correction = "isotropic",
      divisor = "a",
      zerocor = "JonesFoster"
    )

    results_list[[counter]] <- data.frame(
      generator = "CSCP",
      phi_true = phi_true,
      s_true = s_true,
      r_half_true = half_decay_cscp(phi_true, s_true),
      rep = i,
      fit_model = "CSCP",
      phi_hat = fit_c$par[["phi"]],
      s_hat = fit_c$par[["s"]],
      r_half = half_decay_cscp(fit_c$par[["phi"]], fit_c$par[["s"]])
    )
    counter <- counter + 1

    results_list[[counter]] <- data.frame(
      generator = "CSCP",
      phi_true = phi_true,
      s_true = s_true,
      r_half_true = half_decay_cscp(phi_true, s_true),
      rep = i,
      fit_model = "LGCP",
      phi_hat = fit_l$par[["phi"]],
      s_hat = fit_l$par[["s"]],
      r_half = half_decay_lgcp(fit_l$par[["phi"]], fit_l$par[["s"]])
    )
    counter <- counter + 1
  }
}

results <- dplyr::bind_rows(results_list)
```

We first compare the fitted models visually through the distribution of 
estimated clustering strength and half-decay distance across simulations.

```{r echo = FALSE}
ggplot(results, aes(x = fit_model, y = phi_hat, fill = fit_model)) +
  geom_boxplot(outliers = FALSE) +
  geom_hline(aes(yintercept = phi_true),
             linetype = "dashed",
             colour = "black",
             linewidth = 0.4) +
  facet_grid(phi_true ~ s_true, labeller = label_both) +
  labs(
    title = "Estimated clustering strength (CSCP data)",
    x = "Fitted model",
    y = expression(hat(phi))
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r echo = FALSE}
ggplot(results, aes(x = fit_model, y = r_half, fill = fit_model)) +
  geom_boxplot(outliers = FALSE) +
  geom_hline(aes(yintercept = r_half_true),
             linetype = "dashed",
             colour = "black",
             linewidth = 0.4) +
  facet_grid(phi_true ~ s_true, labeller = label_both) +
  labs(
    title = "Half-decay distance (CSCP data)",
    x = "Fitted model",
    y = expression(hat(r)[1/2])
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

To quantify these differences, we summarise the agreement between the two fitted 
models across simulations.

```{r}
#| code-fold: true

# This is nasty work, would come back and clean up if this wasn't just rough 
# notes

# True half-decay for CSCP generator
results <- results %>%
  mutate(
    r_half_true = half_decay_cscp(phi_true, s_true),
    phi_error = phi_hat - phi_true,
    r_half_error = r_half - r_half_true
  )

agreement_table <- results %>%
  dplyr::select(generator, phi_true, s_true, rep, fit_model, phi_hat, r_half) %>%
  tidyr::pivot_wider(
    names_from = fit_model,
    values_from = c(phi_hat, r_half)
  ) %>%
  dplyr::mutate(
    delta_phi = abs(phi_hat_CSCP - phi_hat_LGCP),
    delta_r_half = abs(r_half_CSCP - r_half_LGCP),
    rel_delta_phi = delta_phi / phi_true,
    rel_delta_r_half = delta_r_half / half_decay_cscp(phi_true, s_true)
  ) %>%
  dplyr::group_by(generator, phi_true, s_true) %>%
  dplyr::summarise(
    mean_delta_phi = mean(delta_phi, na.rm = TRUE),
    mean_delta_r_half = mean(delta_r_half, na.rm = TRUE),
    mean_rel_delta_phi = mean(rel_delta_phi, na.rm = TRUE),
    mean_rel_delta_r_half = mean(rel_delta_r_half, na.rm = TRUE),
    median_rel_delta_phi = median(rel_delta_phi, na.rm = TRUE),
    median_rel_delta_r_half = median(rel_delta_r_half, na.rm = TRUE),
    .groups = "drop"
  )

agreement_table_clean <- agreement_table %>%
  mutate(
    median_rel_delta_phi = 100 * median_rel_delta_phi,
    median_rel_delta_r_half = 100 * median_rel_delta_r_half
  ) %>%
  select(
    phi_true, s_true,
    mean_delta_phi,
    median_rel_delta_phi,
    median_rel_delta_r_half
  )
```

```{r}
#| code-fold: true

variability_table <- results %>%
  dplyr::group_by(generator, phi_true, s_true, fit_model) %>%
  dplyr::summarise(
    sd_phi_hat = sd(phi_hat, na.rm = TRUE),
    sd_r_half = sd(r_half, na.rm = TRUE),
    .groups = "drop"
  )

comparison_table <- agreement_table %>%
  left_join(
    variability_table %>%
      group_by(phi_true, s_true) %>%
      summarise(
        sd_phi = mean(sd_phi_hat),
        sd_r_half = mean(sd_r_half),
        .groups = "drop"
      ),
    by = c("phi_true", "s_true")
  ) %>%
  mutate(
    ratio_phi = mean_delta_phi / sd_phi,
    ratio_r_half = mean_delta_r_half / sd_r_half
  ) %>%
  select(phi_true, s_true, ratio_phi, ratio_r_half)
```

:::callout-note
A small number of the fits produced wildly inaccurate estimates, which I'm assuming
is just due to unreliable PCF estimation in general, rather than anything specific 
to the models I'm fitting.

As such, we focus on the *median* relative differences, rather than the mean, 
reducing the influence of the occasional outlier.

Also note, these tables aren't particularly readable, but I don't want to spend
a bunch of time formatting stuff nicely when I don't even know whether the 
methodolgy of the study is correct.
:::

```{r echo = FALSE}
agreement_table_clean %>%
  mutate(
    mean_delta_phi = round(mean_delta_phi, 3),
    median_rel_delta_phi = round(median_rel_delta_phi, 1),
    median_rel_delta_r_half = round(median_rel_delta_r_half, 1)
  ) %>%
  knitr::kable(
    caption = "Agreement between LGCP and CSCP fits on CSCP-generated data"
  )
```

The results show that both LGCP and CSCP models yield highly consistent 
second-order inference when fitted to the same CSCP-generated datasets.

Across all parameter regimes, the median relative differences in estimated 
clustering strength $\hat{\phi}$ are small, typically below 15%. The 
corresponding differences in half-decay distance $\hat{r}_{1/2}$ are somewhat 
larger, but remain moderate, generally within 10–30%.

To assess whether these differences are practically meaningful, we compare them 
to the inherent variability of the estimators.

```{r echo = FALSE}
comparison_table %>%
  mutate(
    ratio_phi = round(ratio_phi, 3),
    ratio_r_half = round(ratio_r_half, 3)
  ) %>%
  knitr::kable(
    caption = "Relative magnitude of cross-model differences compared to estimation variability"
  )
```

For both clustering strength and half-decay distance, the difference between the 
two fitted models is consistently smaller than the standard deviation of the 
estimates. In all scenarios, the ratio of cross-model differences to estimation 
variability is well below one.

This indicates that the differences between LGCP and CSCP fits are negligible 
relative to statistical uncertainty. In particular, the two models lead to 
essentially the same conclusions regarding the strength and spatial extent of 
clustering in the data.

From the perspective of second-order inference, the choice between LGCP and CSCP 
does not materially affect the conclusions drawn from the data.

This provides empirical evidence that second-order summaries alone are 
insufficient to distinguish between these model classes in practice.

### Small aside

So we've seen the difference in fitted models is not very substantial, however
I thought it might be nice to see which actually fits the data better, so we 
take a brief look here:

Again, note that because a small number of fits produced extremely large scale 
estimates, leading to unrealistic values of the half-decay distance. To prevent 
these from dominating mean-based summaries, we report trimmed RMSE values, which 
are more robust to such instabilities.

```{r echo = FALSE}
trimmed_rmse <- function(x, truth, trim = 0.05) {
  err <- (x - truth)^2
  err <- sort(err)
  n <- length(err)
  keep <- err[(floor(n * trim) + 1):(ceiling(n * (1 - trim)))]
  sqrt(mean(keep, na.rm = TRUE))
}

performance_table <- results %>%
  group_by(phi_true, s_true, fit_model) %>%
  summarise(
    phi_rmse = trimmed_rmse(phi_hat, phi_true),
    r_half_rmse = trimmed_rmse(r_half, r_half_true),
    .groups = "drop"
  )

performance_table %>%
  mutate(
    phi_rmse = round(phi_rmse, 3),
    r_half_rmse = round(r_half_rmse, 3)
  ) %>%
  knitr::kable(
    caption = "RMSE of parameter estimates under CSCP-generated data"
  )
```

Across all parameter regimes, the RMSE values for the two fitted models are 
very similar. Differences between LGCP and CSCP fits are small in magnitude 
and show no consistent pattern: in some regimes the CSCP fit yields slightly 
lower error, while in others the LGCP fit performs marginally better.

Importantly, even though the data are generated from a CSCP model, fitting an 
LGCP produces estimates with comparable accuracy. This indicates that, when 
inference is based on second-order summaries, neither model has a clear 
advantage in recovering the underlying clustering structure.

:::callout-note
I find deciding on what results to display *really* difficult, is there some 
sort of default / boiler that I can just always use and be confident I'm 
presenting the right information?
:::

:::callout-note
Really should have a better commentary / analysis on how well the models are 
fitting in general. If they both suck at recovering the true parameter values,
do we really care that they are not very dissimilar?
:::

## LGCP-generated data

We now repeat the study with data generated from an LGCP model.

For each parameter setting $(\lambda, \phi, s)$, we:

1. simulate $n_{\text{rep}}$ point patterns from an LGCP model;
2. estimate the empirical pair correlation function $\hat g(r)$;
3. fit both LGCP and CSCP models using minimum contrast;
4. record the estimated clustering strength $\hat\phi$, the fitted scale 
parameter, and the derived half-decay distance $\hat r_{1/2}$.

We then compare the resulting estimates across the two fitted models.

As before, we fix $\lambda = 300$, and consider

$$
\phi \in \{0.75,\,1.5,\,2.0\}, \qquad s \in \{0.03,\,0.05,\,0.08\}.
$$

These values span a range of clustering strengths and spatial scales, allowing us 
to assess whether the same practical non-identifiability appears when the data are 
generated from the LGCP rather than the CSCP.

```{r echo = FALSE}
pcf_true_df_lgcp <- purrr::map_dfr(seq_len(nrow(scenarios)), function(i) {
  phi_i <- scenarios$phi[i]
  s_i   <- scenarios$s[i]

  data.frame(
    r = r_grid,
    phi = factor(phi_i, levels = c(0.75, 1.5, 2.0)),
    s = factor(s_i, levels = c(0.03, 0.05, 0.08)),
    g = pcf_lgcp_theoretical(r_grid, lambda = 250, phi = phi_i, s = s_i)
  )
})

ggplot(pcf_true_df_lgcp, aes(x = r, y = g)) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 0.3) +
  geom_line(linewidth = 0.9) +
  facet_grid(phi ~ s, labeller = label_both) +
  labs(
    title = "True LGCP pair correlation functions across simulation regimes",
    x = "Distance r",
    y = "g(r)"
  ) +
  theme_minimal()
```

```{r cache = TRUE, warning = FALSE}
#| code-fold: true

results_lgcp_list <- list()
counter <- 1

for (j in 1:nrow(scenarios)) {

  phi_true <- scenarios$phi[j]
  s_true   <- scenarios$s[j]

  for (i in 1:n_rep) {

    X <- sim_lgcp(W, lambda = 300, phi = phi_true, s = s_true)$pp

    fit_l <- fit_lgcp_mc(
      X,
      bw = fit_args$bw,
      correction = "isotropic",
      divisor = "a",
      zerocor = "JonesFoster"
    )

    fit_c <- fit_cscp_mc(
      X,
      bw = fit_args$bw,
      correction = "isotropic",
      divisor = "a",
      zerocor = "JonesFoster"
    )

    results_lgcp_list[[counter]] <- data.frame(
      generator = "LGCP",
      phi_true = phi_true,
      s_true = s_true,
      r_half_true = half_decay_lgcp(phi_true, s_true),
      rep = i,
      fit_model = "LGCP",
      phi_hat = fit_l$par[["phi"]],
      s_hat = fit_l$par[["s"]],
      r_half = half_decay_lgcp(fit_l$par[["phi"]], fit_l$par[["s"]])
    )
    counter <- counter + 1

    results_lgcp_list[[counter]] <- data.frame(
      generator = "LGCP",
      phi_true = phi_true,
      s_true = s_true,
      r_half_true = half_decay_lgcp(phi_true, s_true),
      rep = i,
      fit_model = "CSCP",
      phi_hat = fit_c$par[["phi"]],
      s_hat = fit_c$par[["s"]],
      r_half = half_decay_cscp(fit_c$par[["phi"]], fit_c$par[["s"]])
    )
    counter <- counter + 1
  }
}

results_lgcp <- dplyr::bind_rows(results_lgcp_list)
```

We again compare the fitted models visually through the distribution of estimated
clustering strength and half-decay distance across simulations.

```{r echo = FALSE}
ggplot(results_lgcp, aes(x = fit_model, y = phi_hat, fill = fit_model)) +
  geom_boxplot(outliers = FALSE) +
  geom_hline(aes(yintercept = phi_true),
             linetype = "dashed",
             colour = "black",
             linewidth = 0.4) +
  facet_grid(phi_true ~ s_true, labeller = label_both) +
  labs(
    title = "Estimated clustering strength (LGCP data)",
    x = "Fitted model",
    y = expression(hat(phi))
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r echo = FALSE}
ggplot(results_lgcp, aes(x = fit_model, y = r_half, fill = fit_model)) +
  geom_boxplot(outliers = FALSE) +
  geom_hline(aes(yintercept = r_half_true),
             linetype = "dashed",
             colour = "black",
             linewidth = 0.4) +
  facet_grid(phi_true ~ s_true, labeller = label_both) +
  labs(
    title = "Half-decay distance (LGCP data)",
    x = "Fitted model",
    y = expression(hat(r)[1/2])
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

To quantify these differences, we summarise the agreement between the two fitted
models across simulations.

```{r}
#| code-fold: true

results_lgcp <- results_lgcp %>%
  mutate(
    r_half_true = half_decay_lgcp(phi_true, s_true),
    phi_error = phi_hat - phi_true,
    r_half_error = r_half - r_half_true
  )

agreement_table_lgcp <- results_lgcp %>%
  dplyr::select(generator, phi_true, s_true, rep, fit_model, phi_hat, r_half) %>%
  tidyr::pivot_wider(
    names_from = fit_model,
    values_from = c(phi_hat, r_half)
  ) %>%
  dplyr::mutate(
    delta_phi = abs(phi_hat_CSCP - phi_hat_LGCP),
    delta_r_half = abs(r_half_CSCP - r_half_LGCP),
    rel_delta_phi = delta_phi / phi_true,
    rel_delta_r_half = delta_r_half / half_decay_lgcp(phi_true, s_true)
  ) %>%
  dplyr::group_by(generator, phi_true, s_true) %>%
  dplyr::summarise(
    mean_delta_phi = mean(delta_phi, na.rm = TRUE),
    mean_delta_r_half = mean(delta_r_half, na.rm = TRUE),
    mean_rel_delta_phi = mean(rel_delta_phi, na.rm = TRUE),
    mean_rel_delta_r_half = mean(rel_delta_r_half, na.rm = TRUE),
    median_rel_delta_phi = median(rel_delta_phi, na.rm = TRUE),
    median_rel_delta_r_half = median(rel_delta_r_half, na.rm = TRUE),
    .groups = "drop"
  )

agreement_table_lgcp_clean <- agreement_table_lgcp %>%
  mutate(
    median_rel_delta_phi = 100 * median_rel_delta_phi,
    median_rel_delta_r_half = 100 * median_rel_delta_r_half
  ) %>%
  select(
    phi_true, s_true,
    mean_delta_phi,
    median_rel_delta_phi,
    median_rel_delta_r_half
  )
```

```{r}
#| code-fold: true

variability_table_lgcp <- results_lgcp %>%
  dplyr::group_by(generator, phi_true, s_true, fit_model) %>%
  dplyr::summarise(
    sd_phi_hat = sd(phi_hat, na.rm = TRUE),
    sd_r_half = sd(r_half, na.rm = TRUE),
    .groups = "drop"
  )

comparison_table_lgcp <- agreement_table_lgcp %>%
  left_join(
    variability_table_lgcp %>%
      group_by(phi_true, s_true) %>%
      summarise(
        sd_phi = mean(sd_phi_hat),
        sd_r_half = mean(sd_r_half),
        .groups = "drop"
      ),
    by = c("phi_true", "s_true")
  ) %>%
  mutate(
    ratio_phi = mean_delta_phi / sd_phi,
    ratio_r_half = mean_delta_r_half / sd_r_half
  ) %>%
  select(phi_true, s_true, ratio_phi, ratio_r_half)
```

```{r echo = FALSE}
agreement_table_lgcp_clean %>%
  mutate(
    mean_delta_phi = round(mean_delta_phi, 3),
    median_rel_delta_phi = round(median_rel_delta_phi, 1),
    median_rel_delta_r_half = round(median_rel_delta_r_half, 1)
  ) %>%
  knitr::kable(
    caption = "Agreement between LGCP and CSCP fits on LGCP-generated data"
  )
```

```{r echo = FALSE}
comparison_table_lgcp %>%
  mutate(
    ratio_phi = round(ratio_phi, 3),
    ratio_r_half = round(ratio_r_half, 3)
  ) %>%
  knitr::kable(
    caption = "Relative magnitude of cross-model differences compared to estimation variability"
  )
```

The results again show that both LGCP and CSCP models yield highly consistent
second-order inference when fitted to the same datasets.

Across the parameter regimes considered, the differences in estimated clustering
strength $\hat{\phi}$ and half-decay distance $\hat r_{1/2}$ remain small relative
to the overall Monte Carlo variability of the estimators. Although the fitted
models use different functional forms, they continue to provide very similar
descriptions of the clustering present in the data.

Importantly, this behaviour persists even when the data are generated from the
LGCP itself. Thus, the practical similarity observed in the previous section is
not specific to CSCP-generated data, but appears symmetrically in both directions.

This strengthens the conclusion that second-order summaries alone are not
sufficient to distinguish between these model classes in practice.

### Small aside (again)

As before, we can also compare the accuracy of the two fitted models through RMSE.

```{r echo = FALSE}
performance_table_lgcp <- results_lgcp %>%
  group_by(phi_true, s_true, fit_model) %>%
  summarise(
    phi_rmse = trimmed_rmse(phi_hat, phi_true),
    r_half_rmse = trimmed_rmse(r_half, r_half_true),
    .groups = "drop"
  )

performance_table_lgcp %>%
  mutate(
    phi_rmse = round(phi_rmse, 3),
    r_half_rmse = round(r_half_rmse, 3)
  ) %>%
  knitr::kable(
    caption = "RMSE of parameter estimates under LGCP-generated data"
  )
```

Across all parameter regimes, the RMSE values for the two fitted models remain
very similar. Even when the data are generated from an LGCP model, fitting a
CSCP produces estimates with comparable accuracy.

Taken together with the CSCP-generated results, this suggests that minimum
contrast estimation based on the pair correlation function is primarily recovering
shared second-order structure, rather than features specific to either model class.

## Summary and implications

Across the previous sections, we investigated whether the theoretical similarity 
between the LGCP and CSCP pair correlation functions carries through to practical 
statistical inference.

We first showed that, for suitable parameter choices, the two models can produce 
very similar pair correlation functions. In particular, when expressed in terms of 
the clustering strength $\phi = g(0) - 1$ and a characteristic spatial scale, the 
difference between the two families is often negligible over the range of distances 
relevant for estimation.

We then examined whether this similarity affects inference by conducting simulation 
studies in both directions:

- data generated from a CSCP, fitted with both CSCP and LGCP models;
- data generated from an LGCP, fitted with both LGCP and CSCP models.

In both cases, the results were remarkably consistent. The two fitted models 
produced:

- very similar estimates of clustering strength $\hat{\phi}$;
- very similar estimates of the half-decay distance $\hat{r}_{1/2}$;
- visually indistinguishable fitted pair correlation functions.

Moreover, the differences between the models were small relative to the overall 
Monte Carlo variability of the estimators. In other words, fitting the "wrong" model 
did not meaningfully degrade second-order inference.

### What does this mean?

At a practical level, these results suggest that inference based on second-order 
summaries, such as minimum contrast estimation using the pair correlation 
function, is largely driven by the *shape of the PCF*, rather than the specific 
underlying model.

Since the LGCP and CSCP are capable of generating very similar PCFs, they also 
lead to very similar second-order inference, even under model misspecification.

This implies that, in many realistic settings, it may be difficult to distinguish 
between these model classes using second-order statistics alone.

### Implications for modelling

These findings have two important implications.

First, care should be taken when interpreting fitted Cox process models based only 
on second-order summaries. A good fit to the empirical PCF does not necessarily 
provide strong evidence in favour of one model over another, if multiple models can 
reproduce the same second-order structure.

Second, model choice in this context should be guided not only by goodness-of-fit 
at the level of the PCF, but also by:

- interpretability of the latent structure,
- computational considerations,
- and the specific scientific questions of interest.

:::callout-note
Some thoughts I have:

I really think the two most important pieces of information that need to be 
provided here are (1) that the models actually perform well and (2) that the
difference between the models is smaller than the variation in the empirical
PCF estimator. 

Right now I don't think I do a very good job of framing this.

I will continue to work on this, but regardless, it seems we are not going to be
very successful in distinguishing between the two models using PCF-based methods.
:::

### Future directions

The results of this study naturally raise the question of how the models might be 
distinguished in practice.

One possible direction is to consider features beyond second-order structure. 
For example:

- higher-order interactions,
- conditional or Palm distributions,
- or likelihood-based approaches that incorporate more of the underlying model 
  structure.

Alternatively, one could investigate whether differences in the marginal 
distribution of the intensity field, or in extreme clustering behaviour, provide 
useful distinguishing features.

Exploring these directions may help to identify settings in which the LGCP and CSCP 
can be meaningfully separated, and where model choice has a more substantial impact 
on inference.






















