```{r include = FALSE}
devtools::load_all()
library(spatstat)
```

# Second-order Structure Comparison to LGCPs

An important question we have been investigating is whether a point pattern 
generated by a chi-square Cox process (CSCP) can be distinguished from one 
generated by a log-Gaussian Cox process (LGCP) (single component).

Since the two models arise from different latent constructions, we might expect
them to produce visibly different realizations. A natural place to begin then, 
is with a simple visual comparison: are the differences between the two 
processes large enough that they can be noticed directly from their intensity 
surfaces or point patterns?

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=13}
set.seed(333)

W <- spatstat.geom::owin(c(0, 1), c(0, 2))

lambda <- 500
phi    <- 2
s      <- 0.1
df     <- 1

lgcp_sim <- sim_lgcp(W, lambda = lambda, phi = phi, s = s)
cscp_sim <- sim_cscp(W, lambda = lambda, phi = phi, s = s, df = df)

zlim_shared <- range(
  c(lgcp_sim$intensity$v, cscp_sim$intensity$v),
  finite = TRUE
)

lay <- rbind(
  c(1, 0, 2),
  c(3, 4, 5)
)

layout(
  lay,
  widths  = c(1, 0.025, 1),
  heights = c(0.05, 1)
)

par(
  mar  = c(0, 0, 0, 0),
  oma  = c(0, 0, 0, 0),
  xaxs = "i",
  yaxs = "i"
)

# titles
plot.new()
text(0.5, 0.2, "LGCP", font = 2, cex = 2)

plot.new()
text(0.5, 0.2, "CSCP", font = 2, cex = 2)

# left surface
plot(
  lgcp_sim$intensity,
  main   = "",
  ribbon = FALSE,
  axes   = FALSE,
  box    = FALSE,
  zlim   = zlim_shared
)

# spacer
plot.new()

# right surface
plot(
  cscp_sim$intensity,
  main   = "",
  ribbon = FALSE,
  axes   = FALSE,
  box    = FALSE,
  zlim   = zlim_shared
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=13, dpi=640, dev="ragg_png", fig.retina=2}
lay <- rbind(
  c(1, 0, 2),
  c(3, 4, 5)
)

layout(
  lay,
  widths  = c(1, 0.025, 1),
  heights = c(0.05, 1)
)

par(
  mar  = c(0, 0, 0, 0),
  oma  = c(0, 0, 0, 0),
  xaxs = "i",
  yaxs = "i"
)

# titles
plot.new()
text(0.5, 0.2, "LGCP", font = 2, cex = 2)

plot.new()
text(0.5, 0.2, "CSCP", font = 2, cex = 2)

# left pattern
plot(
  lgcp_sim$pp,
  main = "",
  axes = FALSE,
  box  = FALSE,
  pch  = 16,
  cex  = 1.15
)

# spacer
plot.new()

# right pattern
plot(
  cscp_sim$pp,
  main = "",
  axes = FALSE,
  box  = FALSE,
  pch  = 16,
  cex  = 1.15
)
```

:::callout-note
Will make these nicer later, SO annoying trying to get them to play nice, I 
actually despise the default plotting from spatstat.
:::

When the parameters are matched exactly, the two processes appear slightly
different upon visual inspection of their intensity surfaces (and less so
from the resulting point patterns).

Now obviously visual inspection alone is not sufficient for model comparison in
practice. Instead, it is common to rely on second-order summary statistics,
such as the pair correlation function (PCF), which capture the dependence
structure of the process.

We therefore turn our attention towards a comparison of the theoretical PCFs of 
the two models.

Consider the following plot of 2 processes (LGCP and CSCP), on a 
$[0, 1]\times[0, 1]$ window:

```{r echo = FALSE}
r <- seq(0, 0.5, length.out = 500)

g_lgcp <- pcf_lgcp_theoretical(r, lambda = 1, phi = 1, s = 0.05)
g_cscp <- pcf_cscp_theoretical(r, lambda = 1, phi = 1, s = 0.05, df = 1)

plot(r, g_lgcp, type = "l", lwd = 2, ylim = range(c(g_lgcp, g_cscp)), col = "black", 
     ylab = "g(r)", main = "Theoretical PCFs of LGCP & CSCP (lambda = 1, phi = 1, s = 0.05)")
lines(r, g_cscp, lwd = 2, lty = 2, col = "red")
legend("topright", legend = c("LGCP (s = 0.05)", "CSCP (s = 0.05)"), 
       lty = c(1, 2), lwd = 2, col = c("black", "red"))
```

The two pair correlation functions are clearly different when all parameters are
matched.

However, this comparison is somewhat artificial.

In practice, we obviously do not know the true parameters of the models, and must
estimate them from the observed point pattern. When fitting competing models, each
model is free to select parameter values that best reproduce the observed
structure.

As a result, even if two models differ under identical parameter values, they may
still produce very similar second-order behaviour after fitting.

Importantly, while the parameters $\lambda$ and $\phi = g(0) - 1$ have a
common interpretation across both models, controlling the overall intensity and
the strength of clustering at the origin, the scale parameter $s$ plays a more
subtle role.

In the LGCP, $s$ governs the decay of correlation in the latent Gaussian field,
and leads to a PCF of the form

$$
g(r) = \exp\big(\sigma^2 \exp(-r/s)\big),
$$

whereas in the CSCP, the dependence arises through squared Gaussian fields,
resulting in

$$
g(r) = 1 + \phi \exp(-2r/s).
$$

Thus, although both parameters are referred to as "scale", they do not represent
the same structural feature, and equating them does not lead to a meaningful
comparison of the models.

This suggests a more appropriate comparison: fix the first-order intensity
$\lambda$ and clustering strength $\phi$, and allow the scale parameter to
vary. This reflects the practical setting in which models are fitted to data.

The following plots demonstrate that, under this setup, an LGCP can closely
approximate the PCF of a CSCP by suitably adjusting its scale parameter (and 
vice versa).

```{r echo = FALSE, fig.height=10, fig.width=12}
r <- seq(0, 0.5, length.out = 500)

old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)

par(
  mfrow = c(1, 2),
  mar   = c(4.2, 4.2, 3.2, 1.2), # more room for titles
  oma   = c(0, 0, 0.5, 0), # tiny outer margin
  cex.main = 1.5,
  cex.lab  = 1.15,
  cex.axis = 1.0,
  cex      = 1.0
)

main_txt <- expression("Theoretical PCFs (" * lambda, "==", 1 * ", " * phi, "==", 1 * ")")

### PLOT 1
g_lgcp <- pcf_lgcp_theoretical(r, lambda = 1, phi = 1, s = 0.05)
g_cscp <- pcf_cscp_theoretical(r, lambda = 1, phi = 1, s = 0.05, df = 1)

plot(
  r, g_lgcp,
  type = "l",
  lwd  = 2,
  ylim = range(c(g_lgcp, g_cscp)),
  col  = "black",
  xlab = "r",
  ylab = "g(r)",
  main = main_txt
)

for (scale in seq(0.02, 0.10, length.out = 16)) {
  g_lgcp_ex <- pcf_lgcp_theoretical(r, lambda = 1, phi = 1, s = scale)
  lines(r, g_lgcp_ex, lwd = 0.7, lty = 1, col = "black")
}

lines(r, g_cscp, lwd = 2, lty = 2, col = "red")

legend(
  "topright",
  legend = c("LGCP (s = 0.05)", "CSCP (s = 0.05)"),
  lty    = c(1, 2),
  lwd    = 2,
  col    = c("black", "red"),
  bty    = "n",
  cex    = 1.5
)

### PLOT 2
g_lgcp <- pcf_lgcp_theoretical(r, lambda = 1, phi = 1, s = 0.05)
g_cscp <- pcf_cscp_theoretical(r, lambda = 1, phi = 1, s = 0.05, df = 1)

plot(
  r, g_cscp,
  type = "l",
  lty  = 2,
  lwd  = 2,
  ylim = range(c(g_lgcp, g_cscp)),
  col  = "red",
  xlab = "r",
  ylab = "g(r)",
  main = main_txt
)

for (scale in seq(0.02, 0.10, length.out = 16)) {
  g_cscp_ex <- pcf_cscp_theoretical(r, lambda = 1, phi = 1, s = scale, df = 1)
  lines(r, g_cscp_ex, lwd = 0.7, lty = 2, col = "red")
}

lines(r, g_lgcp, lwd = 2, lty = 1, col = "black")

legend(
  "topright",
  legend = c("LGCP (s = 0.05)", "CSCP (s = 0.05)"),
  lty    = c(1, 2),
  lwd    = 2,
  col    = c("black", "red"),
  bty    = "n",
  cex    = 1.5
)
```

:::callout-note
Title clipping, will fix later. 
:::

And consider a single example, demonstrating a good approximation of LGCP PCF 
with scale = 0.05 by a CSCP PCF with scale = 0.085.

```{r echo = FALSE}
r <- seq(0, 0.5, length.out = 500)

g_lgcp <- pcf_lgcp_theoretical(r, lambda = 1, phi = 1, s = 0.05)
g_cscp <- pcf_cscp_theoretical(r, lambda = 1, phi = 1, s = 0.085, df = 1)

plot(r, g_lgcp, type = "l", lwd = 2, ylim = range(c(g_lgcp, g_cscp)), col = "black", 
     ylab = "g(r)", main = "Theoretical PCFs of LGCP & CSCP (lambda = 1, phi = 1)")
lines(r, g_cscp, lwd = 2, lty = 2, col = "red")
legend("topright", legend = c("LGCP (s = 0.05)", "CSCP (s = 0.085)"), 
       lty = c(1, 2), lwd = 2, col = c("black", "red"))
```

Even after fixing first-order intensity and clustering strength, the models 
retain enough flexibility through their scale parameter to closely match each 
other’s pair correlation functions.

This suggests that the two models may be difficult to distinguish based on
second-order structure alone, even though they arise from fundamentally different
constructions.

This motivates the central question of this chapter:

> To what extent are LGCPs and CSCPs identifiable from their pair correlation
functions?

and perhaps in the future:

> What aspects of the models (if any) allow them to be distinguished in practice?

