```{r include = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(patchwork)
library(spatstat.geom)
library(spatstat.explore)
devtools::load_all()
```

# 🚧 Empirical comparison of PCF families

## Numerical optimisation

To study the approximation problem, we numerically solve

$$
p^*(\phi) \in \arg \min_{p > 0} \sup_{x \in [0,1]} 
\left| \exp(c x^p) - 1 - \phi x \right|,
\quad c = \log(1 + \phi),
$$

for values of $\phi \in [0,2]$.

The supremum over $x$ is approximated using a fine grid over $[0,1]$, and the
minimisation over $p$ is carried out using one-dimensional optimisation.

Code:

```{r}
#| code-fold: true
# Shouldve just used my package functions, but was still under construction
g_cscp <- function(r, phi, s_C) {
  1 + phi * exp(-2 * r / s_C)
}

g_lgcp <- function(r, phi, s_L) {
  (1 + phi)^(exp(-r / s_L))
}

# E(phi, p) = sup_{x in [0,1]} | exp(c x^p) - 1 - phi x |
pcf_error_p <- function(phi, p, x_grid = seq(0, 1, length.out = 4000)) {
  cval <- log(1 + phi)
  max(abs(exp(cval * x_grid^p) - 1 - phi * x_grid))
}

# same thing, but using a scale ratio alpha = s_L / s_C
pcf_error_alpha <- function(phi, alpha, x_grid = seq(0, 1, length.out = 4000)) {
  p <- 1 / (2 * alpha)
  pcf_error_p(phi, p, x_grid = x_grid)
}

# Slope matching - intially thought I would include as a 
# "rule-of-thumb" type thing, but it's too shitty to keep tbh
alpha_slope <- function(phi) {
  ifelse(phi == 0, 1/2, ((1 + phi) * log(1 + phi)) / (2 * phi))
}

p_slope <- function(phi) {
  1 / (2 * alpha_slope(phi))
}

find_optimal_match <- function(phi,
                               x_grid = seq(0, 1, length.out = 4000),
                               alpha_lower = 0.05,
                               alpha_upper = 5) {
  
  obj_alpha <- function(alpha) {
    pcf_error_alpha(phi, alpha, x_grid = x_grid)
  }
  
  opt <- optimize(obj_alpha, interval = c(alpha_lower, alpha_upper))
  
  alpha_opt <- opt$minimum
  err_opt   <- opt$objective
  alpha_h   <- alpha_slope(phi)
  err_h     <- obj_alpha(alpha_h)
  
  tibble(
    phi = phi,
    alpha_opt = alpha_opt,
    alpha_slope = alpha_h,
    p_opt = 1 / (2 * alpha_opt),
    p_slope = p_slope(phi),
    err_opt = err_opt,
    err_slope = err_h
  )
}
```

## Optimal scale relationship

We first examine how the optimal LGCP scale relates to the CSCP scale.

Define the scale ratio

$$
\alpha^*(\phi) = \frac{s_L^*(\phi)}{s_C}.
$$

The figure below shows $\alpha^*(\phi)$ as a function of $\phi$, along with the
slope-matching approximation

$$
\alpha_{\text{slope}}(\phi) = \frac{(1+\phi)\log(1+\phi)}{2\phi}.
$$

:::callout-note
Originally, I wanted to have some sort of closed-form approximation of the best
approximating ratio for a given $\phi$, but it seems less useful than just
making a function which numerically calculates it for you basically instantly.
:::

```{r}
#| code-fold: true

phi_grid <- seq(0.01, 2, length.out = 200)

opt_results <- map_dfr(phi_grid, find_optimal_match)

# opt_results
scale_plot_df <- opt_results %>%
  select(phi, alpha_opt, alpha_slope) %>%
  pivot_longer(cols = c(alpha_opt, alpha_slope),
               names_to = "method",
               values_to = "alpha")

scale_plot_df$method <- factor(
  scale_plot_df$method,
  levels = c("alpha_opt", "alpha_slope"),
  labels = c("Optimal", "Slope-matching")
)

ggplot(scale_plot_df, aes(x = phi, y = alpha, linetype = method)) +
  geom_line(linewidth = 0.9) +
  labs(
    x = expression(phi),
    y = expression(s[L]^"*" / s[C]),
    linetype = NULL
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

## Approximation error

We now examine the minimum discrepancy:

$$
E^*(\phi) := \inf_{p > 0} \sup_{x \in [0,1]} \left| \exp(c x^p) - 1 - \phi x \right|
$$

The figure below shows $E^*(\phi)$ as a function of $\phi$, along with the
discrepancy obtained using the slope-matching approximation.

```{r}
#| code-fold: true

err_plot_df <- opt_results %>%
  select(phi, err_opt, err_slope) %>%
  pivot_longer(cols = c(err_opt, err_slope),
               names_to = "method",
               values_to = "error")

err_plot_df$method <- factor(
  err_plot_df$method,
  levels = c("err_opt", "err_slope"),
  labels = c("Optimal", "Slope-matching")
)

ggplot(err_plot_df, aes(x = phi, y = error, linetype = method)) +
  geom_line(linewidth = 0.9) +
  labs(
    x = expression(phi),
    y = "Minimum discrepancy",
    linetype = NULL
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

:::callout-note
Might be nice to replot this as some type of error relative to the curves. 

For example, maybe relative to the current height of the curve above 1?
:::

## Visual comparison of PCFs

To better understand the approximation, we compare the PCF curves directly
for selected values of $\phi$.

In each panel we show:

- the CSCP PCF (solid line)

- the best-fitting LGCP PCF (dashed line)

```{r}
#| code-fold: true

phi_panel <- c(0.1, 0.8, 1.5, 2.0)
s_C_fixed <- 1
r_grid <- seq(0, 4, length.out = 1000)

panel_match <- map_dfr(phi_panel, find_optimal_match)

pcf_panel_df <- map_dfr(seq_along(phi_panel), function(i) {
  phi_i <- phi_panel[i]
  alpha_i <- panel_match$alpha_opt[panel_match$phi == phi_i]
  s_L_i <- alpha_i * s_C_fixed
  
  tibble(
    r = rep(r_grid, 2),
    g = c(
      g_cscp(r_grid, phi = phi_i, s_C = s_C_fixed),
      g_lgcp(r_grid, phi = phi_i, s_L = s_L_i)
    ),
    model = rep(c("CSCP", "Best LGCP"), each = length(r_grid)),
    phi_lab = paste0("phi = ", phi_i)
  )
})

ggplot(pcf_panel_df, aes(x = r, y = g, colour = model, linetype = model)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ phi_lab, ncol = 2, scales = "fixed") +
  labs(
    x = "r",
    y = "PCF",
    colour = NULL,
    linetype = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

## Sampling variability vs model difference

To assess whether the two models can be distinguished in practice, we compare
the difference between the PCF curves to the sampling variability of the PCF
estimator.

We:

- simulate a CSCP point pattern,
- estimate the PCF,
- construct simulation envelopes,
- and overlay the true CSCP PCF and the best-fitting LGCP PCF.

:::callout-note
Not finished, but you get the idea of what I'm trying to show here, right?
:::

```{r echo = FALSE}
# Example placeholder:
X_cscp <- sim_cscp(lambda = 2, phi = 0.8, s = 1, W = owin(c(0, 10), c(0, 10)))$pp

phi_env <- 0.8
s_C_env <- 1

match_env <- find_optimal_match(phi_env)
s_L_env <- match_env$alpha_opt * s_C_env

# radius grid used in plotting theoretical curves
r_env <- seq(0, 4, length.out = 400)

# empirical pcf estimate
pcf_hat <- pcf(X_cscp, correction = "isotropic", divisor = "a", zerocor = "JonesFoster")

# envelope of pcf under repeated resampling / bootstrap-style replacement
env <- envelope(
  X_cscp,
  fun = pcf,
  nsim = 39,
  correction = "isotropic",
  divisor = "a",
  zerocor = "JonesFoster",
  simulate = expression(rpoispp(intensity(X_cscp), win = Window(X_cscp))),
  verbose = FALSE,
  savefuns = TRUE
)

# plot envelope
plot(env,
     main = "",
     xlab = "r",
     ylab = "PCF",
     lwd = 1)

# overlay empirical estimate again for clarity
lines(pcf_hat$r, pcf_hat$iso, lwd = 2)

# overlay true CSCP and best LGCP approximation
lines(r_env, g_cscp(r_env, phi = phi_env, s_C = s_C_env), lwd = 2, lty = 1)
lines(r_env, g_lgcp(r_env, phi = phi_env, s_L = s_L_env), lwd = 2, lty = 2)

legend(
  "topright",
  legend = c("Empirical PCF", "CSCP truth", "Best LGCP"),
  lty = c(1, 1, 2),
  lwd = c(2, 2, 2),
  bty = "n"
)
```

```{r echo = FALSE}
env_df <- tibble(
  r = env$r,
  lo = env$lo,
  hi = env$hi,
  theo = env$theo,
  obs = env$obs
)

overlay_df <- tibble(
  r = r_env,
  cscp = g_cscp(r_env, phi = phi_env, s_C = s_C_env),
  lgcp = g_lgcp(r_env, phi = phi_env, s_L = s_L_env)
)

ggplot() +
  geom_ribbon(data = env_df, aes(x = r, ymin = lo, ymax = hi), alpha = 0.25) +
  geom_line(data = env_df, aes(x = r, y = obs), linewidth = 0.9) +
  geom_line(data = overlay_df, aes(x = r, y = cscp, linetype = "CSCP truth"), linewidth = 0.9) +
  geom_line(data = overlay_df, aes(x = r, y = lgcp, linetype = "Best LGCP"), linewidth = 0.9) +
  labs(
    x = "r",
    y = "PCF",
    linetype = NULL
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```


## Interpretation

Across all values of $\phi$, the discrepancy between the LGCP and CSCP PCFs can
be made very small by appropriately rescaling the LGCP.

Moreover, the difference between the two PCF curves is typically small relative
to the sampling variability of standard PCF estimators.

This suggests that, in practice, the two models are difficult to distinguish
based on second-order structure alone.

## What’s next?

So far we have shown:

- The PCFs of the LGCP and CSCP can approximate each other closely, and  
- The differences between the two families are small relative to the sampling 
variability of standard PCF estimators.

This is nice, however what we would *like* to claim is:

- Procedures based on second-order summaries, such as minimum contrast 
  estimation, lead to similar conclusions under LGCP and CSCP models.

The next section investigates this empirically by fitting each models to data
generated by both LGCPs and CSCPs, and compares the resulting fits and parameter
estimates.

## Extras 

### Difference between PCFs

To highlight where the two models differ, we plot

$$
g_{LGCP}(r) - g_{CSCP}(r)
$$

for selected values of $\phi$.

This is looking at how different the two curves are as we increase $r$.

```{r}
#| code-fold: true
diff_df <- map_dfr(seq_along(phi_panel), function(i) {
  phi_i <- phi_panel[i]
  alpha_i <- panel_match$alpha_opt[panel_match$phi == phi_i]
  s_L_i <- alpha_i * s_C_fixed
  
  tibble(
    r = r_grid,
    diff = g_lgcp(r_grid, phi = phi_i, s_L = s_L_i) -
      g_cscp(r_grid, phi = phi_i, s_C = s_C_fixed),
    phi_lab = paste0("phi = ", phi_i)
  )
})

ggplot(diff_df, aes(x = r, y = diff)) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ phi_lab, ncol = 2, scales = "fixed") +
  labs(
    x = "r",
    y = expression(g[LGCP](r) - g[CSCP](r))
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
  )
```



