## 🚧 Local Aggregates

🚧 WIP 🚧 WIP🚧 WIP🚧 WIP 🚧 WIP 🚧

Haven't heard about this:

Instead of just the marginal $\Lambda(u)$, instead consider:

$$
\Lambda(B) = \int_B\Lambda(u)\ du
$$

So it's somewhere "inbetween" looking at just the marginal, versus the full joint?

I'm assuming that it is going to be more annoying to estimate properties / structure 
of it from observed point pattern, because it is "containing more information" about
the full joint?

So I guess the justification is that while the LGCP and CSCP may look similar 
pointwise and pairwise, they differ in how intensity accumulates over space?

Z^2 more "spiky" than exp(Z)

:::callout-note

The key point is that two fields can have:

- very similar first-order behaviour,
- very similar second-order summaries,
- even somewhat similar pointwise marginals,

while still differing in how mass accumulates over a small neighbourhood.

That is exactly what $\Lambda(B)$ explores.

Or otherwise stated nicely by the robot:

>Even when two Cox process models are matched to have similar first- and 
second-order structure, they may still differ in the distribution of total 
intensity over local regions. These local aggregates retain some dependence 
information that is lost in pointwise marginals, while remaining much simpler 
than the full joint distribution.

:::

```{r}
# Local aggregate moment comparison for LGCP vs shifted CSCP
dist2d <- function(x, y) {
  sqrt((x[1] - y[1])^2 + (x[2] - y[2])^2)
}

sample_square <- function(n, ell) {
  cbind(runif(n, 0, ell), runif(n, 0, ell))
}

# Pair correlation functions
g_lgcp <- function(r, phi, s) {
  exp(log(1 + phi) * exp(-r / s))
}

g_cscp <- function(r, phi, s) {
  1 + phi * exp(-2 * r / s)
}

# Third-order product density ratios g3
g3_lgcp <- function(r12, r13, r23, phi, s) {
  c <- log(1 + phi)
  exp(c * exp(-r12 / s) +
      c * exp(-r13 / s) +
      c * exp(-r23 / s))
}

g3_cscp <- function(r12, r13, r23, phi, s) {
  rho12 <- exp(-r12 / s)
  rho13 <- exp(-r13 / s)
  rho23 <- exp(-r23 / s)

  1 +
    phi * (rho12^2 + rho13^2 + rho23^2) +
    2 * sqrt(2) * phi^(3/2) * rho12 * rho13 * rho23
}

# Monte Carlo moments of A(B)
mc_moments_A <- function(model = c("LGCP", "CSCP"),
                         ell,
                         lambda,
                         phi,
                         s,
                         n_mc = 20000) {
  model <- match.arg(model)

  area <- ell^2

  # First moment
  EA <- lambda * area

  # Second moment via E[g(U,V)]
  UV <- sample_square(2 * n_mc, ell)
  U <- UV[1:n_mc, , drop = FALSE]
  V <- UV[(n_mc + 1):(2 * n_mc), , drop = FALSE]

  rUV <- sqrt((U[,1] - V[,1])^2 + (U[,2] - V[,2])^2)

  g_vals <- if (model == "LGCP") {
    g_lgcp(rUV, phi = phi, s = s)
  } else {
    g_cscp(rUV, phi = phi, s = s)
  }

  EA2 <- lambda^2 * area^2 * mean(g_vals)
  VarA <- EA2 - EA^2

  # Third moment via E[g3(U,V,W)]
  XYZ <- sample_square(3 * n_mc, ell)
  X <- XYZ[1:n_mc, , drop = FALSE]
  Y <- XYZ[(n_mc + 1):(2 * n_mc), , drop = FALSE]
  Z <- XYZ[(2 * n_mc + 1):(3 * n_mc), , drop = FALSE]

  r12 <- sqrt((X[,1] - Y[,1])^2 + (X[,2] - Y[,2])^2)
  r13 <- sqrt((X[,1] - Z[,1])^2 + (X[,2] - Z[,2])^2)
  r23 <- sqrt((Y[,1] - Z[,1])^2 + (Y[,2] - Z[,2])^2)

  g3_vals <- if (model == "LGCP") {
    g3_lgcp(r12, r13, r23, phi = phi, s = s)
  } else {
    g3_cscp(r12, r13, r23, phi = phi, s = s)
  }

  EA3 <- lambda^3 * area^3 * mean(g3_vals)

  list(
    mean_A = EA,
    var_A  = VarA,
    EA2    = EA2,
    EA3    = EA3
  )
}

# Convert aggregate moments to count moments for N(B)
moments_N_from_A <- function(mean_A, var_A, EA2, EA3) {
  # For a Poisson mixture:
  # E[N]   = E[A]
  # E[N^2] = E[A^2] + E[A]
  # E[N^3] = E[A^3] + 3E[A^2] + E[A]

  EN  <- mean_A
  EN2 <- EA2 + mean_A
  EN3 <- EA3 + 3 * EA2 + mean_A

  VarN <- EN2 - EN^2

  mu3 <- EN3 - 3 * EN * EN2 + 2 * EN^3
  skewN <- mu3 / (VarN^(3/2))

  list(
    mean_N = EN,
    var_N = VarN,
    skew_N = skewN
  )
}
```

```{r}
library(dplyr)
library(purrr)
library(ggplot2)

lambda <- 100
phi    <- 1.5
s_cscp <- 0.05
s_lgcp <- 0.68 * s_cscp   # or whatever matched scale you want

ell_grid <- seq(0.02, 0.25, length.out = 12)

out <- map_dfr(ell_grid, function(ell) {
  A_lgcp <- mc_moments_A("LGCP", ell, lambda, phi, s_lgcp, n_mc = 10000)
  A_cscp <- mc_moments_A("CSCP", ell, lambda, phi, s_cscp, n_mc = 10000)

  N_lgcp <- moments_N_from_A(A_lgcp$mean_A, A_lgcp$var_A, A_lgcp$EA2, A_lgcp$EA3)
  N_cscp <- moments_N_from_A(A_cscp$mean_A, A_cscp$var_A, A_cscp$EA2, A_cscp$EA3)

  bind_rows(
    tibble(
      ell = ell,
      model = "LGCP",
      mean_N = N_lgcp$mean_N,
      var_N  = N_lgcp$var_N,
      skew_N = N_lgcp$skew_N
    ),
    tibble(
      ell = ell,
      model = "CSCP",
      mean_N = N_cscp$mean_N,
      var_N  = N_cscp$var_N,
      skew_N = N_cscp$skew_N
    )
  )
})

ggplot(out, aes(x = ell, y = skew_N, linetype = model, colour = model)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Square side length",
    y = "Skewness of N(B)",
    title = "Local count skewness by region size"
  ) +
  theme_bw()
```






