```{r include = FALSE}
library(tidyverse)
```

# 🚧 Inspecting the Margin(al)s

Having established that LGCP and CSCP models can exhibit nearly identical
second-order structure, we now ask: where do these models differ?

Our first foray into non-first- and second-order statistics begins with the
marginal distributions of the latent fields. While the pair correlation function 
captures second-order dependence, it does not fully characterise the distribution 
of $\Lambda(u)$ at a fixed location.

Importantly: although the latent intensity is not directly observed, it still
determines the local behaviour of the process, and therefore influences 
observable quantities such as counts in small regions. The hope is that the 
differences are large enough that they are able to survive what is becoming 
increasingly apparent as a total nuclear homogenizing poisson sampling step.

The most obvious way to start investigating the difference is to plot the two
processes marginal intensity distributions, for a fixed $(\lambda, \phi)$, and
obviously using the best-matching scale parameters (now easily calculatable using 
the amazing `match_lgcp_scale()` function).

:::callout-note
I don't know why I thought we needed to vary the scale parameters here. As we will
see below, the scale has no effect on the marginal distribution. Reason being: 
the scale describes the decay of the spatial dependence. The marginal distribution
is for a single location, no spatial dependence is involved. 
:::

## Recapping

So before we cover the differences in marginal distributions, it is probably a 
good idea for us to revise what the marginal distributions even are.

So recall, a cox process is defined by it's random intensity function $\Lambda(u)$.

Now, this random intensity function is a Random Field, meaning that each location
$u$ over which $\Lambda$ is defined, is distributed according to the full joint
distribution specified by $\Lambda$.

What does this actually mean though?

Well, $\Lambda(u)$ is essentially assigning a random distribution to each one of 
the uncountably infinite number of locations over which it is defined, and the exact
distributions being assigned are determined by the functional form and parameters
of $\Lambda$.

In the case of the LGCP (and CSCP), the random field $\Lambda$, is a special kind
of random field called a Gaussian Random Field. This means that if we consider
any finite set of locations, say $\{u_1, u_2, u_3, \dots, u_n\}$, then they will
be distributed according to a multivariate normal distribution (with WHAT parameters?).

So you could imagine drawing an infinite number of random distributions, positioned 
vertically at each location, and transforming them according to whatever functional
form $\Lambda$ takes. This is exactly what $\Lambda$ represents. An inifite number
of *possible* inhomogeneous intensity surfaces.

Now, what we have described is the full joint-distribution of $\Lambda$. But,
this is a *very* difficult object to estimate for observed point patterns.

If we want to find differences between two $\Lambda$'s, and we can directly 
observe, or even estimate it well, then perhaps we could look at some *properties*
of the $\Lambda$'s, and if the $\Lambda$'s we are comparing are sufficiently different,
then we might be able to tell them apart (poorly phrased).

One example of this we saw in the last few chapters, where the "property" we were
looking at was the PCF, a summary of essentially the *covariance* between two locations.

This section, we are considering a different "property", that is the *marginal*
distribution of the $\Lambda$'s.

So how exactly are the marginal distributions different than the full joint
distribution?

Essentially, we are "accounting for" all possible values of the other locations
and taking the correlation of them into account, and "averaging out" the effects
due to the correlation / covariance.

The marginal distribution is essentially telling us: in the complete absence of 
information about any other locations, what would the distribution of the intensity
look like at this single point.

In other words: If we repeatedly resampled from *just this single point*, after
resampling the entire surface $\Lambda$ many times, what would the distribution 
of the intensity surfaces look like?

## Recap - corrected

So before we cover the differences in marginal distributions, it is probably a 
good idea for us to revise what the marginal distributions even are.

Recall, a cox process is defined by its random intensity function $\Lambda(u)$.

This is a function which takes a location $u$ as an argument, and outputs a 
random variable, which describes the intensity of the field at that location.

The form $\Lambda$ takes, is a transformation of some Random Field. 

Note that both LGCP and CSCP transform *Gaussian* random fields. GRFs are random
fields such that any finite subset of them will be multivariate-normally 
distributed.

Now note, $\Lambda(u)$ determines the full joint distribution of the Cox Process, 
over all locations. This means (better would be: can be thought of as...) it is essentially responsible for assigning a 
random variable to each one of the uncountably infinite number of locations 
over which it is defined, and the exact distributions of these RVs being assigned are 
determined by the functional form and parameters of $\Lambda$.

This full joint distribution is a *very* difficult object to estimate from just 
observed point patterns.

So if we want to find differences between two $\Lambda$'s, and we can't directly 
observe, or even estimate them well, then perhaps we could look at some *properties*
of the $\Lambda$'s, and if the $\Lambda$'s we are comparing are sufficiently different,
then we might be able to tell them apart (poorly phrased).

One example of this we saw in the last few chapters, where the "property" we were
looking at was the PCF, a summary of essentially the *covariance* between two locations.

This section, we are considering a different "property", that is the *marginal*
distribution of the $\Lambda$'s.

So how exactly are the marginal distributions different than the full joint
distribution?

Essentially, we are "accounting for" all possible values of the other locations
and taking the correlation of them into account, and integrating out the 
dependence on all other locations.

The marginal distribution is essentially telling us: in the complete absence of 
information about any other locations, what would the distribution of the intensity
look like at this single point.

In other words, if we repeatedly simulated the full intensity surface $\Lambda$,
and recorded the value at a fixed location $u$, the resulting values would follow
the marginal distribution of $\Lambda(u)$.

## Trying to understand what is going on

OK so I'm happy with the *idea* of what the marginal distribution is.

I'm thinking about it almost as a summary / property of the full distribution. 
(Is this ok?)

So we want to use this unobservable summary to distinguish between the two 
processes, in the same way we were trying to use the unobserved *true* PCFs to
distinguish between them.

Note that the way we approached the PCFs, was by esitmating them using minimum 
contrast, and then basically "hoping" we did a good enough job (turns out we
didn't). So we would now like to do the same thing here. << Robot didn't like this.

:::callout-note
Some thoughts as I process:

The PCF is a summary / property of the joint-distribution, which just so 
happens to be observable in realized point patterns.

The marginal is NOT reflected in a direct way, in the observable point pattern. 
So essentially, we will have to use a summary of the marginal of some kind i.e. 
the quadrat stuff. But then we are essentially trying to ditnguish between two 
lambdas using a summary of a summary, which seems not good?
:::

Ok trying again:

So we saw previously that using the PCF failed to distinguish between the two
processes.

The PCF is just one possible summary of the true full joint distribution, $\Lambda$.
In particular, the PCF is a quantification of how often pairs of points occur
at a distance, $r$. This is useful, as this quantity is a good (clear and 
interpretable) way of describing what we might think of as "clustering". Unfortunately
however, as we saw, the clustering behaivour alone is not enough (or not able 
to be estimated accurately enough) to distnguish between the two processes.

So now, we have made the decision to try a different summary of the true full 
joint distribution, in the marginal distribution, $\Lambda(u)$. Now, notice there
is a *big* difference in using the marginal as a summary of the true $\Lambda$, 
versus the PCF: unlike the PCF, the marginal is NOT directly observable from the
observed point patterns. The PCF is very nice, because while it is a "direct" 
summary / property of the true underlying $\Lambda$, we can directly estimate it
using the observed point pattern. For the marginal however, *this cannot be done*.

So instead of estimating the marginal directly, we are effectively going to have
to use a summary derived from the observed point process, in order to estimate 
the marginal, which in turn is a summary of the true $\Lambda$. 

Now the hope is that the marginal distributions are different enough between the
two processes, that even looking at summaries, they will be distinguishable.

:::callout-note
OK robot didn't like "PCF is a summary of $\Lambda$."

Kind of confused by this, why is the marginal distribution considered a summary
of $\Lambda$, but the PCF is not?

What is the PCF a summary of? Is it a summary of the *covariance* between two 
locations in $\Lambda$?

What the fuck is a functional omg bruh 
:::

So with this in mind we now need to address two key questions:

1. What are the *true* marginal distributions of the two processes?

2. How can these marginal properties be accessed or approximated from observed point pattern data?
Note here we are not trying to estimate the actual marginal distributions, just
estimating properties of it.

3. Are the marginals actually distinguishable in practice?

:::callout-note
So I'm wondering about estimating the actual marginals.

Is this easier than estimating the full joint distribution? I would assume so.

But how much more difficult? Apparently kernel smoothing, then histogram of 
intensity values over discretized grid? Not very complicated, but subject to issues
with bandwidth and pixel size?

And then some deconvolution, poisson-mixture thing.
:::

## True marginals

Obvious first step. Derive these bad boys. Starting with the CSCP.

### Marginal distribution of the CSCP

Recall the definition of our beautiful CSCP:

$$
\Lambda(u) = \mu + Z(u)^2, \quad Z(u) \sim N(0, \sigma^2)
$$

Fix a location $u$. Since

$$
Z(u) \sim N(0, \sigma^2)
$$

we can standardise, to obtain:

$$
\frac{Z(u)}{\sigma} \sim N(0, 1)
$$

Now squaring both sides, we get:

$$
(\frac{Z(u)}{\sigma})^2 \sim \chi_1^2
$$

since the square of a standard normal RV is chi-square with 1 df.

So we have:

$$
\frac{Z(u)^2}{\sigma^2} \sim \chi_1^2
$$

Now, recall that

$$
\Lambda(u) = \mu + Z(u)^2
$$

so

$$
\Lambda(u) - \mu = Z(u)^2 \Longrightarrow \frac{\Lambda(u) - \mu}{\sigma^2} = \frac{Z(u)^2}{\sigma^2}
$$

And using the above we have:

$$
\frac{\Lambda(u) - \mu}{\sigma^2} \sim \chi_1^2
$$

OK cool, so now what?

Let 

$$
Y = \frac{\Lambda(u) - \mu}{\sigma^2} 
$$

Then, from above we have $Y \sim \chi_1^2$ and also

$$
\Lambda(u) = \mu + \sigma^2 \Big(\frac{\Lambda(u) - \mu}{\sigma^2}\Big) = \mu + \sigma^2Y
$$

And so we can use the change-of-variable formula:

$$
f_{\Lambda(u)}(\lambda) = f_Y(\frac{\lambda - \mu}{\sigma^2})|\frac{d}{d\lambda}\Big( \frac{\lambda - \mu}{\sigma^2}\Big)|
$$

and since

$$
\frac{d}{d\lambda}\Big( \frac{\lambda - \mu}{\sigma^2}\Big) = \frac{1}{\sigma^2}
$$

we obtain

$$
f_{\Lambda(u)}(\lambda) = \frac{1}{\sigma^2} f_{\chi_1^2}\Big( \frac{\lambda - \mu}{\sigma^2}\Big), \quad \lambda > \mu
$$

Also since $Z(u)^2 \ge 0$, we must have $\Lambda(u) \ge \mu$, so the density is
zero for $\lambda \le \mu$.

Now, if we want to be extra explicit, we have:

$$
f_{\chi_1^2} = \frac{1}{\sqrt{2\pi}}x^{-1/2}e^{-x/2},\quad x > 0
$$

and so subbing this in we get:

$$
f_{\Lambda(u)}(\lambda) = \frac{1}{\sigma^2} \frac{1}{\sqrt{2\pi}}\Big( \frac{\lambda - \mu}{\sigma^2}\Big)^{-1/2}\exp\Big(- \frac{\lambda - \mu}{2\sigma^2}\Big), \quad \lambda > \mu
$$

Or equivalently:

$$
f_{\Lambda(u)}(\lambda) =  \frac{1}{\sqrt{2\pi \sigma^2 (\lambda - \mu)}}\exp\Big(- \frac{\lambda - \mu}{2\sigma^2}\Big), \quad \lambda > \mu
$$

So, at a fixed location $u$, the marginal distribution follows a shifted and 
scaled Chi-square.

In particular:

- it has support over $(\mu, \infty)$

- explodes to infinity (?) as $\lambda \to \mu$

- Right tail decays exponentially


### Marginal distribution of the LGCP

Now recall the definition of the LGCP:

$$
\Lambda(u) = \exp(Z(u)), \quad Z(u) \sim N(\mu, \sigma^2)
$$

Fix a location $u$. Since

$$
Z(u) \sim N(\mu, \sigma^2)
$$

and

$$
\Lambda(u) = \exp(Z(u))
$$

we can take logs of both sides to obtain

$$
\log \Lambda(u) = Z(u)
$$

But then, since $Z(u) \sim N(\mu, \sigma^2)$, we immediately have

$$
\log \Lambda(u) \sim N(\mu, \sigma^2)
$$

And this is exactly the definition of a lognormal random variable. Hence

$$
\Lambda(u) \sim \operatorname{Lognormal}(\mu, \sigma^2)
$$

OK cool, so now what?

If we want to write down the density explicitly, let

$$
Y = Z(u)
$$

Then

$$
Y \sim N(\mu, \sigma^2)
$$

and also

$$
\Lambda(u) = e^Y
$$

Equivalently,

$$
Y = \log \Lambda(u)
$$

So we can use the change-of-variable formula:

$$
f_{\Lambda(u)}(\lambda) = f_Y(\log \lambda)\Big|\frac{d}{d\lambda}(\log \lambda)\Big|,
\qquad \lambda > 0
$$

Now since

$$
\frac{d}{d\lambda}(\log \lambda) = \frac{1}{\lambda}
$$

we obtain

$$
f_{\Lambda(u)}(\lambda) = \frac{1}{\lambda} f_{N(\mu,\sigma^2)}(\log \lambda),
\qquad \lambda > 0
$$

Now, if we want to be extra explicit, recall that the density of a normal random
variable is

$$
f_{N(\mu,\sigma^2)}(y)
=
\frac{1}{\sigma \sqrt{2\pi}}
\exp\Big(
-\frac{(y-\mu)^2}{2\sigma^2}
\Big), \quad y \in \mathbb{r echo = FALSE}
$$

So substituting $y = \log \lambda$, we get

$$
f_{\Lambda(u)}(\lambda)
=
\frac{1}{\lambda}
\frac{1}{\sigma \sqrt{2\pi}}
\exp\Big(
-\frac{(\log \lambda - \mu)^2}{2\sigma^2}
\Big), \quad \lambda > 0
$$

That is,

$$
f_{\Lambda(u)}(\lambda)
=
\frac{1}{\lambda \sigma \sqrt{2\pi}}
\exp\Big(
-\frac{(\log \lambda - \mu)^2}{2\sigma^2}
\Big), \quad \lambda > 0
$$

Also, since $\Lambda(u)=\exp(Z(u)) > 0$ always, the density is zero for
$\lambda \le 0$.

So, at a fixed location $u$, the marginal intensity follows a lognormal random
variable.

In particular:

- it has support over $(0, \infty)$

- tends to $0$ as $\lambda \to 0$

- has a heavier right tail than the CSCP marginal

## Plotting the PDF

Setup

```{r echo = FALSE}
#| code-fold: true
devtools::load_all()

# Params
lambda <- 100
phi <- 2
s <- 0.05 # not used in the marginals, but included for consistency   
df <- 1

# Get parameterisations using helpers
lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

# Extract
mu_log <- lgcp_pars$mu
sigma_log <- lgcp_pars$sigma

mu_cscp <- cscp_pars$mu
sigma2_cscp <- cscp_pars$var

# Marginal density functions

# LGCP marginal density: Lognormal(mu_log, sigma_log^2)
d_lgcp_marginal <- function(x, mu_log, sigma_log) {
  dlnorm(x, meanlog = mu_log, sdlog = sigma_log)
}

# Shifted CSCP marginal density:
# Lambda(u) = mu + sigma^2 * Y, Y ~ chisq(df = 1)
d_cscp_marginal <- function(x, mu, sigma2) {
  out <- numeric(length(x))
  ok <- x > mu
  out[ok] <- dchisq((x[ok] - mu) / sigma2, df = 1) / sigma2
  out
}
```

Standard PDF on ordinary scale

```{r echo = FALSE}
x1 <- seq(0.001, 300, length.out = 5000)

y_lgcp_1 <- d_lgcp_marginal(x1, mu_log = mu_log, sigma_log = sigma_log)
y_cscp_1 <- d_cscp_marginal(x1, mu = mu_cscp, sigma2 = sigma2_cscp)

ylim1 <- range(c(y_lgcp_1, y_cscp_1), finite = TRUE)

plot(
  x1, y_lgcp_1,
  type = "l",
  lwd = 2,
  col = "blue",
  xlab = expression(lambda),
  ylab = "Density",
  ylim = ylim1,
  main = "Marginal PDFs"
)
lines(x1, y_cscp_1, col = "red", lwd = 2)

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col = c("blue", "red"),
  lwd = 2,
  bty = "n"
)
```

```{r echo = FALSE}
ylim2 <- c(0, 0.1)
plot(
  x1, y_lgcp_1,
  type = "l",
  lwd = 2,
  col = "blue",
  xlab = expression(lambda),
  ylab = "Density",
  ylim = ylim2,
  main = "Marginal PDFs"
)
lines(x1, y_cscp_1, col = "red", lwd = 2)

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col = c("blue", "red"),
  lwd = 2,
  bty = "n"
)
```

Varying phi:

```{r echo = FALSE}
phi_vals <- c(0.5, 1, 2)
lambda <- 100
s <- 0.05
df <- 1

x1 <- seq(0.001, 300, length.out = 5000)
ylim2 <- c(0, 0.06)

oldpar <- par(mfrow = c(1, 3), mar = c(4,4,3,1))
on.exit(par(oldpar), add = TRUE)

for (phi in phi_vals) {
  # Get parameters using your package
  lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
  cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

  mu_log <- lgcp_pars$mu
  sigma_log  <- lgcp_pars$sigma

  mu_cscp <- cscp_pars$mu
  sigma2_cscp <- cscp_pars$var

  # Densities
  y_lgcp <- d_lgcp_marginal(x1, mu_log, sigma_log)
  y_cscp <- d_cscp_marginal(x1, mu_cscp, sigma2_cscp)

  plot(
    x1, y_lgcp,
    type = "l",
    lwd = 2,
    col = "blue",
    xlab = expression(lambda),
    ylab = "Density",
    ylim = ylim2,
    main = bquote(phi == .(phi))
  )
  
  lines(x1, y_cscp, col = "red", lwd = 2)
}

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col = c("blue", "red"),
  lwd = 2,
  bty = "n"
)
```

Vary lambda (phi = 2):

```{r echo = FALSE}
lambda_vals <- c(50, 100, 200)
phi <- 2
s <- 0.05
df <- 1

x1 <- seq(0.001, 400, length.out = 5000)
ylim2 <- c(0, 0.1)

d_lgcp_marginal <- function(x, mu_log, sigma_log) {
  dlnorm(x, meanlog = mu_log, sdlog = sigma_log)
}

d_cscp_marginal <- function(x, mu, sigma2) {
  out <- numeric(length(x))
  ok <- x > mu
  out[ok] <- dchisq((x[ok] - mu) / sigma2, df = 1) / sigma2
  out
}

oldpar <- par(mfrow = c(1, 3), mar = c(4,4,3,1))
on.exit(par(oldpar), add = TRUE)

for (lambda in lambda_vals) {
  # Get parameters using your package
  lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
  cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

  mu_log <- lgcp_pars$mu
  sigma_log <- lgcp_pars$sigma

  mu_cscp <- cscp_pars$mu
  sigma2_cscp <- cscp_pars$var

  # Densities
  y_lgcp <- d_lgcp_marginal(x1, mu_log, sigma_log)
  y_cscp <- d_cscp_marginal(x1, mu_cscp, sigma2_cscp)

  # Plot
  plot(
    x1, y_lgcp,
    type = "l",
    lwd = 2,
    col = "blue",
    xlab = expression(lambda),
    ylab = "Density",
    ylim = ylim2,
    main = bquote(lambda == .(lambda))
  )
  
  lines(x1, y_cscp, col = "red", lwd = 2)
}

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col = c("blue", "red"),
  lwd = 2,
  bty = "n"
)
```

Vary lambda (phi = 0.1)

```{r echo = FALSE}
lambda_vals <- c(50, 100, 200)
phi <- 0.1
s <- 0.05
df <- 1

x1 <- seq(0.001, 400, length.out = 5000)
ylim2 <- c(0, 0.1)

d_lgcp_marginal <- function(x, mu_log, sigma_log) {
  dlnorm(x, meanlog = mu_log, sdlog = sigma_log)
}

d_cscp_marginal <- function(x, mu, sigma2) {
  out <- numeric(length(x))
  ok <- x > mu
  out[ok] <- dchisq((x[ok] - mu) / sigma2, df = 1) / sigma2
  out
}

oldpar <- par(mfrow = c(1, 3), mar = c(4,4,3,1))
on.exit(par(oldpar), add = TRUE)

for (lambda in lambda_vals) {
  # Get parameters using your package
  lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
  cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

  mu_log <- lgcp_pars$mu
  sigma_log <- lgcp_pars$sigma

  mu_cscp <- cscp_pars$mu
  sigma2_cscp <- cscp_pars$var

  # Densities
  y_lgcp <- d_lgcp_marginal(x1, mu_log, sigma_log)
  y_cscp <- d_cscp_marginal(x1, mu_cscp, sigma2_cscp)

  # Plot
  plot(
    x1, y_lgcp,
    type = "l",
    lwd = 2,
    col = "blue",
    xlab = expression(lambda),
    ylab = "Density",
    ylim = ylim2,
    main = bquote(lambda == .(lambda))
  )
  
  lines(x1, y_cscp, col = "red", lwd = 2)
}

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col = c("blue", "red"),
  lwd = 2,
  bty = "n"
)
```

Zoom near the left boundary

```{r echo = FALSE}
# Params
lambda <- 100
phi <- 2
s <- 0.05 # not used in the marginals, but included for consistency   
df <- 1

# Get parameterisations using helpers
lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

# Extract
mu_log <- lgcp_pars$mu
sigma_log <- lgcp_pars$sigma

mu_cscp <- cscp_pars$mu
sigma2_cscp <- cscp_pars$var

x2 <- seq(0.001, 30, length.out = 5000)

y_lgcp_2 <- d_lgcp_marginal(x2, mu_log = mu_log, sigma_log = sigma_log)
y_cscp_2 <- d_cscp_marginal(x2, mu = mu_cscp, sigma2 = sigma2_cscp)

# Clipping the y-axis a bit so the CSCP spike doesn't completely flatten the LGCP
ylim2 <- c(0, max(y_lgcp_2, na.rm = TRUE) * 3)

plot(
  x2, y_lgcp_2,
  type = "l",
  lwd  = 2,
  col  = "blue",
  xlab = expression(lambda),
  ylab = "Density",
  ylim = ylim2,
  main = "Marginal PDFs: left-endpoint zoom"
)
lines(x2, y_cscp_2, col = "red", lwd = 2)

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

PDF with log-scaled y-axis

```{r echo = FALSE}
x3 <- seq(0.001, 1000, length.out = 10000)

y_lgcp_3 <- d_lgcp_marginal(x3, mu_log = mu_log, sigma_log = sigma_log)
y_cscp_3 <- d_cscp_marginal(x3, mu = mu_cscp, sigma2 = sigma2_cscp)

# Need strictly positive values for log = "y"
eps <- 1e-300
y_lgcp_3[y_lgcp_3 <= 0] <- eps
y_cscp_3[y_cscp_3 <= 0] <- eps

plot(
  x3, y_lgcp_3,
  type = "l",
  lwd  = 2,
  col  = "blue",
  log  = "y",
  xlab = expression(lambda),
  ylab = "Density (log scale)",
  main = "Marginal PDFs: log-scale y-axis"
)
lines(x3, y_cscp_3, col = "red", lwd = 2)

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

### What are we seeing?

???

## Maybe CDF is a little nicer?

Corresponding cumulative distribution functions follow immediately from the 
transformed representations used above.

Hopefully useful as: firstly, they provide a cumulative description
of the one-point marginal laws. secondly, they will later allow us to investigate
the upper-tail behaviour of the two models through their survival functions.

### CDF of the CSCP marginal

Recall that under the shifted CSCP,
$$
\Lambda(u) = \mu + Z(u)^2, \quad Z(u) \sim N(0, \sigma^2)
$$
and from the previous subsection we showed that

$$
\frac{\Lambda(u) - \mu}{\sigma^2} \sim \chi_1^2
$$

So for any $\lambda > \mu$, we have

$$
F_{\Lambda(u)}(\lambda)
=
P(\Lambda(u) \le \lambda)
$$

Using the transformation above, this becomes

$$
P(\Lambda(u) \le \lambda)
=
P\Big( \frac{\Lambda(u) - \mu}{\sigma^2} \le \frac{\lambda - \mu}{\sigma^2} \Big)
$$

Since
$$
\frac{\Lambda(u) - \mu}{\sigma^2} \sim \chi_1^2
$$
it follows that

$$
F_{\Lambda(u)}(\lambda)
=
F_{\chi_1^2}\Big( \frac{\lambda - \mu}{\sigma^2} \Big), \quad \lambda > \mu
$$

Also, since $\Lambda(u) \ge \mu$ always, we have

$$
F_{\Lambda(u)}(\lambda) = 0, \quad \lambda \le \mu
$$

So the full CDF is

$$
F_{\Lambda(u)}(\lambda)
=
\begin{cases}
0, & \lambda \le \mu, \\
F_{\chi_1^2}\Big( \frac{\lambda - \mu}{\sigma^2} \Big), & \lambda > \mu
\end{cases}
$$

So, at a fixed location $u$, the marginal CDF of the CSCP is simply a shifted
and scaled chi-square CDF.

### CDF of the LGCP marginal

Now recall that under the LGCP,
$$
\Lambda(u) = \exp(Z(u)), \quad Z(u) \sim N(\mu, \sigma^2)
$$

From the previous subsection, we showed that

$$
\log \Lambda(u) \sim N(\mu, \sigma^2)
$$

So for any $\lambda > 0$, we have

$$
F_{\Lambda(u)}(\lambda)
=
P(\Lambda(u) \le \lambda)
$$

Taking logs of both sides inside the probability gives

$$
P(\Lambda(u) \le \lambda)
=
P(\log \Lambda(u) \le \log \lambda)
$$

But since
$$
\log \Lambda(u) \sim N(\mu, \sigma^2)
$$
this becomes

$$
F_{\Lambda(u)}(\lambda)
=
P(Z(u) \le \log \lambda)
$$

Now standardising,

$$
P(Z(u) \le \log \lambda)
=
P\Big( \frac{Z(u)-\mu}{\sigma} \le \frac{\log \lambda - \mu}{\sigma} \Big)
$$

Since
$$
\frac{Z(u)-\mu}{\sigma} \sim N(0,1)
$$

it follows that

$$
F_{\Lambda(u)}(\lambda)
=
\Phi\Big( \frac{\log \lambda - \mu}{\sigma} \Big), \quad \lambda > 0
$$

where $\Phi$ is the standard normal CDF.

Also, since $\Lambda(u) = \exp(Z(u)) > 0$ always, we have

$$
F_{\Lambda(u)}(\lambda) = 0, \quad \lambda \le 0
$$

So the full CDF is

$$
F_{\Lambda(u)}(\lambda)
=
\begin{cases}
0, & \lambda \le 0 \\
\Phi\Big( \frac{\log \lambda - \mu}{\sigma} \Big), & \lambda > 0
\end{cases}
$$

So, at a fixed location $u$, the marginal CDF of the LGCP is the CDF of a
lognormal random variable.

### Comparison

While the CSCP and LGCP can exhibit very similar second-order structure,
their one-point marginal cumulative distributions are fundamentally different 
(although *how* different, I feel is not very much).

In particular:

- the CSCP marginal CDF is that of a shifted and scaled chi-square random variable

- the LGCP marginal CDF is that of a lognormal random variable

### Marginal survival functions

Since we are especially interested in how the two marginals differ in their
upper tails, it is natural to also consider the survival functions,

$$
\bar F_{\Lambda(u)}(\lambda) = 1 - F_{\Lambda(u)}(\lambda)
$$

For the CSCP, this gives

$$
\bar F_{\Lambda(u)}(\lambda)
=
\begin{cases}
1, & \lambda \le \mu \\
1 - F_{\chi_1^2}\Big( \frac{\lambda - \mu}{\sigma^2} \Big), & \lambda > \mu
\end{cases}
$$

and for the LGCP,

$$
\bar F_{\Lambda(u)}(\lambda)
=
\begin{cases}
1, & \lambda \le 0 \\
1 - \Phi\Big( \frac{\log \lambda - \mu}{\sigma} \Big), & \lambda > 0
\end{cases}
$$

## Plotting the CDFs

```{r echo = FALSE}
# Settings
lambda <- 100
phi    <- 2
s      <- 0.05
df     <- 1

lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

mu_log     <- lgcp_pars$mu
sigma_log  <- lgcp_pars$sigma

mu_cscp     <- cscp_pars$mu
sigma2_cscp <- cscp_pars$var

# Marginal CDF functions
p_lgcp_marginal <- function(x, mu_log, sigma_log) {
  plnorm(x, meanlog = mu_log, sdlog = sigma_log)
}

p_cscp_marginal <- function(x, mu, sigma2) {
  out <- numeric(length(x))
  ok <- x > mu
  out[ok] <- pchisq((x[ok] - mu) / sigma2, df = 1)
  out
}
```

Standard CDF
```{r echo = FALSE}
x1 <- seq(0.001, 300, length.out = 5000)

F_lgcp_1 <- p_lgcp_marginal(x1, mu_log = mu_log, sigma_log = sigma_log)
F_cscp_1 <- p_cscp_marginal(x1, mu = mu_cscp, sigma2 = sigma2_cscp)

plot(
  x1, F_lgcp_1,
  type = "l",
  lwd  = 2,
  col  = "blue",
  xlab = expression(lambda),
  ylab = "CDF",
  ylim = c(0, 1),
  main = "Marginal CDFs"
)
lines(x1, F_cscp_1, col = "red", lwd = 2)

legend(
  "bottomright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

Varying phi

```{r echo = FALSE}
phi_vals <- c(0.5, 1, 2)
lambda   <- 100
s        <- 0.05
df       <- 1

x1 <- seq(0.001, 300, length.out = 5000)

oldpar <- par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
on.exit(par(oldpar), add = TRUE)

for (phi in phi_vals) {

  lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
  cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

  mu_log      <- lgcp_pars$mu
  sigma_log   <- lgcp_pars$sigma
  mu_cscp     <- cscp_pars$mu
  sigma2_cscp <- cscp_pars$var

  F_lgcp <- p_lgcp_marginal(x1, mu_log, sigma_log)
  F_cscp <- p_cscp_marginal(x1, mu_cscp, sigma2_cscp)

  plot(
    x1, F_lgcp,
    type = "l",
    lwd  = 2,
    col  = "blue",
    xlab = expression(lambda),
    ylab = "CDF",
    ylim = c(0, 1),
    main = bquote(phi == .(phi))
  )

  lines(x1, F_cscp, col = "red", lwd = 2)
}

legend(
  "bottomright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

Left-endpoint zoom

```{r echo = FALSE}
x2 <- seq(0.001, 40, length.out = 5000)

F_lgcp_2 <- p_lgcp_marginal(x2, mu_log = mu_log, sigma_log = sigma_log)
F_cscp_2 <- p_cscp_marginal(x2, mu = mu_cscp, sigma2 = sigma2_cscp)

plot(
  x2, F_lgcp_2,
  type = "l",
  lwd  = 2,
  col  = "blue",
  xlab = expression(lambda),
  ylab = "CDF",
  ylim = c(0, 0.4),
  main = "Marginal CDFs: left-endpoint zoom"
)
lines(x2, F_cscp_2, col = "red", lwd = 2)

legend(
  "bottomright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

And survival plots I guess?

```{r echo = FALSE}
s_lgcp_marginal <- function(x, mu_log, sigma_log) {
  1 - plnorm(x, meanlog = mu_log, sdlog = sigma_log)
}

s_cscp_marginal <- function(x, mu, sigma2) {
  out <- rep(1, length(x))
  ok <- x > mu
  out[ok] <- 1 - pchisq((x[ok] - mu) / sigma2, df = 1)
  out
}

x3 <- seq(0.001, 300, length.out = 5000)

S_lgcp_3 <- s_lgcp_marginal(x3, mu_log = mu_log, sigma_log = sigma_log)
S_cscp_3 <- s_cscp_marginal(x3, mu = mu_cscp, sigma2 = sigma2_cscp)

plot(
  x3, S_lgcp_3,
  type = "l",
  lwd  = 2,
  col  = "blue",
  xlab = expression(lambda),
  ylab = "Survival probability",
  ylim = c(0, 1),
  main = "Marginal survival functions"
)
lines(x3, S_cscp_3, col = "red", lwd = 2)

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
x4 <- seq(0.001, 3000, length.out = 10000)

S_lgcp_4 <- s_lgcp_marginal(x4, mu_log = mu_log, sigma_log = sigma_log)
S_cscp_4 <- s_cscp_marginal(x4, mu = mu_cscp, sigma2 = sigma2_cscp)

eps <- 1e-300
S_lgcp_4[S_lgcp_4 <= 0] <- eps
S_cscp_4[S_cscp_4 <= 0] <- eps

plot(
  x4, S_lgcp_4,
  type = "l",
  lwd  = 2,
  col  = "blue",
  log  = "y",
  xlab = expression(lambda),
  ylab = "Survival probability (log scale)",
  main = "Marginal survival functions: log-scale y-axis"
)
lines(x4, S_cscp_4, col = "red", lwd = 2)

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

Varying phi:

```{r echo = FALSE}
phi_vals <- c(0.5, 1, 2)
lambda   <- 100
s        <- 0.05
df       <- 1

x1 <- seq(0.001, 1500, length.out = 10000)



oldpar <- par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
on.exit(par(oldpar), add = TRUE)

for (phi in phi_vals) {

  lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
  cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

  mu_log      <- lgcp_pars$mu
  sigma_log   <- lgcp_pars$sigma
  mu_cscp     <- cscp_pars$mu
  sigma2_cscp <- cscp_pars$var

  S_lgcp <- s_lgcp_marginal(x1, mu_log, sigma_log)
  S_cscp <- s_cscp_marginal(x1, mu_cscp, sigma2_cscp)

  eps <- 1e-300
  S_lgcp[S_lgcp <= 0] <- eps
  S_cscp[S_cscp <= 0] <- eps

  plot(
    x1, S_lgcp,
    type = "l",
    lwd  = 2,
    col  = "blue",
    log  = "y",
    xlab = expression(lambda),
    ylab = "Survival (log scale)",
    main = bquote(phi == .(phi))
  )

  lines(x1, S_cscp, col = "red", lwd = 2)
}

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

## Relative scale plot

Relative scale survival plot

```{R echo = FALSE}
lambda <- 100
phi    <- 2
s      <- 0.05
df     <- 1

# Use your package helpers
lgcp_pars <- lgcp_params(lambda = lambda, phi = phi, s = s)
cscp_pars <- cscp_params(lambda = lambda, phi = phi, s = s, df = df)

mu_log      <- lgcp_pars$mu
sigma_log   <- lgcp_pars$sigma
mu_cscp     <- cscp_pars$mu
sigma2_cscp <- cscp_pars$var

rel_grid <- seq(0.001, 20, length.out = 10000)   # Lambda / E[Lambda]
x_abs    <- lambda * rel_grid # Convert back to absolute intensity

S_lgcp <- s_lgcp_marginal(x_abs, mu_log = mu_log, sigma_log = sigma_log)
S_cscp <- s_cscp_marginal(x_abs, mu = mu_cscp, sigma2 = sigma2_cscp)

eps <- 1e-300
S_lgcp[S_lgcp <= 0] <- eps
S_cscp[S_cscp <= 0] <- eps

plot(
  rel_grid, S_lgcp,
  type = "l",
  lwd  = 2,
  col  = "blue",
  log  = "y",
  xlab = expression(Lambda(u) / E[Lambda(u)]),
  ylab = "Survival probability (log scale)",
  main = "Marginal survival functions on relative scale"
)
lines(rel_grid, S_cscp, col = "red", lwd = 2)

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

Relative-scale PDF

```{r echo = FALSE}
rel_grid <- seq(0.001, 8, length.out = 10000)
x_abs    <- lambda * rel_grid

f_rel_lgcp <- lambda * d_lgcp_marginal(x_abs, mu_log = mu_log, sigma_log = sigma_log)
f_rel_cscp <- lambda * d_cscp_marginal(x_abs, mu = mu_cscp, sigma2 = sigma2_cscp)

plot(
  rel_grid, f_rel_lgcp,
  type = "l",
  lwd  = 2,
  col  = "blue",
  xlab = expression(Lambda(u) / E[Lambda(u)]),
  ylab = "Density",
  main = "Marginal PDFs on relative scale"
)
lines(rel_grid, f_rel_cscp, col = "red", lwd = 2)

legend(
  "topright",
  legend = c("LGCP", "CSCP"),
  col    = c("blue", "red"),
  lwd    = 2,
  bty    = "n"
)
```

## Quantiles table

```{R}
probs <- c(0.90, 0.95, 0.99, 0.999, 0.9999)

q_lgcp <- qlnorm(probs, meanlog = mu_log, sdlog = sigma_log)
q_cscp <- mu_cscp + sigma2_cscp * qchisq(probs, df = 1)

quantile_tab <- data.frame(
  prob = probs,
  LGCP = q_lgcp,
  CSCP = q_cscp,
  LGCP_over_mean = q_lgcp / lambda,
  CSCP_over_mean = q_cscp / lambda
)

print(quantile_tab, row.names = FALSE)
```

## Tail ratio plot

```{R}
rel_grid <- seq(0.5, 20, length.out = 10000)
x_abs <- lambda * rel_grid

S_lgcp <- s_lgcp_marginal(x_abs, mu_log = mu_log, sigma_log = sigma_log)
S_cscp <- s_cscp_marginal(x_abs, mu = mu_cscp, sigma2 = sigma2_cscp)

tail_ratio <- S_lgcp / S_cscp
tail_ratio[!is.finite(tail_ratio)] <- NA

plot(
  rel_grid, tail_ratio,
  type = "l",
  lwd = 2,
  col = "black",
  log = "y",
  xlab = expression(Lambda(u) / E[Lambda(u)]),
  ylab = "LGCP survival / CSCP survival",
  main = "Relative tail probability ratio"
)
abline(h = 1, lty = 2)
```

ratio > 1: LGCP has heavier upper tail there
ratio < 1: CSCP heavier there

## Final Verdict

So what we thinking??

The two models are nearly indistinguishable across the **majority** of the distribution, 
and only diverge in the extreme tail corresponding to intensities exceeding 
approximately 10–15 times the mean. Not promising at ALL.
