```{r include=FALSE}
devtools::load_all()
library(dplyr)
library(purrr)
library(tidyr)
library(spatstat)
```
# Quadrats 2

```{r include=FALSE}
# All the code from last section, I know I should make this include, sue me

# Calculate quadrat counts using spatstat's quadratcount()
# Example of splitting observation window into 2x2 grid:
#
# > get_quadrat_counts(X_lgcp, 2)
# [1] 32 21 15 28
get_quadrat_counts <- function(X, nx, ny = nx) {
  qc <- quadratcount(X, nx = nx, ny = ny)
  as.numeric(qc)
}

# Calculate sample skewness summary
# Making sure to return NA in the case that all counts 
#       are the same (sample variance = 0)
safe_skewness <- function(x) {
  x <- as.numeric(x)
  if (length(x) < 2) stop("Require at least 2 quadrats")
  
  s2 <- stats::var(x)
  if (!is.finite(s2) || s2 <= 0) return(NA_real_)
  
  # Third moment calculation
  # m3 = mean((N_j - mean(N))^3)
  m3 <- mean((x - mean(x))^3)
  
  # Return standardised skewness
  # gamma1 = m3 / s^3 (s sample SD)
  m3 / (sqrt(s2)^3)
}


# Saves the counts vector from above as a dataframe, with extra sim-info 
# Example with counts = [1] 32 21 15 28:
#
# > count_hist_df(test, "lgcp", 1, 2)
# # A tibble: 4 × 6
#   model   rep    nx    ny     k  freq
#   <chr> <dbl> <dbl> <dbl> <int> <dbl>
# 1 lgcp      1     2     2    15     1
# 2 lgcp      1     2     2    21     1
# 3 lgcp      1     2     2    28     1
# 4 lgcp      1     2     2    32     1
count_hist_df <- function(counts, model, rep, nx, ny = nx) {
  # Tabulating to store more compactly
  # Helps in the case we are using very fine grid resolution
  tab <- table(counts)
  
  tibble(
    model = model, # "LGCP" or "CSCP"
    rep = rep, # replicate number
    nx = nx, # grid resolution
    ny = ny, # grid resolution
    k = as.integer(names(tab)), # The unique count values
    freq = as.numeric(tab) # How many quadrats had each of the above count values
  )
}

# Same as above, but instead of saving the raw counts, it saves the stats info
count_stat_df <- function(counts, model, rep, nx, ny = nx) {
  counts <- as.numeric(counts)
  J <- length(counts)

  # Empirical count distribution
  tab <- table(counts)
  props <- as.numeric(tab) / J
  k_vals <- as.integer(names(tab))

  # Peak height and peak width
  mode_index <- which.max(props)
  T_peak <- props[mode_index]

  # Interquartile width
  q25 <- as.numeric(quantile(counts, 0.25, type = 1))
  q75 <- as.numeric(quantile(counts, 0.75, type = 1))
  T_iqr_width <- q75 - q25

  # Lower / central / upper quantiles
  T_q05 <- as.numeric(quantile(counts, 0.05, type = 1))
  T_median <- as.numeric(quantile(counts, 0.50, type = 1))
  T_q995 <- as.numeric(quantile(counts, 0.995, type = 1))

  # Baseline dispersion
  mean_count <- mean(counts)
  var_count <- stats::var(counts)
  T_disp <- if (mean_count > 0) var_count / mean_count else NA_real_

  tibble(
    model = model,
    rep = rep,
    nx = nx,
    ny = ny,
    T_q05 = T_q05,
    T_peak = T_peak,
    T_iqr_width = T_iqr_width,
    T_median = T_median,
    T_q995 = T_q995,
    T_disp = T_disp
  )
}

run_quadrat_count_study <- function(
    W, # Observation window
    lambda, 
    phi,
    s_cscp = 0.05, # LGCP scale will be best-matching
    grids = c(4, 8, 16, 32),
    n_sims = 200,
    seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)
  
  # Match LGCP scale to CSCP scale
  s_lgcp <- match_lgcp_scale(phi = phi, s_cscp = s_cscp)
  
  # Objects for storing histograms and moments
  hist_list <- vector("list", length = 2 * n_sims * length(grids))
  moment_list <- vector("list", length = 2 * n_sims * length(grids))
  
  # Index is just for keeping calculated counts and summaries together
  idx <- 1
  
  # Study loop
  for (rep in seq_len(n_sims)) {
    # Simulate one LGCP pattern
    X_lgcp <- sim_lgcp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_lgcp
    )$pp
    
    # Simulate one CSCP pattern
    X_cscp <- sim_cscp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_cscp,
      df = 1
    )$pp
    
    # For each grid size
    for (m in grids) {
      # Calculate the counts per grid-tile
      counts_lgcp <- get_quadrat_counts(X_lgcp, nx = m, ny = m)
      counts_cscp <- get_quadrat_counts(X_cscp, nx = m, ny = m)
      
      # Keep track of the counts data plus summaries for each process
      hist_list[[idx]] <- count_hist_df(counts_lgcp, model = "LGCP", rep = rep, nx = m)
      moment_list[[idx]] <- count_stat_df(counts_lgcp, model = "LGCP", rep = rep, nx = m)
      idx <- idx + 1
      
      hist_list[[idx]] <- count_hist_df(counts_cscp, model = "CSCP", rep = rep, nx = m)
      moment_list[[idx]] <- count_stat_df(counts_cscp, model = "CSCP", rep = rep, nx = m)
      idx <- idx + 1
    }
  }
  
  # Combine all the data
  hist_df <- bind_rows(hist_list)
  moment_df <- bind_rows(moment_list)
  
  # Complete missing count values so summaries are comparable
  # For example if LGCP has counts up to 100 but CSCP only to 50
  # Then the below code sets all of CSCP counts from 51 to 100 = 0
  hist_df <- hist_df %>%
    group_by(nx, ny) %>%
    tidyr::complete(model, rep, k = 0:max(k), fill = list(freq = 0)) %>%
    ungroup()
  
  # Calculate proportion of each count within each replicate/grid
  hist_df <- hist_df %>%
    group_by(model, rep, nx, ny) %>%
    mutate(prop = freq / sum(freq)) %>%
    ungroup()
  
  # Histogram envelopes across simulations
  hist_summary <- hist_df %>%
    group_by(model, nx, ny, k) %>%
    summarise(
      mean_prop = mean(prop),
      med_prop = median(prop),
      lo_prop = quantile(prop, 0.025),
      hi_prop = quantile(prop, 0.975),
      .groups = "drop"
    )
  
  # Statistics summaries across simulations
  stat_summary <- moment_df %>%
    group_by(model, nx, ny) %>%
    summarise(
      mean_T_q05 = mean(T_q05, na.rm = TRUE),
      lo_T_q05 = quantile(T_q05, 0.025, na.rm = TRUE),
      hi_T_q05 = quantile(T_q05, 0.975, na.rm = TRUE),
  
      mean_T_peak = mean(T_peak, na.rm = TRUE),
      lo_T_peak = quantile(T_peak, 0.025, na.rm = TRUE),
      hi_T_peak = quantile(T_peak, 0.975, na.rm = TRUE),
  
      mean_T_iqr_width = mean(T_iqr_width, na.rm = TRUE),
      lo_T_iqr_width = quantile(T_iqr_width, 0.025, na.rm = TRUE),
      hi_T_iqr_width = quantile(T_iqr_width, 0.975, na.rm = TRUE),
  
      mean_T_median = mean(T_median, na.rm = TRUE),
      lo_T_median = quantile(T_median, 0.025, na.rm = TRUE),
      hi_T_median = quantile(T_median, 0.975, na.rm = TRUE),
  
      mean_T_q995 = mean(T_q995, na.rm = TRUE),
      lo_T_q995 = quantile(T_q995, 0.025, na.rm = TRUE),
      hi_T_q995 = quantile(T_q995, 0.975, na.rm = TRUE),
  
      mean_T_disp = mean(T_disp, na.rm = TRUE),
      lo_T_disp = quantile(T_disp, 0.025, na.rm = TRUE),
      hi_T_disp = quantile(T_disp, 0.975, na.rm = TRUE),
  
      .groups = "drop"
    )
  
  list(
    hist_df = hist_df,
    hist_summary = hist_summary,
    moment_df = moment_df,
    stat_summary = stat_summary,
    settings = list(
      lambda = lambda,
      phi = phi,
      s_cscp = s_cscp,
      s_lgcp = s_lgcp,
      grids = grids,
      n_sims = n_sims
    )
  )
}
```

```{r include=FALSE}
# Plotting shit, might need to revamp these for this section...
plot_quadrat_histograms_grid_scan <- function(study, max_k = NULL) {
  hist_sum <- study$hist_summary
  grids <- sort(unique(hist_sum$nx))
  n_grids <- length(grids)

  if (is.null(max_k)) {
    max_k_by_grid <- sapply(grids, function(g) {
      sub <- hist_sum[hist_sum$nx == g, ]
      keep <- with(sub, hi_prop > 0.002 | mean_prop > 0.001)
      if (!any(keep)) max(sub$k) else max(sub$k[keep])
    })
  } else {
    max_k_by_grid <- rep(max_k, n_grids)
  }

  global_ymax <- max(
    hist_sum$hi_prop[hist_sum$k <= max(max_k_by_grid)],
    na.rm = TRUE
  )

  n_col <- 3
  n_row <- ceiling(n_grids / n_col)

  op <- par(
    mfrow = c(n_row, n_col),
    mar = c(3, 3, 2, 1),
    oma = c(2, 2.5, 0, 0)
  )

  for (i in seq_along(grids)) {
    g <- grids[i]

    sub <- hist_sum[
      hist_sum$nx == g & hist_sum$k <= max_k_by_grid[i],
    ]

    lgcp <- sub[sub$model == "LGCP", ]
    cscp <- sub[sub$model == "CSCP", ]

    plot(
      range(sub$k), c(0, global_ymax),
      type = "n",
      axes = FALSE,
      xlab = "",
      ylab = "",
      main = paste0(g, " x ", g)
    )

    polygon(
      c(cscp$k, rev(cscp$k)),
      c(cscp$lo_prop, rev(cscp$hi_prop)),
      border = NA,
      col = adjustcolor("tomato", alpha.f = 0.2)
    )

    polygon(
      c(lgcp$k, rev(lgcp$k)),
      c(lgcp$lo_prop, rev(lgcp$hi_prop)),
      border = NA,
      col = adjustcolor("steelblue", alpha.f = 0.2)
    )

    lines(cscp$k, cscp$mean_prop, lwd = 2, col = "tomato")
    lines(lgcp$k, lgcp$mean_prop, lwd = 2, col = "steelblue")

    axis(1)
    axis(2)
    box()

    if (i == 1) {
      legend(
        "topright",
        legend = c("LGCP", "CSCP"),
        col = c("steelblue", "tomato"),
        lwd = 2,
        bty = "n"
      )
    }
  }

  mtext("Cell count", side = 1, outer = TRUE, line = 0.8)
  mtext("Proportion of cells", side = 2, outer = TRUE, line = 1.2)

  par(op)
}


plot_quadrat_statistics <- function(study) {
  ss <- study$stat_summary
  grids <- sort(unique(ss$nx))

  lgcp <- ss[ss$model == "LGCP", ]
  cscp <- ss[ss$model == "CSCP", ]

  op <- par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

  plot_one <- function(mean_col, lo_col, hi_col, ylab, main) {
    y_lim <- range(
      lgcp[[lo_col]], lgcp[[hi_col]],
      cscp[[lo_col]], cscp[[hi_col]],
      na.rm = TRUE
    )

    plot(
      grids, lgcp[[mean_col]],
      type = "b", pch = 16, lwd = 2, col = "steelblue",
      ylim = y_lim,
      xlab = "Grid size m",
      ylab = ylab,
      main = main
    )

    arrows(
      grids, lgcp[[lo_col]],
      grids, lgcp[[hi_col]],
      angle = 90, code = 3, length = 0.05,
      col = "steelblue"
    )

    lines(
      grids, cscp[[mean_col]],
      type = "b", pch = 16, lwd = 2, col = "tomato"
    )

    arrows(
      grids, cscp[[lo_col]],
      grids, cscp[[hi_col]],
      angle = 90, code = 3, length = 0.05,
      col = "tomato"
    )

    legend(
      "topright",
      legend = c("LGCP", "CSCP"),
      col = c("steelblue", "tomato"),
      lwd = 2,
      pch = 16,
      bty = "n"
    )
  }

  plot_one(
    "mean_T_q05", "lo_T_q05", "hi_T_q05",
    expression(T[0.05]), "5% quantile"
  )

  plot_one(
    "mean_T_peak", "lo_T_peak", "hi_T_peak",
    expression(T[peak]), "Peak height"
  )

  plot_one(
    "mean_T_iqr_width", "lo_T_iqr_width", "hi_T_iqr_width",
    expression(T[IQR]), "Interquartile width"
  )
  
  plot_one(
    "mean_T_median", "lo_T_median", "hi_T_median",
    expression(T[median]), "Median"
  )

  plot_one(
    "mean_T_q995", "lo_T_q995", "hi_T_q995",
    expression(T[0.995]), "99.5% quantile"
  )

  plot_one(
    "mean_T_disp", "lo_T_disp", "hi_T_disp",
    expression(T[disp]), "Index of dispersion"
  )

  par(op)
}

plot_lambda_histograms <- function(lambda_scan, grid_choice, max_k = NULL) {
  sub_all <- lambda_scan$hist_summary %>%
    dplyr::filter(nx == grid_choice)

  lambdas <- sort(unique(sub_all$lambda))

  get_max_k <- function(s) {
    if (!is.null(max_k)) return(max_k)

    keep <- with(s, hi_prop > 0.002 | mean_prop > 0.001)

    if (!any(keep)) {
      max(s$k, na.rm = TRUE)
    } else {
      max(s$k[keep], na.rm = TRUE)
    }
  }

  max_k_by_lambda <- sapply(lambdas, function(lam) {
    get_max_k(sub_all[sub_all$lambda == lam, ])
  })

  op <- par(
    mfrow = c(1, length(lambdas)),
    mar = c(3, 2, 2, 1),
    oma = c(2, 2.5, 0, 0)
  )

  for (i in seq_along(lambdas)) {
    lam <- lambdas[i]

    s <- sub_all %>%
      dplyr::filter(lambda == lam, k <= max_k_by_lambda[i])

    lgcp <- s[s$model == "LGCP", ]
    cscp <- s[s$model == "CSCP", ]

    ymax <- max(s$hi_prop, na.rm = TRUE)

    plot(
      range(s$k), c(0, ymax),
      type = "n",
      axes = FALSE,
      main = paste0(expression(lambda), " = ", lam),
      xlab = "",
      ylab = ""
    )

    polygon(
      c(cscp$k, rev(cscp$k)),
      c(cscp$lo_prop, rev(cscp$hi_prop)),
      col = adjustcolor("tomato", 0.2),
      border = NA
    )

    polygon(
      c(lgcp$k, rev(lgcp$k)),
      c(lgcp$lo_prop, rev(lgcp$hi_prop)),
      col = adjustcolor("steelblue", 0.2),
      border = NA
    )

    lines(cscp$k, cscp$mean_prop, col = "tomato", lwd = 2)
    lines(lgcp$k, lgcp$mean_prop, col = "steelblue", lwd = 2)

    axis(1)
    axis(2)
    box()

    if (i == 1) {
      legend(
        "topright",
        legend = c("LGCP", "CSCP"),
        col = c("steelblue", "tomato"),
        lwd = 2,
        bty = "n"
      )
    }
  }

  mtext("Cell count", side = 1, outer = TRUE, line = 0.8)
  mtext("Proportion of cells", side = 2, outer = TRUE, line = 1.2)

  par(op)
}

plot_lambda_stat_differences_all_grids <- function(
    lambda_scan,
    grid_choices = c(10, 30, 60)
) {
  ss <- lambda_scan$stat_summary %>%
    dplyr::filter(nx %in% grid_choices)

  stat_info <- list(
    list(mean = "mean_T_q05",        ylab = expression(Delta*T[0.05]),  main = "5% quantile"),
    list(mean = "mean_T_peak",       ylab = expression(Delta*T[peak]),  main = "Peak height"),
    list(mean = "mean_T_iqr_width",  ylab = expression(Delta*T[IQR]),   main = "Interquartile width"),
    list(mean = "mean_T_median",     ylab = expression(Delta*T[median]), main = "Median"),
    list(mean = "mean_T_q995",       ylab = expression(Delta*T[0.995]), main = "99.5% quantile"),
    list(mean = "mean_T_disp",       ylab = expression(Delta*T[disp]),  main = "Index of dispersion")
  )

  cols <- c("black", "grey40", "grey70")
  pchs <- c(16, 17, 15)

  op <- par(
    mfrow = c(2, 3),
    mar = c(3.0, 3.0, 1.0, 0.6),
    oma = c(2.6, 2.8, 1.2, 0.5),
    mgp = c(1.7, 0.5, 0),
    tcl = -0.25
  )
  on.exit(par(op))

  for (info in stat_info) {

    diff_list <- lapply(grid_choices, function(g) {
      lgcp <- ss %>%
        dplyr::filter(model == "LGCP", nx == g) %>%
        dplyr::arrange(lambda)

      cscp <- ss %>%
        dplyr::filter(model == "CSCP", nx == g) %>%
        dplyr::arrange(lambda)

      data.frame(
        nx = g,
        lambda = lgcp$lambda,
        diff = lgcp[[info$mean]] - cscp[[info$mean]]
      )
    })

    d_all <- do.call(rbind, diff_list)

    y_lim <- range(d_all$diff, 0, na.rm = TRUE)
    pad <- 0.08 * diff(y_lim)
    if (pad == 0) pad <- 1
    y_lim <- y_lim + c(-pad, pad)

    plot(
      range(d_all$lambda, na.rm = TRUE),
      y_lim,
      type = "n",
      axes = FALSE,
      xlab = "",
      ylab = "",
      main = ""
    )

    abline(h = 0, lty = 2, col = "grey60")

    for (ii in seq_along(grid_choices)) {
      g <- grid_choices[ii]
      d <- d_all[d_all$nx == g, ]

      lines(
        d$lambda,
        d$diff,
        type = "b",
        pch = pchs[ii],
        lwd = 2,
        col = cols[ii]
      )
    }

    usr <- par("usr")
    text(
      x = usr[1] + 0.04 * diff(usr[1:2]),
      y = usr[4] - 0.06 * diff(usr[3:4]),
      labels = info$main,
      adj = c(0, 1),
      cex = 0.9
    )

    axis(1)
    axis(2)
    box()
  }

  legend(
    "top",
    inset = -0.02,
    legend = paste0("m = ", grid_choices),
    col = cols,
    pch = pchs,
    lwd = 2,
    horiz = TRUE,
    bty = "n",
    xpd = NA
  )

  mtext(expression(lambda), side = 1, outer = TRUE, line = 1.4)
  mtext(expression(T[LGCP] - T[CSCP]), side = 2, outer = TRUE, line = 1.6)
}

plot_lambda_histograms_3x3 <- function(
    lambda_scan,
    grid_choices = c(10, 30, 60),
    lambda_choices = NULL,
    max_k = NULL
) {
  hist_sum <- lambda_scan$hist_summary

  if (is.null(lambda_choices)) {
    lambda_choices <- sort(unique(hist_sum$lambda))
  }

  hist_sum <- hist_sum %>%
    dplyr::filter(
      nx %in% grid_choices,
      lambda %in% lambda_choices
    )

  get_max_k <- function(s) {
    if (nrow(s) == 0) return(0)

    keep <- with(s, hi_prop > 0.002 | mean_prop > 0.001)

    if (!any(keep, na.rm = TRUE)) {
      max(s$k, na.rm = TRUE)
    } else {
      max(s$k[keep], na.rm = TRUE)
    }
  }

  max_k_lookup <- expand.grid(
    nx = grid_choices,
    lambda = lambda_choices
  )

  max_k_lookup$max_k <- mapply(function(g, lam) {
    s <- hist_sum[hist_sum$nx == g & hist_sum$lambda == lam, ]
    get_max_k(s)
  }, max_k_lookup$nx, max_k_lookup$lambda)

  ## Row-dependent y limits
  ymax_lookup <- data.frame(
    nx = grid_choices,
    ymax = NA_real_
  )

  for (ii in seq_along(grid_choices)) {
    g <- grid_choices[ii]
    row_ymax <- 0

    for (lam in lambda_choices) {
      mk <- max_k_lookup$max_k[
        max_k_lookup$nx == g &
          max_k_lookup$lambda == lam
      ]

      s <- hist_sum[
        hist_sum$nx == g &
          hist_sum$lambda == lam &
          hist_sum$k <= mk,
      ]

      if (nrow(s) > 0) {
        row_ymax <- max(row_ymax, s$hi_prop, na.rm = TRUE)
      }
    }

    ymax_lookup$ymax[ymax_lookup$nx == g] <- row_ymax
  }

  op <- par(
    mfrow = c(length(grid_choices), length(lambda_choices)),
    mar = c(1.4, 1.6, 0.4, 0.4),
    oma = c(3, 3.5, 1.2, 0.5),
    mgp = c(1.5, 0.35, 0),
    tcl = -0.25
  )

  on.exit(par(op))

  for (g in grid_choices) {
    for (lam in lambda_choices) {
      mk <- max_k_lookup$max_k[
        max_k_lookup$nx == g &
          max_k_lookup$lambda == lam
      ]

      s <- hist_sum[
        hist_sum$nx == g &
          hist_sum$lambda == lam &
          hist_sum$k <= mk,
      ]

      if (nrow(s) == 0 || all(!is.finite(s$k))) {
        plot.new()
        text(
          0.5, 0.95,
          paste0("m = ", g, ",  λ = ", lam, "\n(no data)"),
          adj = c(0.5, 1),
          cex = 0.85
        )
        next
      }

      lgcp <- s[s$model == "LGCP", ]
      cscp <- s[s$model == "CSCP", ]

      row_ymax <- ymax_lookup$ymax[ymax_lookup$nx == g]

      plot(
        range(s$k), c(0, row_ymax),
        type = "n",
        axes = FALSE,
        xlab = "",
        ylab = "",
        main = ""
      )

      polygon(
        c(cscp$k, rev(cscp$k)),
        c(cscp$lo_prop, rev(cscp$hi_prop)),
        col = adjustcolor("tomato", alpha.f = 0.2),
        border = NA
      )

      polygon(
        c(lgcp$k, rev(lgcp$k)),
        c(lgcp$lo_prop, rev(lgcp$hi_prop)),
        col = adjustcolor("steelblue", alpha.f = 0.2),
        border = NA
      )

      lines(cscp$k, cscp$mean_prop, col = "tomato", lwd = 2)
      lines(lgcp$k, lgcp$mean_prop, col = "steelblue", lwd = 2)

      ## Title inside plot window
      usr <- par("usr")
      text(
        x = usr[1] + 0.04 * diff(usr[1:2]),
        y = usr[4] - 0.06 * diff(usr[3:4]),
        labels = paste0("m = ", g, ",  λ = ", lam),
        adj = c(0, 1),
        cex = 1
      )

      axis(1)
      axis(2)
      box()
    }
  }

  legend(
    "bottomright",
    inset = -0.02,
    legend = c("LGCP", "CSCP"),
    col = c("steelblue", "tomato"),
    lwd = 2,
    horiz = TRUE,
    bty = "n",
    xpd = NA
  )

  mtext("Cell count", side = 1, outer = TRUE, line = 1.5)
  mtext("Proportion of cells", side = 2, outer = TRUE, line = 2.2)
}

plot_phi_histograms_3x3 <- function(
    phi_scan,
    grid_choices = c(10, 30, 60),
    phi_choices = NULL,
    max_k = NULL
) {
  hist_sum <- phi_scan$hist_summary

  if (is.null(phi_choices)) {
    phi_choices <- sort(unique(hist_sum$phi))
  }

  hist_sum <- hist_sum %>%
    dplyr::filter(
      nx %in% grid_choices,
      phi %in% phi_choices
    )

  get_max_k <- function(s) {
    if (nrow(s) == 0) return(0)
    if (!is.null(max_k)) return(max_k)

    keep <- with(s, hi_prop > 0.002 | mean_prop > 0.001)

    if (!any(keep, na.rm = TRUE)) {
      max(s$k, na.rm = TRUE)
    } else {
      max(s$k[keep], na.rm = TRUE)
    }
  }

  max_k_lookup <- expand.grid(
    nx = grid_choices,
    phi = phi_choices
  )

  max_k_lookup$max_k <- mapply(function(g, ph) {
    s <- hist_sum[hist_sum$nx == g & hist_sum$phi == ph, ]
    get_max_k(s)
  }, max_k_lookup$nx, max_k_lookup$phi)

  ymax_lookup <- data.frame(
    nx = grid_choices,
    ymax = NA_real_
  )

  for (ii in seq_along(grid_choices)) {
    g <- grid_choices[ii]
    row_ymax <- 0

    for (ph in phi_choices) {
      mk <- max_k_lookup$max_k[
        max_k_lookup$nx == g &
          max_k_lookup$phi == ph
      ]

      s <- hist_sum[
        hist_sum$nx == g &
          hist_sum$phi == ph &
          hist_sum$k <= mk,
      ]

      if (nrow(s) > 0) {
        row_ymax <- max(row_ymax, s$hi_prop, na.rm = TRUE)
      }
    }

    ymax_lookup$ymax[ymax_lookup$nx == g] <- row_ymax
  }

  op <- par(
    mfrow = c(length(grid_choices), length(phi_choices)),
    mar = c(1.4, 1.6, 0.4, 0.4),
    oma = c(3, 3.5, 1.2, 0.5),
    mgp = c(1.5, 0.35, 0),
    tcl = -0.25
  )
  on.exit(par(op))

  for (g in grid_choices) {
    for (ph in phi_choices) {
      mk <- max_k_lookup$max_k[
        max_k_lookup$nx == g &
          max_k_lookup$phi == ph
      ]

      s <- hist_sum[
        hist_sum$nx == g &
          hist_sum$phi == ph &
          hist_sum$k <= mk,
      ]

      if (nrow(s) == 0 || all(!is.finite(s$k))) {
        plot.new()
        text(
          0.5, 0.95,
          paste0("m = ", g, ",  phi = ", ph, "\n(no data)"),
          adj = c(0.5, 1),
          cex = 0.85
        )
        next
      }

      lgcp <- s[s$model == "LGCP", ]
      cscp <- s[s$model == "CSCP", ]

      row_ymax <- ymax_lookup$ymax[ymax_lookup$nx == g]

      plot(
        range(s$k), c(0, row_ymax),
        type = "n",
        axes = FALSE,
        xlab = "",
        ylab = "",
        main = ""
      )

      polygon(
        c(cscp$k, rev(cscp$k)),
        c(cscp$lo_prop, rev(cscp$hi_prop)),
        col = adjustcolor("tomato", alpha.f = 0.2),
        border = NA
      )

      polygon(
        c(lgcp$k, rev(lgcp$k)),
        c(lgcp$lo_prop, rev(lgcp$hi_prop)),
        col = adjustcolor("steelblue", alpha.f = 0.2),
        border = NA
      )

      lines(cscp$k, cscp$mean_prop, col = "tomato", lwd = 2)
      lines(lgcp$k, lgcp$mean_prop, col = "steelblue", lwd = 2)

      usr <- par("usr")
      text(
        x = usr[1] + 0.04 * diff(usr[1:2]),
        y = usr[4] - 0.06 * diff(usr[3:4]),
        labels = bquote(m == .(g) * "," ~~ phi == .(ph)),
        adj = c(0, 1),
        cex = 1
      )

      axis(1)
      axis(2)
      box()
    }
  }

  legend(
    "right",
    inset = -0.02,
    legend = c("LGCP", "CSCP"),
    col = c("steelblue", "tomato"),
    lwd = 2,
    horiz = TRUE,
    bty = "n",
    xpd = NA
  )

  mtext("Cell count", side = 1, outer = TRUE, line = 1.5)
  mtext("Proportion of cells", side = 2, outer = TRUE, line = 2.2)
}
```

*Choosing quadrat resolutions for testing?*

:::callout-note
I think the primary goals of this section should be:

1. It is not immediately clear whether there exists a single grid size for any given 
parameter setting, which provides the best separation between the two curves

2. Different grid sizes emphasise the different aspects of the seperation

And I guess secondary goals might be:

1. Demonstrating *how* "turning the knobs" effects the two curves

2. Demonstrating how the scale parameter effects the two curves
:::

## Aim

The goal of this section is to explore how the quadrat-count distribution is
affected by both the parameters of the latent field and the grid resolution used.

In particular, we investigate how changes in clustering strength, clustering
scale, intensity and grid size alter the separation between the CSCP and LGCP
quadrat-count distributions.

This section is primarily visual and exploratory. Its purpose is to identify
which parameter regimes and grid resolutions appear most promising for
distinguishing the two models, before carrying out a more principled assessment
of classification power in the following section.

## Effect of grid resolution

Given a grid of $m \times m$ quadrats, we obtain a collection of counts

$$
\{N_1, \dots, N_J\}, \quad J = m^2,
$$

where each $N_j$ represents the number of points observed in quadrat $B_j$.

As in the last section, we summarise these counts through their empirical 
distribution, that is, the proportion of quadrats with count $k$,

$$
\hat p_m(k) = \frac{1}{J} \sum_{j=1}^J \mathbf{1}\{N_j = k\}.
$$

This provides a discrete approximation to the distribution of local point
intensity at the scale determined by the grid.

The choice of grid resolution plays a key role in this representation:

- *Finer grids* (large $ m $) produce many small quadrats, leading to
  noisier counts that reflect local variability in the process.

- *Coarser grids* (small $ m $) aggregate over larger regions, smoothing
  local fluctuations and emphasising broader clustering structure.

As a result, differences between the LGCP and CSCP may appear more clearly at
some grid resolutions than others.

## Grid resolution setup

We begin by fixing the model parameters and varying only the quadrat grid
resolution. The aim is to isolate the effect of grid resolution on the quadrat
count distribution, and to understand how the apparent differences between the
LGCP and CSCP change across scales.

We use three different grid resolutions to reflect a coarse, intermediate and fine
partitioning of the window:

- **Coarse** - 15x15

- **Intermediate** - 30x30

- **Fine** - 45x45

As in the last section, we also fix 

$$
\lambda = 25000, \qquad \phi = 1, \qquad s_{\text{CSCP}} = 0.1,
$$

and use the corresponding matched LGCP scale parameter. As in the last section,
these parameter settings are chosen to represent a regime where clustering is clearly 
present, and where the expected number of points is sufficiently large to ensure 
that quadrat counts are informative across a range of grid resolutions.

An example of two of the point patterns, partitioned into quadrats is provided 
below:

```{r echo=FALSE}
example_lambda <- 25000
example_phi <- 1
example_s_cscp <- 0.05
example_s_lgcp <- match_lgcp_scale(phi = example_phi, s_cscp = example_s_cscp)

W <- owin(c(0, 1), c(0, 1))
set.seed(1)

ex_lgcp <- sim_lgcp(
  W = W,
  lambda = example_lambda,
  phi = example_phi,
  s = example_s_lgcp
)$pp

ex_cscp <- sim_cscp(
  W = W,
  lambda = example_lambda,
  phi = example_phi,
  s = example_s_cscp,
  df = 1
)$pp

par(mfrow = c(1, 2), mar = c(0,0,2,0))
plot(ex_cscp, pch = 16, cex = 0.35, main = "CSCP")
plot(ex_lgcp, pch = 16, cex = 0.35, main = "LGCP")
```

We run the study as below:

```{r cache=TRUE}
grid_study <- run_quadrat_count_study(
  W = owin(),
  lambda = 25000,
  phi = 1,
  s_cscp = 0.1,
  grids = c(15, 30, 45),
  n_sims = 200,
  seed = 123
)
```

The following figure shows the mean quadrat count distributions for each grid
resolution, with pointwise Monte Carlo intervals.

```{r echo=FALSE}
plot_quadrat_histograms_grid_scan(grid_study)
```

To aid interpretation, we also track several scalar summaries of the count
distribution.

```{r echo=FALSE}
plot_quadrat_statistics(grid_study)
```

Several clear patterns emerge.

As the grid becomes finer, the average number of points per quadrat decreases.
This shifts the count distributions toward smaller values and increases the
importance of discreteness. At the finest resolutions, most quadrats contain 
only a small number of points, often close to zero or one. This makes the 
distributions more sensitive to random variation, and less stable across simulations.

At the coarsest resolutions, quadrat counts are large, but each cell averages
over a substantial region of the intensity surface. Because each quadrat covers 
a larger region, the counts average over local fluctuations in the intensity. 
This makes the two distributions appear more similar overall.

Between these extremes, at the intermediate grid resolution, the differences 
between the LGCP and CSCP appear most clearly visible (subjective, and quite 
strong statement, not sure if justified tbh). In particular:

- Moderate grid resolution shows differences in the overall shape of the
  distribution, including peak height and spread.

- Coarser grid resolutions can produce larger raw differences in lower-tail
  (near-zero) summaries, but these differences are accompanied by substantially 
  greater Monte Carlo variability.

This highlights an important point: large differences alone are not sufficient
for choosing a grid resolution. A useful grid should produce differences that
are both substantial and stable across simulations.

Taken together, these results suggest that different grid resolutions highlight
different features of the quadrat count distribution. While intermediate grids
appear to offer a useful balance between signal and stability, the evidence here
is not sufficient to identify a single clearly optimal choice.

A more formal comparison of classification performance across grid resolutions
is therefore required, which we undertake in the following section.

We now investigate how this behaviour changes as the model parameters vary. In
particular, we now examine the effect of overall intensity $\lambda$.

## Effect of intensity

We now investigate how the effectiveness of different grid resolutions depends
on the overall intensity $\lambda$.

This is important because the quadrat count distribution depends not only on
the structure of the intensity field, but also on the expected number of points
per quadrat. For a grid with $m \times m$ cells over a window $W$, the
expected number of points per quadrat is approximately

$$
\bar n_q = \frac{\lambda |W|}{m^2}
$$

Thus, the same grid resolution can correspond to very different distributions
depending on the value of $\lambda$. A grid that is too fine at low intensity
may become appropriate at higher intensity, as the counts per quadrat increase.

```{r include=FALSE}
run_quadrat_lambda_scan <- function(
    W,
    lambda_vals,
    phi = 1,
    s_cscp = 0.1,
    grids = c(15, 30, 45),
    n_sims = 200,
    seed = NULL
) {
  hist_out <- vector("list", length(lambda_vals))
  stat_out <- vector("list", length(lambda_vals))
  settings_out <- vector("list", length(lambda_vals))

  for (i in seq_along(lambda_vals)) {
    lambda_i <- lambda_vals[i]

    study_i <- run_quadrat_count_study(
      W = W,
      lambda = lambda_i,
      phi = phi,
      s_cscp = s_cscp,
      grids = grids,
      n_sims = n_sims,
      seed = if (is.null(seed)) NULL else seed + i - 1
    )

    hist_out[[i]] <- study_i$hist_summary %>%
      mutate(
        lambda = lambda_i,
        phi = phi,
        s_cscp = s_cscp,
        s_lgcp = study_i$settings$s_lgcp
      )

    stat_out[[i]] <- study_i$stat_summary %>%
      mutate(
        lambda = lambda_i,
        phi = phi,
        s_cscp = s_cscp,
        s_lgcp = study_i$settings$s_lgcp
      )

    settings_out[[i]] <- study_i$settings
  }

  list(
    hist_summary = bind_rows(hist_out),
    stat_summary = bind_rows(stat_out),
    settings = settings_out
  )
}
```

```{r cache=TRUE}
lambda_scan <- run_quadrat_lambda_scan(
  W = owin(),
  lambda_vals = c(5000, 25000, 100000),
  phi = 1,
  s_cscp = 0.1,
  grids = c(15, 30, 45),
  n_sims = 200,
  seed = 123
)
```

```{r echo=FALSE}
plot_lambda_histograms_3x3(
  lambda_scan,
  grid_choices = c(15, 30, 45),
  lambda_choices = c(5000, 25000, 100000)
)
```

```{r echo=FALSE}
# plot_lambda_stat_differences_all_grids(
#   lambda_scan,
#   grid_choices = c(15, 30, 60)
# )
```

As $\lambda$ increases, the expected number of points per quadrat increases
for any fixed grid resolution. This reduces discreteness and stabilises the
quadrat count distributions, making differences between the LGCP and CSCP more
clearly visible.

At low values of $\lambda$, many quadrats contain few points. In this regime,
Poisson variability dominates, and the differences between the two models are
more difficult to detect, particularly for finer grids.

As $\lambda$ increases, the same grid resolutions become more informative.
Differences in both the lower tail and the overall shape of the distribution
become more stable, and the Monte Carlo variability decreases.

This shows that grid resolution cannot be chosen independently of $\lambda$.
Instead, it should be interpreted relative to the expected number of points per
quadrat. In particular, useful grid resolutions are those for which quadrat
counts are sufficiently large to avoid excessive discreteness, but not so large
that spatial averaging removes local structure.

These results suggest that the appropriate grid resolution becomes finer as
$\lambda$ increases. Equivalently, grid choice should be guided by the
expected count per quadrat, rather than the absolute number of grid cells.

:::callout-note
This is so mind numbing at this point. I dont even see the point if I'm being honest, 
I would rather just jump straight to a study of power. 

I understand we have to motivate our choices for candidate grid resolutions, but
I'm so bored of plotting these and adding commentary jeesus
:::

## How this changes with $\phi$

We next investigate how the effectiveness of different grid resolutions changes
with the clustering strength $\phi$. Here $\lambda$ and
$s_{\text{CSCP}}$ are fixed, while $\phi$ is varied.

This is important because $\phi$ controls the strength of second-order
clustering in both models. Larger values of $\phi$ produce stronger local
variation in the intensity field, which may make differences between the LGCP
and CSCP count distributions easier to detect.

For each value of $\phi$, the LGCP scale parameter is re-matched to the CSCP
scale parameter so that the models remain comparable at the level of the pair
correlation function.

blah blah blah you get the idea...

```{r include=FALSE}
run_quadrat_phi_scan <- function(
    W,
    phi_vals,
    lambda = 25000,
    s_cscp = 0.1,
    grids = c(10, 30, 60),
    n_sims = 200,
    seed = NULL
) {
  hist_out <- vector("list", length(phi_vals))
  stat_out <- vector("list", length(phi_vals))
  settings_out <- vector("list", length(phi_vals))

  for (i in seq_along(phi_vals)) {
    phi_i <- phi_vals[i]

    study_i <- run_quadrat_count_study(
      W = W,
      lambda = lambda,
      phi = phi_i,
      s_cscp = s_cscp,
      grids = grids,
      n_sims = n_sims,
      seed = if (is.null(seed)) NULL else seed + i - 1
    )

    hist_out[[i]] <- study_i$hist_summary %>%
      mutate(
        lambda = lambda,
        phi = phi_i,
        s_cscp = s_cscp,
        s_lgcp = study_i$settings$s_lgcp
      )

    stat_out[[i]] <- study_i$stat_summary %>%
      mutate(
        lambda = lambda,
        phi = phi_i,
        s_cscp = s_cscp,
        s_lgcp = study_i$settings$s_lgcp
      )

    settings_out[[i]] <- study_i$settings
  }

  list(
    hist_summary = bind_rows(hist_out),
    stat_summary = bind_rows(stat_out),
    settings = settings_out
  )
}
```

```{r cache=TRUE}
phi_scan <- run_quadrat_phi_scan(
  W = owin(),
  phi_vals = c(0.5, 1.25, 2),
  lambda = 25000,
  s_cscp = 0.1,
  grids = c(15, 30, 45),
  n_sims = 200,
  seed = 456
)
```

```{r echo=FALSE}
plot_phi_histograms_3x3(
  phi_scan,
  grid_choices = c(15, 30, 45),
  phi_choices = c(0.5, 1.25, 2)
)
```

Same story again: no clear best grid resolution, phi has a noticable effect, 
yada yada yada.

## Brief scale check

Will need to demonstrate how scale effects things, I'm sick of slaving away here.

Just too much plotting code, slowly osing my sanity. Will finish at a later date.

## Key takeaways

The results of this section show that it is unclear whether there is a single 
quadrat resolution that is uniformly optimal across parameter settings.

Different grid resolutions highlight different features of the quadrat count
distribution:

- Moderate grids reveal differences in overall shape, including peak height
  and spread.

- Coarser grids tend to exaggerate differences in the lower tail, but these
  effects are often less stable across simulations.

- Finer grids produce many small counts, making the distributions more sensitive
  to random variation unless the intensity is sufficiently large.

Moreover, the effectiveness of a given grid depends on the underlying parameters,
particularly the intensity $\lambda$, through the expected number of points per
quadrat.

Taken together, these results suggest that selecting a single "optimal" grid
resolution is not appropriate. Instead, grid choice should be treated as part of
the design of the test statistic itself.

In practice, one could tailor the grid choice to a fitted model by estimating
parameters and selecting the grid (or set of grids) with the highest power for
that regime. However, this introduces additional complexity and is not considered
further here.

Instead, we aim to select a fixed set of grid resolutions that performs well
across a range of clustering regimes, and use this set for the remainder of the
thesis. In the next section, we conduct a formal power study to assess the
performance of candidate grid choices and determine an appropriate selection.

:::callout-note
Think this is slightly too verbose.

Basically, we will be testing a *bunch* of candidates, and seeing how they perform
across multiple parameter regimes. 

Then for this thesis, we will choose the one which seems the most robust across 
all regimes.

In practice, it would be better to fit your model, and then do a power-test to 
figure out which grid set has the greatest power for your fitted model, and then
use this in your MC hypothesis test. 

Note that it is important if you were to perform the grid-selection using a power
study, that you only use the fitted parameters, and never compare grid performance
on the observed data.
:::




