```{r include=FALSE}
# All the code from last section, I know I should make this include, sue me AGAIN

# Calculate quadrat counts using spatstat's quadratcount()
# Example of splitting observation window into 2x2 grid:
#
# > get_quadrat_counts(X_lgcp, 2)
# [1] 32 21 15 28
get_quadrat_counts <- function(X, nx, ny = nx) {
  qc <- quadratcount(X, nx = nx, ny = ny)
  as.numeric(qc)
}

# Calculate sample skewness summary
# Making sure to return NA in the case that all counts 
#       are the same (sample variance = 0)
safe_skewness <- function(x) {
  x <- as.numeric(x)
  if (length(x) < 2) stop("Require at least 2 quadrats")
  
  s2 <- stats::var(x)
  if (!is.finite(s2) || s2 <= 0) return(NA_real_)
  
  # Third moment calculation
  # m3 = mean((N_j - mean(N))^3)
  m3 <- mean((x - mean(x))^3)
  
  # Return standardised skewness
  # gamma1 = m3 / s^3 (s sample SD)
  m3 / (sqrt(s2)^3)
}


# Saves the counts vector from above as a dataframe, with extra sim-info 
# Example with counts = [1] 32 21 15 28:
#
# > count_hist_df(test, "lgcp", 1, 2)
# # A tibble: 4 × 6
#   model   rep    nx    ny     k  freq
#   <chr> <dbl> <dbl> <dbl> <int> <dbl>
# 1 lgcp      1     2     2    15     1
# 2 lgcp      1     2     2    21     1
# 3 lgcp      1     2     2    28     1
# 4 lgcp      1     2     2    32     1
count_hist_df <- function(counts, model, rep, nx, ny = nx) {
  # Tabulating to store more compactly
  # Helps in the case we are using very fine grid resolution
  tab <- table(counts)
  
  tibble(
    model = model, # "LGCP" or "CSCP"
    rep = rep, # replicate number
    nx = nx, # grid resolution
    ny = ny, # grid resolution
    k = as.integer(names(tab)), # The unique count values
    freq = as.numeric(tab) # How many quadrats had each of the above count values
  )
}

# Same as above, but instead of saving the raw counts, it saves the stats info
count_stat_df <- function(counts, model, rep, nx, ny = nx) {
  counts <- as.numeric(counts)
  J <- length(counts)

  # Empirical count distribution
  tab <- table(counts)
  props <- as.numeric(tab) / J
  k_vals <- as.integer(names(tab))

  # Peak height and peak width
  mode_index <- which.max(props)
  T_peak <- props[mode_index]

  # Interquartile width
  q25 <- as.numeric(quantile(counts, 0.25, type = 1))
  q75 <- as.numeric(quantile(counts, 0.75, type = 1))
  T_iqr_width <- q75 - q25

  # Lower / central / upper quantiles
  T_q05 <- as.numeric(quantile(counts, 0.05, type = 1))
  T_median <- as.numeric(quantile(counts, 0.50, type = 1))
  T_q995 <- as.numeric(quantile(counts, 0.995, type = 1))

  # Baseline dispersion
  mean_count <- mean(counts)
  var_count <- stats::var(counts)
  T_disp <- if (mean_count > 0) var_count / mean_count else NA_real_

  tibble(
    model = model,
    rep = rep,
    nx = nx,
    ny = ny,
    T_q05 = T_q05,
    T_peak = T_peak,
    T_iqr_width = T_iqr_width,
    T_median = T_median,
    T_q995 = T_q995,
    T_disp = T_disp
  )
}

run_quadrat_count_study <- function(
    W, # Observation window
    lambda, 
    phi,
    s_cscp = 0.05, # LGCP scale will be best-matching
    grids = c(4, 8, 16, 32),
    n_sims = 200,
    seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)
  
  # Match LGCP scale to CSCP scale
  s_lgcp <- match_lgcp_scale(phi = phi, s_cscp = s_cscp)
  
  # Objects for storing histograms and moments
  hist_list <- vector("list", length = 2 * n_sims * length(grids))
  moment_list <- vector("list", length = 2 * n_sims * length(grids))
  
  # Index is just for keeping calculated counts and summaries together
  idx <- 1
  
  # Study loop
  for (rep in seq_len(n_sims)) {
    # Simulate one LGCP pattern
    X_lgcp <- sim_lgcp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_lgcp
    )$pp
    
    # Simulate one CSCP pattern
    X_cscp <- sim_cscp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_cscp,
      df = 1
    )$pp
    
    # For each grid size
    for (m in grids) {
      # Calculate the counts per grid-tile
      counts_lgcp <- get_quadrat_counts(X_lgcp, nx = m, ny = m)
      counts_cscp <- get_quadrat_counts(X_cscp, nx = m, ny = m)
      
      # Keep track of the counts data plus summaries for each process
      hist_list[[idx]] <- count_hist_df(counts_lgcp, model = "LGCP", rep = rep, nx = m)
      moment_list[[idx]] <- count_stat_df(counts_lgcp, model = "LGCP", rep = rep, nx = m)
      idx <- idx + 1
      
      hist_list[[idx]] <- count_hist_df(counts_cscp, model = "CSCP", rep = rep, nx = m)
      moment_list[[idx]] <- count_stat_df(counts_cscp, model = "CSCP", rep = rep, nx = m)
      idx <- idx + 1
    }
  }
  
  # Combine all the data
  hist_df <- bind_rows(hist_list)
  moment_df <- bind_rows(moment_list)
  
  # Complete missing count values so summaries are comparable
  # For example if LGCP has counts up to 100 but CSCP only to 50
  # Then the below code sets all of CSCP counts from 51 to 100 = 0
  hist_df <- hist_df %>%
    group_by(nx, ny) %>%
    tidyr::complete(model, rep, k = 0:max(k), fill = list(freq = 0)) %>%
    ungroup()
  
  # Calculate proportion of each count within each replicate/grid
  hist_df <- hist_df %>%
    group_by(model, rep, nx, ny) %>%
    mutate(prop = freq / sum(freq)) %>%
    ungroup()
  
  # Histogram envelopes across simulations
  hist_summary <- hist_df %>%
    group_by(model, nx, ny, k) %>%
    summarise(
      mean_prop = mean(prop),
      med_prop = median(prop),
      lo_prop = quantile(prop, 0.025),
      hi_prop = quantile(prop, 0.975),
      .groups = "drop"
    )
  
  # Statistics summaries across simulations
  stat_summary <- moment_df %>%
    group_by(model, nx, ny) %>%
    summarise(
      mean_T_q05 = mean(T_q05, na.rm = TRUE),
      lo_T_q05 = quantile(T_q05, 0.025, na.rm = TRUE),
      hi_T_q05 = quantile(T_q05, 0.975, na.rm = TRUE),
  
      mean_T_peak = mean(T_peak, na.rm = TRUE),
      lo_T_peak = quantile(T_peak, 0.025, na.rm = TRUE),
      hi_T_peak = quantile(T_peak, 0.975, na.rm = TRUE),
  
      mean_T_iqr_width = mean(T_iqr_width, na.rm = TRUE),
      lo_T_iqr_width = quantile(T_iqr_width, 0.025, na.rm = TRUE),
      hi_T_iqr_width = quantile(T_iqr_width, 0.975, na.rm = TRUE),
  
      mean_T_median = mean(T_median, na.rm = TRUE),
      lo_T_median = quantile(T_median, 0.025, na.rm = TRUE),
      hi_T_median = quantile(T_median, 0.975, na.rm = TRUE),
  
      mean_T_q995 = mean(T_q995, na.rm = TRUE),
      lo_T_q995 = quantile(T_q995, 0.025, na.rm = TRUE),
      hi_T_q995 = quantile(T_q995, 0.975, na.rm = TRUE),
  
      mean_T_disp = mean(T_disp, na.rm = TRUE),
      lo_T_disp = quantile(T_disp, 0.025, na.rm = TRUE),
      hi_T_disp = quantile(T_disp, 0.975, na.rm = TRUE),
  
      .groups = "drop"
    )
  
  list(
    hist_df = hist_df,
    hist_summary = hist_summary,
    moment_df = moment_df,
    stat_summary = stat_summary,
    settings = list(
      lambda = lambda,
      phi = phi,
      s_cscp = s_cscp,
      s_lgcp = s_lgcp,
      grids = grids,
      n_sims = n_sims
    )
  )
}
```

# Quadrats 4

The pilot study above suggests that the grid strategy

$$
\mathcal E = \{2,5,10,20,40\}
$$

provides a robust default among the candidate strategies considered. However,
this should not be interpreted as universally optimal. In practice, the power of
the procedure should be assessed under the fitted parameter values, since the
most informative quadrat scales may depend on the intensity, clustering
strength, and correlation range.

We now use the same quadrat-distance idea to define a formal Monte Carlo test.

## Test statistic

Let $M_0$ denote the null model and $M_1$ denote the alternative model. For an
observed point pattern $X$, define

$$
T(X) = D(X;M_0) - D(X;M_1).
$$

Large positive values of $T(X)$ indicate that the observed pattern is closer to
the alternative model than to the null model.

For example, when testing

$$
H_0: X \sim \text{CSCP}
\qquad \text{against} \qquad
H_1: X \sim \text{LGCP},
$$

we use

$$
T(X) = D(X;\text{CSCP}) - D(X;\text{LGCP}),
$$

and reject the CSCP null for large positive values.

## Helper functions

```{r include=FALSE}
expected_count_to_grid <- function(W, lambda, eta) {
  area_W <- spatstat.geom::area.owin(W)
  m <- round(sqrt(lambda * area_W / eta))
  pmax(1, unique(m))
}

make_grid_sets_from_eta <- function(W, lambda, eta_sets) {
  purrr::map(eta_sets, ~ expected_count_to_grid(W, lambda, .x))
}

make_default_grids <- function(W, lambda, eta = c(2, 5, 10, 20, 40)) {
  expected_count_to_grid(W, lambda, eta)
}
```

```{r}
quadrat_distance_to_model <- function(X, ref_curves, grids, model_name) {
  total_dist <- 0

  for (m in grids) {
    q <- spatstat.geom::quadratcount(X, nx = m, ny = m)
    counts <- as.vector(q)

    obs_tab <- tibble::tibble(k = counts) |>
      dplyr::count(k, name = "n") |>
      dplyr::mutate(prop = n / sum(n))

    ref_tab <- ref_curves |>
      dplyr::filter(model == model_name, nx == m, ny == m) |>
      dplyr::select(k, ref_prop = mean_prop)

    dist_m <- dplyr::full_join(obs_tab, ref_tab, by = "k") |>
      dplyr::mutate(
        prop = tidyr::replace_na(prop, 0),
        ref_prop = tidyr::replace_na(ref_prop, 0)
      ) |>
      dplyr::summarise(dist = sum(abs(prop - ref_prop))) |>
      dplyr::pull(dist)

    total_dist <- total_dist + dist_m
  }

  total_dist
}
```

```{r}
quadrat_test_statistic <- function(X, ref_curves, grids, null_model, alt_model) {
  d_null <- quadrat_distance_to_model(
    X = X,
    ref_curves = ref_curves,
    grids = grids,
    model_name = null_model
  )

  d_alt <- quadrat_distance_to_model(
    X = X,
    ref_curves = ref_curves,
    grids = grids,
    model_name = alt_model
  )

  d_null - d_alt
}
```

## Monte Carlo p-value

```{r}
simulate_from_model <- function(model, W, lambda, phi, s_cscp, s_lgcp) {
  if (model == "CSCP") {
    sim_cscp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_cscp,
      df = 1
    )$pp
  } else if (model == "LGCP") {
    sim_lgcp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_lgcp
    )$pp
  } else {
    stop("model must be either 'CSCP' or 'LGCP'")
  }
}
```

```{r}
run_quadrat_mc_test <- function(
    X_obs,
    W,
    lambda,
    phi,
    s_cscp,
    null_model = "CSCP",
    alt_model = "LGCP",
    eta = c(2, 5, 10, 20, 40),
    n_ref = 300,
    n_mc = 999,
    seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  grids <- make_default_grids(W, lambda, eta)

  ref_study <- run_quadrat_count_study(
    W = W,
    lambda = lambda,
    phi = phi,
    s_cscp = s_cscp,
    grids = grids,
    n_sims = n_ref,
    seed = seed
  )

  ref_curves <- ref_study$hist_summary
  s_lgcp <- ref_study$settings$s_lgcp

  T_obs <- quadrat_test_statistic(
    X = X_obs,
    ref_curves = ref_curves,
    grids = grids,
    null_model = null_model,
    alt_model = alt_model
  )

  T_null <- numeric(n_mc)

  for (i in seq_len(n_mc)) {
    X_i <- simulate_from_model(
      model = null_model,
      W = W,
      lambda = lambda,
      phi = phi,
      s_cscp = s_cscp,
      s_lgcp = s_lgcp
    )

    T_null[i] <- quadrat_test_statistic(
      X = X_i,
      ref_curves = ref_curves,
      grids = grids,
      null_model = null_model,
      alt_model = alt_model
    )
  }

  p_value <- (1 + sum(T_null >= T_obs)) / (n_mc + 1)

  list(
    T_obs = T_obs,
    T_null = T_null,
    p_value = p_value,
    grids = grids,
    eta = eta,
    null_model = null_model,
    alt_model = alt_model,
    settings = list(
      lambda = lambda,
      phi = phi,
      s_cscp = s_cscp,
      s_lgcp = s_lgcp,
      n_ref = n_ref,
      n_mc = n_mc
    )
  )
}
```

## Example test

```{r}
W <- spatstat.geom::owin(c(0, 1), c(0, 1))

lambda <- 5000
phi <- 1
s_cscp <- 0.1

s_lgcp <- match_lgcp_scale(phi = phi, s_cscp = s_cscp)

X_obs <- sim_lgcp(
  W = W,
  lambda = lambda,
  phi = phi,
  s = s_lgcp
)$pp

par(mar = c(0,0,2,0))
plot(X_obs, pch = 16, cex = 0.45)
```

```{r}
mc_test <- run_quadrat_mc_test(
  X_obs = X_obs,
  W = W,
  lambda = lambda,
  phi = phi,
  s_cscp = s_cscp,
  null_model = "CSCP",
  alt_model = "LGCP",
  eta = c(2, 5, 10, 20, 40),
  n_ref = 300,
  n_mc = 100,
  seed = 123
)

mc_test$p_value
mc_test$grids
```

```{r}
hist(
  mc_test$T_null,
  breaks = 30,
  main = "Monte Carlo null distribution",
  xlab = "Test statistic"
)

abline(v = mc_test$T_obs, lwd = 2, lty = 2)
```

The dashed vertical line shows the observed test statistic. If this value lies in
the upper tail of the Monte Carlo null distribution, this provides evidence
against the null model.

# Power study

The Monte Carlo test gives a formal p-value for one observed point pattern. We
now use repeated simulation to estimate the power of the test under different
parameter regimes.

For a fixed regime, power is estimated by simulating point patterns from the
alternative model and recording the proportion of times the null hypothesis is
rejected at level $\alpha = 0.05$.

```{r}
run_quadrat_power_one_regime <- function(
    W,
    lambda,
    phi,
    s_cscp,
    null_model = "CSCP",
    alt_model = "LGCP",
    eta = c(2, 5, 10, 20, 40),
    alpha = 0.05,
    n_ref = 300,
    n_mc = 499,
    n_power = 100,
    seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  grids <- make_default_grids(W, lambda, eta)

  ref_study <- run_quadrat_count_study(
    W = W,
    lambda = lambda,
    phi = phi,
    s_cscp = s_cscp,
    grids = grids,
    n_sims = n_ref,
    seed = seed
  )

  ref_curves <- ref_study$hist_summary
  s_lgcp <- ref_study$settings$s_lgcp

  T_null <- numeric(n_mc)

  for (i in seq_len(n_mc)) {
    X_i <- simulate_from_model(
      model = null_model,
      W = W,
      lambda = lambda,
      phi = phi,
      s_cscp = s_cscp,
      s_lgcp = s_lgcp
    )

    T_null[i] <- quadrat_test_statistic(
      X = X_i,
      ref_curves = ref_curves,
      grids = grids,
      null_model = null_model,
      alt_model = alt_model
    )
  }

  p_values <- numeric(n_power)

  for (j in seq_len(n_power)) {
    X_alt <- simulate_from_model(
      model = alt_model,
      W = W,
      lambda = lambda,
      phi = phi,
      s_cscp = s_cscp,
      s_lgcp = s_lgcp
    )

    T_alt <- quadrat_test_statistic(
      X = X_alt,
      ref_curves = ref_curves,
      grids = grids,
      null_model = null_model,
      alt_model = alt_model
    )

    p_values[j] <- (1 + sum(T_null >= T_alt)) / (n_mc + 1)
  }

  tibble::tibble(
    lambda = lambda,
    phi = phi,
    s_cscp = s_cscp,
    s_lgcp = s_lgcp,
    null_model = null_model,
    alt_model = alt_model,
    eta = paste(eta, collapse = ", "),
    grids = paste(grids, collapse = ", "),
    alpha = alpha,
    n_ref = n_ref,
    n_mc = n_mc,
    n_power = n_power,
    power = mean(p_values <= alpha),
    mean_p_value = mean(p_values),
    median_p_value = median(p_values)
  )
}
```

## Scan over intensity

```{r}
lambda_grid <- c(500, 1000, 2000, 5000, 10000, 20000)

power_lambda_scan <- purrr::map_dfr(seq_along(lambda_grid), function(i) {
  run_quadrat_power_one_regime(
    W = W,
    lambda = lambda_grid[i],
    phi = 1,
    s_cscp = 0.1,
    null_model = "CSCP",
    alt_model = "LGCP",
    eta = c(2, 5, 10, 20, 40),
    alpha = 0.05,
    n_ref = 300,
    n_mc = 499,
    n_power = 100,
    seed = 1000 + i
  )
})

power_lambda_scan
```

```{r}
plot(
  power_lambda_scan$lambda,
  power_lambda_scan$power,
  type = "b",
  log = "x",
  ylim = c(0, 1),
  xlab = expression(lambda),
  ylab = "Estimated power",
  main = "Power of quadrat-distance test"
)

abline(h = 0.95, lty = 2)
```

```{r}
min_lambda_95 <- power_lambda_scan |>
  dplyr::filter(power >= 0.95) |>
  dplyr::summarise(min_lambda = min(lambda)) |>
  dplyr::pull(min_lambda)

min_lambda_95
```

The smallest value of $\lambda$ for which the estimated power exceeds $0.95$
gives an approximate minimum intensity required for reliable discrimination
under this parameter regime.

# Recommended practical workflow

The results above suggest the following practical procedure.

1. Fit both the CSCP and LGCP to the observed point pattern.
2. Match the models using the fitted first- and second-order properties.
3. Choose a default grid strategy, such as
   $\mathcal E = \{2,5,10,20,40\}$.
4. Simulate reference quadrat-count distributions under both fitted models.
5. Compute the quadrat-distance test statistic.
6. Calibrate the statistic using Monte Carlo simulation under the null model.
7. Estimate power under the fitted alternative model.

The pilot study provides a robust default grid strategy, but the final power of
the test should always be assessed for the fitted parameter regime.
