```{r include=FALSE}
devtools::load_all()
library(dplyr)
library(purrr)
library(tidyr)
library(spatstat)
```
# Quadrats 3

```{r include=FALSE}
# All the code from last section, I know I should make this include, sue me

# Calculate quadrat counts using spatstat's quadratcount()
# Example of splitting observation window into 2x2 grid:
#
# > get_quadrat_counts(X_lgcp, 2)
# [1] 32 21 15 28
get_quadrat_counts <- function(X, nx, ny = nx) {
  qc <- quadratcount(X, nx = nx, ny = ny)
  as.numeric(qc)
}

# Calculate sample skewness summary
# Making sure to return NA in the case that all counts 
#       are the same (sample variance = 0)
safe_skewness <- function(x) {
  x <- as.numeric(x)
  if (length(x) < 2) stop("Require at least 2 quadrats")
  
  s2 <- stats::var(x)
  if (!is.finite(s2) || s2 <= 0) return(NA_real_)
  
  # Third moment calculation
  # m3 = mean((N_j - mean(N))^3)
  m3 <- mean((x - mean(x))^3)
  
  # Return standardised skewness
  # gamma1 = m3 / s^3 (s sample SD)
  m3 / (sqrt(s2)^3)
}


# Saves the counts vector from above as a dataframe, with extra sim-info 
# Example with counts = [1] 32 21 15 28:
#
# > count_hist_df(test, "lgcp", 1, 2)
# # A tibble: 4 × 6
#   model   rep    nx    ny     k  freq
#   <chr> <dbl> <dbl> <dbl> <int> <dbl>
# 1 lgcp      1     2     2    15     1
# 2 lgcp      1     2     2    21     1
# 3 lgcp      1     2     2    28     1
# 4 lgcp      1     2     2    32     1
count_hist_df <- function(counts, model, rep, nx, ny = nx) {
  # Tabulating to store more compactly
  # Helps in the case we are using very fine grid resolution
  tab <- table(counts)
  
  tibble(
    model = model, # "LGCP" or "CSCP"
    rep = rep, # replicate number
    nx = nx, # grid resolution
    ny = ny, # grid resolution
    k = as.integer(names(tab)), # The unique count values
    freq = as.numeric(tab) # How many quadrats had each of the above count values
  )
}

# Same as above, but instead of saving the raw counts, it saves the stats info
count_stat_df <- function(counts, model, rep, nx, ny = nx) {
  counts <- as.numeric(counts)
  J <- length(counts)

  # Empirical count distribution
  tab <- table(counts)
  props <- as.numeric(tab) / J
  k_vals <- as.integer(names(tab))

  # Peak height and peak width
  mode_index <- which.max(props)
  T_peak <- props[mode_index]

  # Interquartile width
  q25 <- as.numeric(quantile(counts, 0.25, type = 1))
  q75 <- as.numeric(quantile(counts, 0.75, type = 1))
  T_iqr_width <- q75 - q25

  # Lower / central / upper quantiles
  T_q05 <- as.numeric(quantile(counts, 0.05, type = 1))
  T_median <- as.numeric(quantile(counts, 0.50, type = 1))
  T_q995 <- as.numeric(quantile(counts, 0.995, type = 1))

  # Baseline dispersion
  mean_count <- mean(counts)
  var_count <- stats::var(counts)
  T_disp <- if (mean_count > 0) var_count / mean_count else NA_real_

  tibble(
    model = model,
    rep = rep,
    nx = nx,
    ny = ny,
    T_q05 = T_q05,
    T_peak = T_peak,
    T_iqr_width = T_iqr_width,
    T_median = T_median,
    T_q995 = T_q995,
    T_disp = T_disp
  )
}

run_quadrat_count_study <- function(
    W, # Observation window
    lambda, 
    phi,
    s_cscp = 0.05, # LGCP scale will be best-matching
    grids = c(4, 8, 16, 32),
    n_sims = 200,
    seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)
  
  # Match LGCP scale to CSCP scale
  s_lgcp <- match_lgcp_scale(phi = phi, s_cscp = s_cscp)
  
  # Objects for storing histograms and moments
  hist_list <- vector("list", length = 2 * n_sims * length(grids))
  moment_list <- vector("list", length = 2 * n_sims * length(grids))
  
  # Index is just for keeping calculated counts and summaries together
  idx <- 1
  
  # Study loop
  for (rep in seq_len(n_sims)) {
    # Simulate one LGCP pattern
    X_lgcp <- sim_lgcp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_lgcp
    )$pp
    
    # Simulate one CSCP pattern
    X_cscp <- sim_cscp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_cscp,
      df = 1
    )$pp
    
    # For each grid size
    for (m in grids) {
      # Calculate the counts per grid-tile
      counts_lgcp <- get_quadrat_counts(X_lgcp, nx = m, ny = m)
      counts_cscp <- get_quadrat_counts(X_cscp, nx = m, ny = m)
      
      # Keep track of the counts data plus summaries for each process
      hist_list[[idx]] <- count_hist_df(counts_lgcp, model = "LGCP", rep = rep, nx = m)
      moment_list[[idx]] <- count_stat_df(counts_lgcp, model = "LGCP", rep = rep, nx = m)
      idx <- idx + 1
      
      hist_list[[idx]] <- count_hist_df(counts_cscp, model = "CSCP", rep = rep, nx = m)
      moment_list[[idx]] <- count_stat_df(counts_cscp, model = "CSCP", rep = rep, nx = m)
      idx <- idx + 1
    }
  }
  
  # Combine all the data
  hist_df <- bind_rows(hist_list)
  moment_df <- bind_rows(moment_list)
  
  # Complete missing count values so summaries are comparable
  # For example if LGCP has counts up to 100 but CSCP only to 50
  # Then the below code sets all of CSCP counts from 51 to 100 = 0
  hist_df <- hist_df %>%
    group_by(nx, ny) %>%
    tidyr::complete(model, rep, k = 0:max(k), fill = list(freq = 0)) %>%
    ungroup()
  
  # Calculate proportion of each count within each replicate/grid
  hist_df <- hist_df %>%
    group_by(model, rep, nx, ny) %>%
    mutate(prop = freq / sum(freq)) %>%
    ungroup()
  
  # Histogram envelopes across simulations
  hist_summary <- hist_df %>%
    group_by(model, nx, ny, k) %>%
    summarise(
      mean_prop = mean(prop),
      med_prop = median(prop),
      lo_prop = quantile(prop, 0.025),
      hi_prop = quantile(prop, 0.975),
      .groups = "drop"
    )
  
  # Statistics summaries across simulations
  stat_summary <- moment_df %>%
    group_by(model, nx, ny) %>%
    summarise(
      mean_T_q05 = mean(T_q05, na.rm = TRUE),
      lo_T_q05 = quantile(T_q05, 0.025, na.rm = TRUE),
      hi_T_q05 = quantile(T_q05, 0.975, na.rm = TRUE),
  
      mean_T_peak = mean(T_peak, na.rm = TRUE),
      lo_T_peak = quantile(T_peak, 0.025, na.rm = TRUE),
      hi_T_peak = quantile(T_peak, 0.975, na.rm = TRUE),
  
      mean_T_iqr_width = mean(T_iqr_width, na.rm = TRUE),
      lo_T_iqr_width = quantile(T_iqr_width, 0.025, na.rm = TRUE),
      hi_T_iqr_width = quantile(T_iqr_width, 0.975, na.rm = TRUE),
  
      mean_T_median = mean(T_median, na.rm = TRUE),
      lo_T_median = quantile(T_median, 0.025, na.rm = TRUE),
      hi_T_median = quantile(T_median, 0.975, na.rm = TRUE),
  
      mean_T_q995 = mean(T_q995, na.rm = TRUE),
      lo_T_q995 = quantile(T_q995, 0.025, na.rm = TRUE),
      hi_T_q995 = quantile(T_q995, 0.975, na.rm = TRUE),
  
      mean_T_disp = mean(T_disp, na.rm = TRUE),
      lo_T_disp = quantile(T_disp, 0.025, na.rm = TRUE),
      hi_T_disp = quantile(T_disp, 0.975, na.rm = TRUE),
  
      .groups = "drop"
    )
  
  list(
    hist_df = hist_df,
    hist_summary = hist_summary,
    moment_df = moment_df,
    stat_summary = stat_summary,
    settings = list(
      lambda = lambda,
      phi = phi,
      s_cscp = s_cscp,
      s_lgcp = s_lgcp,
      grids = grids,
      n_sims = n_sims
    )
  )
}

get_curve_for_pattern <- function(X, grids) {
  purrr::map_dfr(grids, function(m) {
    counts <- get_quadrat_counts(X, nx = m, ny = m)
    tab <- table(counts)

    tibble(
      nx = m,
      ny = m,
      k = as.integer(names(tab)),
      prop = as.numeric(tab) / length(counts)
    )
  })
}

curve_distance <- function(obs_curve, ref_curve, grids) {
  total <- 0

  for (m in grids) {
    obs_m <- obs_curve %>% filter(nx == m)
    ref_m <- ref_curve %>% filter(nx == m)

    all_k <- sort(unique(c(obs_m$k, ref_m$k)))

    obs_p <- rep(0, length(all_k))
    ref_p <- rep(0, length(all_k))

    obs_p[match(obs_m$k, all_k)] <- obs_m$prop
    ref_p[match(ref_m$k, all_k)] <- ref_m$mean_prop

    total <- total + sum(abs(obs_p - ref_p))
  }

  total
}

classify_pattern <- function(X, ref_curves, grids) {
  obs_curve <- get_curve_for_pattern(X, grids)

  d_lgcp <- curve_distance(
    obs_curve,
    ref_curves %>% filter(model == "LGCP"),
    grids
  )

  d_cscp <- curve_distance(
    obs_curve,
    ref_curves %>% filter(model == "CSCP"),
    grids
  )

  if (d_lgcp < d_cscp) "LGCP" else "CSCP"
}

run_grid_power_pilot <- function(
    W,
    lambda,
    phi,
    s_cscp,
    grid_sets,
    n_train = 300,
    n_test = 200,
    seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  all_grids <- sort(unique(unlist(grid_sets)))

  train_study <- run_quadrat_count_study(
    W = W,
    lambda = lambda,
    phi = phi,
    s_cscp = s_cscp,
    grids = all_grids,
    n_sims = n_train,
    seed = seed
  )

  ref_curves <- train_study$hist_summary
  s_lgcp <- train_study$settings$s_lgcp

  results <- list()
  idx <- 1

  for (rep in seq_len(n_test)) {
    X_lgcp <- sim_lgcp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_lgcp
    )$pp

    X_cscp <- sim_cscp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_cscp,
      df = 1
    )$pp

    for (set_name in names(grid_sets)) {
      grids <- grid_sets[[set_name]]

      pred_lgcp <- classify_pattern(X_lgcp, ref_curves, grids)
      pred_cscp <- classify_pattern(X_cscp, ref_curves, grids)

      results[[idx]] <- tibble(
        rep = rep,
        grid_set = set_name,
        true_model = "LGCP",
        pred_model = pred_lgcp
      )
      idx <- idx + 1

      results[[idx]] <- tibble(
        rep = rep,
        grid_set = set_name,
        true_model = "CSCP",
        pred_model = pred_cscp
      )
      idx <- idx + 1
    }
  }

  bind_rows(results) %>%
    group_by(grid_set) %>%
    summarise(
      accuracy = mean(true_model == pred_model),
      lgcp_accuracy = mean(pred_model[true_model == "LGCP"] == "LGCP"),
      cscp_accuracy = mean(pred_model[true_model == "CSCP"] == "CSCP"),
      .groups = "drop"
    ) %>%
    arrange(desc(accuracy))
}

# Convert eta-sets to named grid-sets for a given regime
make_named_grid_sets_from_eta <- function(W, lambda, eta_sets) {
  purrr::imap(eta_sets, function(eta, nm) {
    sort(expected_count_to_grid(W, lambda, eta))
  })
}

# Distance classification pilot for ONE parameter regime
run_eta_grid_pilot_one_regime <- function(
    W,
    lambda,
    phi,
    s_cscp,
    eta_sets,
    n_train = 300,
    n_test = 200,
    seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  grid_sets <- make_named_grid_sets_from_eta(W, lambda, eta_sets)
  all_grids <- sort(unique(unlist(grid_sets)))

  message(
    "Running regime: lambda=", lambda,
    ", phi=", phi,
    ", s_cscp=", s_cscp,
    ", grids={", paste(all_grids, collapse = ", "), "}"
  )

  train_study <- run_quadrat_count_study(
    W = W,
    lambda = lambda,
    phi = phi,
    s_cscp = s_cscp,
    grids = all_grids,
    n_sims = n_train,
    seed = seed
  )

  ref_curves <- train_study$hist_summary
  s_lgcp <- train_study$settings$s_lgcp

  out <- vector("list", length = 2 * n_test * length(grid_sets))
  idx <- 1

  for (rep in seq_len(n_test)) {
    X_lgcp <- sim_lgcp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_lgcp
    )$pp

    X_cscp <- sim_cscp(
      W = W,
      lambda = lambda,
      phi = phi,
      s = s_cscp,
      df = 1
    )$pp

    for (set_name in names(grid_sets)) {
      grids <- grid_sets[[set_name]]

      pred_lgcp <- classify_pattern(X_lgcp, ref_curves, grids)
      pred_cscp <- classify_pattern(X_cscp, ref_curves, grids)

      out[[idx]] <- tibble(
        lambda = lambda,
        phi = phi,
        s_cscp = s_cscp,
        s_lgcp = s_lgcp,
        rep = rep,
        eta_set = set_name,
        eta_values = paste(eta_sets[[set_name]], collapse = ", "),
        grid_values = paste(grids, collapse = ", "),
        true_model = "LGCP",
        pred_model = pred_lgcp
      )
      idx <- idx + 1

      out[[idx]] <- tibble(
        lambda = lambda,
        phi = phi,
        s_cscp = s_cscp,
        s_lgcp = s_lgcp,
        rep = rep,
        eta_set = set_name,
        eta_values = paste(eta_sets[[set_name]], collapse = ", "),
        grid_values = paste(grids, collapse = ", "),
        true_model = "CSCP",
        pred_model = pred_cscp
      )
      idx <- idx + 1
    }
  }

  bind_rows(out)
}
```

```{r include=FALSE, eval=FALSE}
grid_sets <- list(
  "10" = c(10),
  "15" = c(15),
  "20" = c(20),
  "25" = c(25),
  "30" = c(30),
  "40" = c(40),
  "60" = c(60),
  "15_20_25_30" = c(15, 20, 25, 30)
  # "10_15_20_25_30" = c(10, 15, 20, 25, 30),
  # "10_15_20" = c(10, 15, 20),
  # "10_30" = c(10, 30),
  # "10_30_60" = c(10, 30, 60),
  # "10_15_20_60" = c(10, 15, 20, 60),
  # "10_15_20_30_60" = c(10, 15, 20, 30, 60),
  # "10_15_20_30__35_40_60" = c(10, 15, 20, 30, 35, 40, 60)
)

grid_power <- run_grid_power_pilot(
  W = owin(),
  lambda = 2000,
  phi = 2,
  s_cscp = 0.1,
  grid_sets = grid_sets,
  n_train = 300,
  n_test = 200,
  seed = 789
)

grid_power
```

As we saw in the previous section, it is not clear whether there exists a single
grid resolution, or rule for selecting grid resolutions, that performs well
across all parameter regimes.

In this section, we therefore focus on selecting a grid (or set of grids) that
provides robust discrimination between the LGCP and CSCP across a range of
settings.

To do this, we compare candidate grid sets using a simulation-based pilot
study. Rather than constructing a formal hypothesis test at this stage, we use a
simple classifier as a convenient way to quantify how well the two models can be 
separated. In this context, classification accuracy serves as a proxy for 
statistical power: a grid strategy with high classification accuracy is one 
where the models are easily distinguishable.

Since the aim is not to develop an optimal classifier, but to determine whether
the two processes can be separated using observable quadrat-count behaviour, we
use a simple distance between empirical quadrat-count distributions as the 
statistic of classification.

For a given grid resolution $m$, define

$$
D_m(X; M) = \sum_k \left| \hat p_m(k;X)-\bar p_{M,m}(k) \right|
$$

where $\hat p_m(k;X)$ is the empirical quadrat-count distribution for the
observed pattern, and $\bar p_{M,m}(k)$ is the mean quadrat-count distribution
under model $M$, estimated via simulation. $\bar p_{M,m}(k)$ acts as an
approximation of the true theoretical distribution of quadrats counts.

:::callout-note
Why is $\bar p_{M,m}(k)$, the mean of the simulated-pattern quadrat count
distributions, an approximation of the the true distribution?

[Elaborate]
:::

:::callout-note
$\bar p_{M,m}(k)$ is straight nasty notation. Can definitely do better.
:::

For a set of grid resolutions $\mathcal M$, we define the overall distance

$$
D(X; M) = \sum_{m \in \mathcal M} D_m(X; M)
$$

and classify the pattern as the model with the smaller distance.

:::callout-note
In simpler terms, $D(X; M)$ is the sum of the distances of *each* empirical
distribution to the true distribution. That is, for every grid resolution within
the grid-set:

1. Calculate the distance to each of the true distributions.

2. Then add them all up (so we will end up with two sums, representing the distances
from the CSCP and LGCP truth respectively).

3. Classify based on whichever distance is smallest.
:::

This classification rule is used only within this pilot study to compare grid
strategies. In the next section, we use the same distance-based statistic to
construct a more formal Monte Carlo hypothesis test.

## Relative grid sizing

There is one last thing we need to address before conducting the study.

Rather than selecting grid resolutions directly, we index candidate grids by the
expected number of points per quadrat,

$$
\eta_m = \frac{\lambda |W|}{m^2}
$$

This gives a scale-free way to describe grid coarseness. For example,
$\eta_m = 20$ corresponds to a grid where each quadrat contains 20 points on
average, regardless of the size of the observation window.

For a target expected quadrat count $\eta$, the corresponding square grid
resolution is

$$
m(\eta) =
\left\lceil
\sqrt{\frac{\lambda |W|}{\eta}}
\right\rceil
$$

:::callout-note
Should this be floor instead of ceiling?
:::

Thus, candidate grid strategies are defined in terms of sets of expected counts,
such as

$$
\mathcal E = \{5, 10, 20, 40\},
$$

rather than fixed values of $m$. This makes the resulting rule easier to apply
outside the simulation setting, where the observation window and intensity may
differ.

```{r echo = FALSE}
expected_count_to_grid <- function(W, lambda, eta) {
  area_W <- spatstat.geom::area.owin(W)
  m <- round(sqrt(lambda * area_W / eta))
  pmax(1, unique(m))
}

make_grid_sets_from_eta <- function(W, lambda, eta_sets) {
  purrr::map(eta_sets, ~ expected_count_to_grid(W, lambda, .x))
}
```

Below we show an example, where the candidate grid-sets have expected counts
$\{5 \}$ and $\{5, 20\}$, and $\lambda = 2000$.

```{r}
eta_sets <- list(
  "eta_5" = 5,
  "eta_5_20" = c(5, 20)
)

grid_sets <- make_grid_sets_from_eta(
  W = owin(),
  lambda = 2000,
  eta_sets = eta_sets
)
grid_sets
```

We can see, we can see the associated grid resolutions for the above expected
quadrat counts.

## Pilot design

For each parameter regime, we generate independent training and test sets from
both the CSCP and matched LGCP. The training set is used only to estimate the
reference quadrat-count distributions $\bar p_{M,m}$. The test set is then
classified using the distance rule above.

We repeat this over a grid of parameter regimes,

$$
\lambda \in \{\cdots\}, \qquad
\phi \in \{\cdots\}, \qquad
s_{\text{CSCP}} \in \{\cdots\}.
$$

For each candidate grid strategy $\mathcal E$, we record the classification
accuracy within each regime, together with its mean, minimum, and interquartile
width across regimes.

Since the goal is robustness rather than best-case performance, candidate
strategies are ranked primarily by their worst-case accuracy across regimes.
Ties are broken using mean accuracy and then interquartile width.

## Pilot

Candidate grid sets:

```{r}
eta_sets <- list(
  "eta_2" = 2,
  "eta_5" = 5,
  "eta_10" = 10,
  "eta_20" = 20,
  "eta_40" = 40,
  "eta_80" = 80,

  "eta_2_5_10" = c(2, 5, 10),
  "eta_5_10_20" = c(5, 10, 20),
  "eta_10_20_40" = c(10, 20, 40),
  "eta_20_40_80" = c(20, 40, 80),

  "eta_2_5_10_20" = c(2, 5, 10, 20),
  "eta_5_10_20_40" = c(5, 10, 20, 40),
  "eta_10_20_40_80" = c(10, 20, 40, 80),

  "eta_2_5_10_20_40" = c(2, 5, 10, 20, 40),
  "eta_5_10_20_40_80" = c(5, 10, 20, 40, 80),
  "eta_2_5_10_20_40_80" = c(2, 5, 10, 20, 40, 80)
)
```

And regimes:

```{r}
regimes <- tidyr::crossing(
  lambda = c(1000, 5000, 10000),
  phi = c(1, 1.5, 2),
  s_cscp = c(0.1)
)
```

```{r include=FALSE}
run_eta_grid_pilot <- function(
    W,
    regimes,
    eta_sets,
    n_train = 300,
    n_test = 200,
    seed = 1
) {
  purrr::pmap_dfr(
    regimes,
    function(lambda, phi, s_cscp) {
      regime_seed <- seed +
        100000 * match(lambda, unique(regimes$lambda)) +
        1000 * match(phi, unique(regimes$phi)) +
        10 * match(s_cscp, unique(regimes$s_cscp))

      run_eta_grid_pilot_one_regime(
        W = W,
        lambda = lambda,
        phi = phi,
        s_cscp = s_cscp,
        eta_sets = eta_sets,
        n_train = n_train,
        n_test = n_test,
        seed = regime_seed
      )
    }
  )
}
```

Running the pilot:

```{r message=FALSE, results=FALSE, cache=TRUE}
eta_grid_results <- run_eta_grid_pilot(
  W = spatstat.geom::owin(c(0, 1), c(0, 1)),
  regimes = regimes,
  eta_sets = eta_sets,
  n_train = 300,
  n_test = 100,
  seed = 789
)
```

Summarise by regime:

```{r include=FALSE}
eta_grid_by_regime <- eta_grid_results %>%
  group_by(lambda, phi, s_cscp, eta_set, eta_values, grid_values) %>%
  summarise(
    accuracy = mean(true_model == pred_model),
    lgcp_accuracy = mean(pred_model[true_model == "LGCP"] == "LGCP"),
    cscp_accuracy = mean(pred_model[true_model == "CSCP"] == "CSCP"),
    .groups = "drop"
  )
```

```{r echo=FALSE}
eta_grid_ranking <- eta_grid_by_regime %>%
  group_by(eta_set, eta_values) %>%
  summarise(
    mean_accuracy = mean(accuracy),
    min_accuracy = min(accuracy),
    q25_accuracy = quantile(accuracy, 0.25),
    median_accuracy = median(accuracy),
    q75_accuracy = quantile(accuracy, 0.75),
    iqr_accuracy = IQR(accuracy),
    sd_accuracy = sd(accuracy),
    mean_lgcp_accuracy = mean(lgcp_accuracy),
    mean_cscp_accuracy = mean(cscp_accuracy),
    .groups = "drop"
  ) %>%
  arrange(
    desc(min_accuracy),
    desc(mean_accuracy),
    iqr_accuracy
  )

grid_lookup <- eta_grid_by_regime %>%
  distinct(eta_set, eta_values, lambda, grid_values) %>%
  arrange(eta_set, lambda) %>%
  group_by(eta_set, eta_values) %>%
  summarise(
    induced_grids = paste0(
      "$\\lambda=", lambda, "$: {", grid_values, "}",
      collapse = ";"
    ),
    .groups = "drop"
  )

eta_grid_table <- eta_grid_ranking %>%
  left_join(grid_lookup, by = c("eta_set", "eta_values")) %>%
  slice_head(n = 8) %>%
  transmute(
    `Expected counts per quadrat` = paste0("{", eta_values, "}"),
    # `Induced grid resolutions` = induced_grids,
    `Worst-case accuracy` = round(min_accuracy, 3),
    `Mean accuracy` = round(mean_accuracy, 3),
    `Median accuracy` = round(median_accuracy, 3),
    `IQR` = round(iqr_accuracy, 3),
    `Mean LGCP accuracy` = round(mean_lgcp_accuracy, 3),
    `Mean CSCP accuracy` = round(mean_cscp_accuracy, 3)
  )

knitr::kable(
  eta_grid_table,
  escape = FALSE,
  align = "llrrrrrr",
  caption = "Top candidate grid strategies ranked by worst-case classification accuracy across all parameter regimes."
)
```

Best within each regime:

```{r echo=FALSE}
eta_grid_best_by_regime <- eta_grid_by_regime %>%
  group_by(lambda, phi, s_cscp) %>%
  slice_max(
    order_by = accuracy,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  transmute(
    `lambda` = lambda,
    `phi` = phi,
    `s_CSCP` = s_cscp,
    `Best expected counts per quadrat` = paste0("{", eta_values, "}"),
    `Induced grid resolutions` = paste0("{", grid_values, "}"),
    `Accuracy` = round(accuracy, 3),
    `LGCP accuracy` = round(lgcp_accuracy, 3),
    `CSCP accuracy` = round(cscp_accuracy, 3)
  )

knitr::kable(
  eta_grid_best_by_regime,
  escape = FALSE,
  align = "rrrllrrr",
  caption = "Best-performing candidate grid strategy within each parameter regime."
)
```

Interesting that when $\phi = 2$ and $\lambda = 5000$, the best grid set is just
$\{2\}$. 

Nice plot of results:

```{r echo=FALSE}
top_eta_sets <- eta_grid_ranking$eta_set[1:5]

plot_df <- eta_grid_by_regime %>%
  dplyr::filter(eta_set %in% top_eta_sets) %>%
  dplyr::mutate(
    regime = paste0(
      "lambda=", lambda,
      ", phi=", phi,
      ", s=", s_cscp
    )
  )

regime_levels <- plot_df %>%
  dplyr::distinct(lambda, phi, s_cscp, regime) %>%
  dplyr::arrange(lambda, phi, s_cscp) %>%
  dplyr::pull(regime)

plot_df <- plot_df %>%
  dplyr::mutate(
    regime = factor(regime, levels = regime_levels)
  )

cols <- c("steelblue", "tomato", "darkgreen", "purple", "orange")
names(cols) <- top_eta_sets

plot(
  NA,
  xlim = c(1, length(regime_levels)),
  ylim = range(plot_df$accuracy, 0.5, 1, na.rm = TRUE),
  xaxt = "n",
  xlab = "",
  ylab = "Classification accuracy",
  main = "Accuracy across parameter regimes"
)

abline(h = seq(0.5, 1, by = 0.1), col = "grey90", lty = 1)
abline(h = 0.5, col = "grey50", lty = 2)

axis(
  1,
  at = seq_along(regime_levels),
  labels = seq_along(regime_levels)
)

for (i in seq_along(top_eta_sets)) {
  this_set <- top_eta_sets[i]

  d <- plot_df %>%
    dplyr::filter(eta_set == this_set) %>%
    dplyr::arrange(regime)

  lines(
    seq_along(regime_levels),
    d$accuracy,
    type = "b",
    pch = 16,
    lwd = 2,
    col = cols[this_set]
  )
}

legend_labels <- eta_grid_ranking %>%
  dplyr::filter(eta_set %in% top_eta_sets) %>%
  dplyr::arrange(match(eta_set, top_eta_sets)) %>%
  dplyr::mutate(
    label = paste0("eta = {", eta_values, "}")
  ) %>%
  dplyr::pull(label)

legend(
  "bottomright",
  legend = legend_labels,
  col = cols[top_eta_sets],
  lwd = 2,
  pch = 16,
  bty = "n",
  cex = 0.8
)

mtext("Parameter regime index", side = 1, line = 2.2)
```

```{r echo=FALSE}
regime_key <- plot_df %>%
  distinct(regime, lambda, phi, s_cscp) %>%
  arrange(lambda, phi, s_cscp) %>%
  mutate(`Regime index` = row_number()) %>%
  select(
    `Regime index`,
    lambda,
    phi,
    `s_CSCP` = s_cscp
  )

knitr::kable(
  regime_key,
  align = "rrrr",
  caption = "Parameter regimes used in the pilot power study."
)
```

