# 🚧🚧 Marginal Distribution of $\Lambda(u)$

Tight tight tight

In the previous sections, we showed that LGCP and CSCP models can produce nearly
identical second-order structure, leading to similar inference based on the pair
correlation function.

We now investigate whether the models can be distinguished through the marginal
distribution of the underlying intensity field $\Lambda(u)$.

## Visual inspection

So we are looking at the marginal distribution of the LGCP and CSCP. 

We will start by just doing some visual inspection of the intensity surfaces.

```{r}
# 2x2 panel: intensity surfaces
# rows = different realisations
# cols = CSCP (left), LGCP (right)
```

```{r}
# 2x2 panel plotting the 3D!!! intensity surface for LGCP CSCP with the same lambda, phi
# CSCP on the left, LGCP on the right
```

Comments on the plot.

## Observed data

Now if we were able to have access to the true intensity surface, obviously we
could distinguish between the two processes, but unfortunately we only observe the
point patterns. Taking a look at some example patterns:

```{r}
# 5x2 panel plotting the 2D point patterns for LGCP CSCP with the same lambda, phi
# CSCP on the left, LGCP on the right
# Increasing lambda as you go down row-wise
```

Probably much harder to distinguish from the patterns alone.

So we would ideally like to have a procedure which does not rely on our visual 
interpretation, which brings us to quadrat counts.

## Quadrat count idea

To extract information about the marginal distribution of $\Lambda(u)$, we consider
quadrat counts.

The idea is that, over sufficiently small regions, the number of points reflects
the local intensity. Differences in the distribution of $\Lambda(u)$ may therefore
be detectable through the distribution of quadrat counts.

```{r}
# Example: same pattern, quadrat overlay
# left = CSCP, right = LGCP
```

```{r}
# Histogram / distribution of quadrat counts
# compare CSCP vs LGCP
```

We expect the LGCP to produce more extreme counts, corresponding to its
heavier-tailed intensity distribution, while the CSCP produces more moderate
variability.

This suggests that quadrat-based summaries may provide a practical way to
distinguish between the models, even when their second-order structure is similar.





