<!-- Complete dogshit here -->
<div id="pp-wrap" style="width:100%; display:flex; justify-content:center; margin:0; padding:0; margin-bottom: -16px;">
  <canvas id="pp-canvas" style="display:block; width:100%; max-width:980px; height:auto; cursor:pointer;"></canvas>
</div>

<script>
(() => {
  // ============================================================
  // Tunable settings
  // ============================================================
  const WIDTH = 980;
  const HEIGHT = 350;

  // Mask generation
  const MASK_RES = 1;
  const NOISE_SCALE_1 = 0.010;
  const NOISE_SCALE_2 = 0.022;
  const NOISE_SCALE_3 = 0.055;
  const THRESHOLD = 0.485;
  const WARP = 45;
  const BLUR_PASSES = 2;

  // Point process
  const LAMBDA = 115;
  const MEAN_LIFETIME = 8.5;
  const MAX_POINTS = 500;
  const POINT_RADIUS = 5.6;
  const STROKE_WIDTH = 1.7;
  const POINT_FILL = "#f2f2f2";
  const POINT_STROKE = "#2b2b2b";

  // Appearance
  const BORDER_COLOR = "#666";
  const BORDER_WIDTH = 1.5;
  const BLACK = "#000000";
  const WHITE = "#ffffff";

  // ============================================================
  // Setup
  // ============================================================
  const canvas = document.getElementById("pp-canvas");
  const ctx = canvas.getContext("2d");
  canvas.width = WIDTH;
  canvas.height = HEIGHT;

  const off = document.createElement("canvas");
  const offCtx = off.getContext("2d");
  off.width = WIDTH;
  off.height = HEIGHT;

  const maskCanvas = document.createElement("canvas");
  const maskCtx = maskCanvas.getContext("2d");

  const maskW = Math.floor(WIDTH / MASK_RES);
  const maskH = Math.floor(HEIGHT / MASK_RES);
  maskCanvas.width = maskW;
  maskCanvas.height = maskH;

  let spawnMask = new Uint8Array(maskW * maskH);
  let whitePixels = [];
  let valueGrid = new Map();
  let rng = null;

  // ============================================================
  // RNG
  // ============================================================
  function mulberry32(a) {
    return function() {
      let t = a += 0x6D2B79F5;
      t = Math.imul(t ^ (t >>> 15), t | 1);
      t ^= t + Math.imul(t ^ (t >>> 7), t | 61);
      return ((t ^ (t >>> 14)) >>> 0) / 4294967296;
    };
  }

  function reseed(seed = (Date.now() ^ Math.floor(Math.random() * 1e9)) >>> 0) {
    rng = mulberry32(seed);
    valueGrid = new Map();
  }

  // ============================================================
  // Noise helpers
  // ============================================================
  function hash2(ix, iy) {
    const key = `${ix},${iy}`;
    if (!valueGrid.has(key)) valueGrid.set(key, rng());
    return valueGrid.get(key);
  }

  function smoothstep(t) {
    return t * t * (3 - 2 * t);
  }

  function lerp(a, b, t) {
    return a + (b - a) * t;
  }

  function valueNoise(x, y, scale) {
    const sx = x * scale;
    const sy = y * scale;

    const x0 = Math.floor(sx);
    const y0 = Math.floor(sy);
    const x1 = x0 + 1;
    const y1 = y0 + 1;

    const tx = smoothstep(sx - x0);
    const ty = smoothstep(sy - y0);

    const v00 = hash2(x0, y0);
    const v10 = hash2(x1, y0);
    const v01 = hash2(x0, y1);
    const v11 = hash2(x1, y1);

    const a = lerp(v00, v10, tx);
    const b = lerp(v01, v11, tx);
    return lerp(a, b, ty);
  }

  function fbm(x, y) {
    return (
      0.56 * valueNoise(x, y, NOISE_SCALE_1) +
      0.29 * valueNoise(x + 173.1, y - 91.7, NOISE_SCALE_2) +
      0.15 * valueNoise(x - 71.3, y + 44.2, NOISE_SCALE_3)
    );
  }

  // ============================================================
  // Mask generation
  // ============================================================
  function generateMask() {
    const img = maskCtx.createImageData(maskW, maskH);
    const data = img.data;
    const raw = new Float32Array(maskW * maskH);

    for (let j = 0; j < maskH; j++) {
      for (let i = 0; i < maskW; i++) {
        const idx = j * maskW + i;
        const x = i * MASK_RES;
        const y = j * MASK_RES;

        const wx = x + WARP * (fbm(x + 400, y + 100) - 0.5);
        const wy = y + WARP * (fbm(x - 200, y + 700) - 0.5);

        let z = fbm(wx, wy);

        const ridges =
          Math.abs(fbm(wx + 80, wy - 120) - 0.5) +
          0.8 * Math.abs(fbm(wx - 250, wy + 40) - 0.5);

        z = 0.72 * z + 0.28 * ridges;
        raw[idx] = z;
      }
    }

    let cur = raw;
    for (let pass = 0; pass < BLUR_PASSES; pass++) {
      const nxt = new Float32Array(maskW * maskH);
      for (let j = 0; j < maskH; j++) {
        for (let i = 0; i < maskW; i++) {
          let sum = 0;
          let cnt = 0;
          for (let dy = -1; dy <= 1; dy++) {
            for (let dx = -1; dx <= 1; dx++) {
              const xx = i + dx;
              const yy = j + dy;
              if (xx >= 0 && xx < maskW && yy >= 0 && yy < maskH) {
                sum += cur[yy * maskW + xx];
                cnt++;
              }
            }
          }
          nxt[j * maskW + i] = sum / cnt;
        }
      }
      cur = nxt;
    }

    whitePixels = [];

    for (let j = 0; j < maskH; j++) {
      for (let i = 0; i < maskW; i++) {
        const idx = j * maskW + i;
        const v = cur[idx];

        const isWhite = v < THRESHOLD ? 1 : 0;
        spawnMask[idx] = isWhite;

        if (isWhite) whitePixels.push(idx);

        const c = isWhite ? 255 : 0;
        const k = idx * 4;
        data[k] = c;
        data[k + 1] = c;
        data[k + 2] = c;
        data[k + 3] = 255;
      }
    }

    maskCtx.putImageData(img, 0, 0);
  }

  function drawBackground() {
    offCtx.clearRect(0, 0, WIDTH, HEIGHT);
    offCtx.fillStyle = WHITE;
    offCtx.fillRect(0, 0, WIDTH, HEIGHT);
    offCtx.imageSmoothingEnabled = true;
    offCtx.drawImage(maskCanvas, 0, 0, WIDTH, HEIGHT);
    offCtx.lineWidth = BORDER_WIDTH;
    offCtx.strokeStyle = BORDER_COLOR;
    offCtx.strokeRect(0.5, 0.5, WIDTH - 1, HEIGHT - 1);
  }

  // ============================================================
  // Point process
  // ============================================================
  function sampleSpawnPoint() {
    if (whitePixels.length === 0) return null;

    const idx = whitePixels[Math.floor(rng() * whitePixels.length)];
    const j = Math.floor(idx / maskW);
    const i = idx - j * maskW;

    return {
      x: (i + rng()) * MASK_RES,
      y: (j + rng()) * MASK_RES
    };
  }

  function expSample(mean) {
    return -mean * Math.log(1 - rng());
  }

  const points = [];
  let lastTime = performance.now();

  function refillPoints() {
    points.length = 0;
    for (let i = 0; i < Math.round(LAMBDA * 0.9); i++) {
      const pt = sampleSpawnPoint();
      if (pt) {
        const lifetime = expSample(MEAN_LIFETIME);
        points.push({
          x: pt.x,
          y: pt.y,
          age: rng() * lifetime,
          lifetime
        });
      }
    }
  }

  function maybeSpawn(dt) {
    const birthRate = LAMBDA / MEAN_LIFETIME;
    let expectedBirths = birthRate * dt;

    while (expectedBirths > 0) {
      const p = Math.min(1, expectedBirths);
      if (rng() < p) {
        const pt = sampleSpawnPoint();
        if (pt && points.length < MAX_POINTS) {
          points.push({
            x: pt.x,
            y: pt.y,
            age: 0,
            lifetime: expSample(MEAN_LIFETIME)
          });
        }
      }
      expectedBirths -= 1;
    }
  }

  function update(dt) {
    maybeSpawn(dt);

    for (let i = points.length - 1; i >= 0; i--) {
      points[i].age += dt;
      if (points[i].age >= points[i].lifetime) {
        points.splice(i, 1);
      }
    }
  }

  function drawPoints() {
    for (const p of points) {
      const t = p.age / p.lifetime;
      let alpha = 1;
      if (t < 0.08) alpha = t / 0.08;
      else if (t > 0.88) alpha = (1 - t) / 0.12;

      ctx.globalAlpha = Math.max(0, Math.min(1, alpha));

      ctx.beginPath();
      ctx.arc(p.x, p.y, POINT_RADIUS, 0, Math.PI * 2);
      ctx.fillStyle = POINT_FILL;
      ctx.fill();
      ctx.lineWidth = STROKE_WIDTH;
      ctx.strokeStyle = POINT_STROKE;
      ctx.stroke();
    }
    ctx.globalAlpha = 1;
  }

  function frame(now) {
    const dt = Math.min(0.05, (now - lastTime) / 1000);
    lastTime = now;

    update(dt);

    ctx.clearRect(0, 0, WIDTH, HEIGHT);
    ctx.drawImage(off, 0, 0);
    drawPoints();

    requestAnimationFrame(frame);
  }

  // ============================================================
  // Regenerate on click
  // ============================================================
  function regenerate() {
    reseed();
    generateMask();
    drawBackground();
    refillPoints();
  }

  canvas.addEventListener("click", regenerate);

  // ============================================================
  // Initialise
  // ============================================================
  regenerate();
  requestAnimationFrame(frame);
})();
</script>

# Preface {.unnumbered}

This webpage serves as both documentation for the `cscp` package, as well as 
an up-to-date reference for my thesis progress.

Skip to:

<div class="cta-container">

<a href="thesis/cscp_preface.html" class="cta-button secondary">
📘 CSCP I (Thesis Material)
</a>

<a href="thesis/lgcp_motivation.html" class="cta-button secondary">
📘 CSCP II (Thesis Material)
</a>

<a href="package_docs/basics.html" class="cta-button secondary">
📦 Package Documentation
</a>

</div>
## Installation

You can install the `cscp` package using `devtools::install_github()`:

```{r eval = FALSE}
devtools::install_github("https://github.com/JosephReps/coxLab.git")
```

## Usage

Under construction...

---

```{r}
#| label: game-data
#| echo: false
#| message: false
#| warning: false

library(jsonlite)
source("../R/sim.R")
source("../R/util.R")

make_round <- function(
  family,
  W = spatstat.geom::owin(c(0, 1), c(0, 1)),
  lambda = 100,
  phi = 1.5,
  s = 0.05,
  df = 1,
  npix = 64
) {
  sim <- if (family == "LGCP") {
    sim_lgcp(W = W, lambda = lambda, phi = phi, s = s)
  } else {
    sim_cscp(W = W, lambda = lambda, phi = phi, s = s, df = df)
  }

  # Extract the simulated intensity image directly onto a matrix
  z <- sim$intensity$v
  
  # Optional: keep output size manageable / consistent if needed
  # If your simulated image is not already npix x npix, then resample separately.

  pts <- data.frame(x = sim$pp$x, y = sim$pp$y)

  list(
    family = family,
    lambda = lambda,
    phi = phi,
    s = s,
    z = unname(split(z, row(z))),
    points = unname(lapply(seq_len(nrow(pts)), function(i) {
      list(x = pts$x[i], y = pts$y[i])
    }))
  )
}

rounds <- c(
  lapply(seq_len(40), function(i) make_round("LGCP")),
  lapply(seq_len(40), function(i) make_round("CSCP"))
)

rounds_json <- jsonlite::toJSON(rounds, auto_unbox = TRUE)
```

```{r}
#| echo: false
#| results: asis

cat('<script id="cscp-game-data" type="application/json">')
cat(rounds_json)
cat('</script>')
```


::: {.guess-game-wrap}
::: {.guess-game-top}
<div class="guess-game-score">Score: <span id="score-correct">0</span>/<span id="score-attempts">0</span></div>
<button id="score-reset" class="guess-game-reset" type="button">Reset</button>
:::

::: {.guess-game-canvas-wrap}
<canvas id="guess-game-canvas" width="520" height="520"></canvas>
<div id="guess-game-feedback" class="guess-game-feedback"></div>
:::

::: {.guess-game-buttons}
<button id="guess-lgcp" class="guess-btn" type="button">LGCP</button>
<button id="guess-cscp" class="guess-btn" type="button">CSCP</button>
:::
:::

<script>
function spatstatTerrain(t) {
  const stops = [
    [0, 0, 130],      // dark blue
    [0, 120, 255],    // blue
    [0, 180, 120],    // green
    [200, 220, 0],    // yellow-green
    [255, 160, 0],    // orange
    [200, 0, 0]       // red
  ];

  const n = stops.length - 1;
  const i = Math.min(Math.floor(t * n), n - 1);
  const f = t * n - i;

  const c0 = stops[i];
  const c1 = stops[i + 1];

  const r = Math.round(c0[0] + f * (c1[0] - c0[0]));
  const g = Math.round(c0[1] + f * (c1[1] - c0[1]));
  const b = Math.round(c0[2] + f * (c1[2] - c0[2]));

  return `rgb(${r}, ${g}, ${b})`;
}

(function initGuessGame() {
  const dataNode   = document.getElementById("cscp-game-data");
  const canvas     = document.getElementById("guess-game-canvas");
  const feedback   = document.getElementById("guess-game-feedback");
  const correctEl  = document.getElementById("score-correct");
  const attemptsEl = document.getElementById("score-attempts");
  const btnLGCP    = document.getElementById("guess-lgcp");
  const btnCSCP    = document.getElementById("guess-cscp");
  const btnReset   = document.getElementById("score-reset");

  if (!dataNode || !canvas || !feedback || !correctEl || !attemptsEl ||
      !btnLGCP || !btnCSCP || !btnReset) {
    console.error("Game widget elements not found.");
    return;
  }

  const rounds = JSON.parse(dataNode.textContent);
  const ctx = canvas.getContext("2d");

  let scoreCorrect = 0;
  let scoreAttempts = 0;
  let currentRound = null;
  let locked = false;

  function updateScore() {
    correctEl.textContent = scoreCorrect;
    attemptsEl.textContent = scoreAttempts;
  }

  function drawSurface(z) {
    const nRow = z.length;
    const nCol = z[0].length;

    let min = Infinity;
    let max = -Infinity;

    for (let r = 0; r < nRow; r++) {
      for (let c = 0; c < nCol; c++) {
        const v = z[r][c];
        if (v < min) min = v;
        if (v > max) max = v;
      }
    }

    const cellW = canvas.width / nCol;
    const cellH = canvas.height / nRow;

    ctx.clearRect(0, 0, canvas.width, canvas.height);

    for (let r = 0; r < nRow; r++) {
      for (let c = 0; c < nCol; c++) {
        const v = z[r][c];
        const t = (v - min) / ((max - min) || 1);
        
        ctx.fillStyle = spatstatTerrain(t);
        ctx.fillRect(c * cellW, r * cellH, cellW + 0.5, cellH + 0.5);
      }
    }
  }

  function sampleRound() {
    currentRound = rounds[Math.floor(Math.random() * rounds.length)];
    drawSurface(currentRound.z);
  }

  function showFeedback(isCorrect) {
    feedback.textContent = isCorrect ? "✓" : "✕";
    feedback.className = "guess-game-feedback show " + (isCorrect ? "correct" : "wrong");

    setTimeout(() => {
      feedback.className = "guess-game-feedback";
    }, 800);
  }

  function handleGuess(guess) {
    if (locked || !currentRound) return;

    locked = true;

    const isCorrect = guess === currentRound.family;
    scoreAttempts += 1;
    if (isCorrect) scoreCorrect += 1;

    updateScore();
    showFeedback(isCorrect);

    setTimeout(() => {
      sampleRound();
      locked = false;
    }, 900);
  }

  function resetGame() {
    scoreCorrect = 0;
    scoreAttempts = 0;
    updateScore();
    sampleRound();
  }

  btnLGCP.addEventListener("click", () => handleGuess("LGCP"));
  btnCSCP.addEventListener("click", () => handleGuess("CSCP"));
  btnReset.addEventListener("click", resetGame);

  updateScore();
  sampleRound();
})();
</script>

























