```{r include = FALSE}
devtools::load_all()
library(spatstat)
```
# Minimum contrast

I've implemented my own minimum contrast. You can find the code 
[here](https://github.com/JosephReps/coxLab)

## Examples

Fitting a CSCP minimum contrast model:

```{r}
#| code-fold: true
set.seed(1)
# Simulating a point pattern realization from a CSCP on window [0,1]x[0,1] 
cscp_pp <- sim_cscp(lambda = 100, phi = 1.5, s = 0.05)$pp

# Fitting a MC model
cscp_fit <- fit_cscp_mc(cscp_pp, bw = 0.02)

# Comparing the model fit to the true PCF
r_vals <- seq(0, 2.5, by = 0.001)

fitted_pcf_vals <- pcf_cscp_theoretical(r_vals, lambda = cscp_fit$par[["lambda"]], 
                                        phi = cscp_fit$par[["phi"]], 
                                        s = cscp_fit$par[["s"]])
theoretical_pcf_vals <- pcf_cscp_theoretical(r_vals, lambda = 100, phi = 1.5, s = 0.05)

plot(pcf(cscp_pp, correction = "isotropic", divisor = "a",
                        zerocor = "JonesFoster", bw = 0.02), 
     ylim = c(0.8, 3), legend = F, main = "Example of MC CSCP fit (lambda = 100, phi = 1.5, s = 0.05)")
lines(r_vals, theoretical_pcf_vals, col = "forestgreen", lwd = 2)
lines(r_vals, fitted_pcf_vals, col = "red", lwd = 2)
legend("topright", legend = c("Empirical", "True", "Fitted"), col = c("black", "forestgreen", "red"), lwd = c(1, 2, 2))
```

Fitting an LGCP minimum contrast model:

```{r}
#| code-fold: true
set.seed(1)
# Simulating a point pattern realization from a CSCP on window [0,1]x[0,1] 
lgcp_pp <- sim_lgcp(lambda = 100, phi = 5, s = 0.05)$pp

# Fitting a MC model
# Notice we have used a smaller bandwidth for this fit
lgcp_fit <- fit_lgcp_mc(lgcp_pp, bw = 0.01)

# Comparing the model fit to the true PCF
r_vals <- seq(0, 1, by = 0.001)

fitted_pcf_vals <- pcf_lgcp_theoretical(r_vals, lambda = lgcp_fit$par[["lambda"]], 
                                        phi = lgcp_fit$par[["phi"]], 
                                        s = lgcp_fit$par[["s"]])
theoretical_pcf_vals <- pcf_lgcp_theoretical(r_vals, lambda = 100, phi = 5, s = 0.05)

plot(pcf(cscp_pp, correction = "isotropic", divisor = "a",
                        zerocor = "JonesFoster", bw = 0.01), 
     ylim = c(0.8, 6.5), legend = F, main = "Example of MC LGCP fit (lambda = 100, phi = 5, s = 0.05)")
lines(r_vals, theoretical_pcf_vals, col = "forestgreen", lwd = 2)
lines(r_vals, fitted_pcf_vals, col = "red", lwd = 2)
legend("topright", legend = c("Empirical", "True", "Fitted"), col = c("black", "forestgreen", "red"), lwd = c(1, 2, 2))
```

## Implementation notes

Firstly, `spatstat` already provides a very useful `mincontrast()` function, and
it is probably faster and more reliable than my implementation, but if I didn't 
do this I would never bother to understand what is happening under the hood. I would
eventually like to implement the non-parameteric PCF estimators as well.

I will comment on the CSCP implementation below, the LGCP is almost the same. I've
left the in-line documentation out here, see the source file linked above if you
prefer reading that, or use `?fit_cscp_mc` after loading the library.

```{r eval = FALSE}
#| code-fold: true
estimate_pcf_for_mc <- function(X,
                                bw = 0.2,
                                correction = "isotropic",
                                divisor = "a",
                                zerocor = "JonesFoster",
                                r_min = 0.02,
                                r_max = NULL) {
  stopifnot(inherits(X, "ppp"))

  obs <- spatstat.explore::pcf(
    X,
    bw = bw,
    correction = correction,
    divisor = divisor,
    zerocor = zerocor
  )

  y_name <- switch(
    correction,
    isotropic = "iso",
    translate = "trans",
    Ripley = "iso",
    "iso"
  )

  if (!y_name %in% names(obs)) {
    y_name <- intersect(c("iso", "trans", "un"), names(obs))[1]
  }

  r <- obs$r
  g <- obs[[y_name]]

  keep <- is.finite(r) & is.finite(g) & r > r_min
  if (!is.null(r_max)) keep <- keep & r <= r_max

  list(
    fv = obs,
    r = r[keep],
    g = g[keep],
    correction_used = y_name,
    bw = bw,
    divisor = divisor,
    zerocor = zerocor,
    r_min = r_min,
    r_max = r_max
  )
}

mc_pcf_objective <- function(par,
                             r,
                             g_obs,
                             model = c("lgcp", "cscp"),
                             lambda,
                             df = 1,
                             k = 0.25,
                             p = 2) {
  model <- match.arg(model)

  if (model == "lgcp") {
    phi <- exp(par[1]) # phi in (0, Inf)              
    s   <- exp(par[2]) # s > 0              
    g_th <- pcf_lgcp_theoretical(
      r,
      lambda = lambda,
      phi = phi,
      s = s
    )
  } else {
    # Note for CSCP we need to transform the parameters in order to
    # use the pcf_cscp_theoretical() function.
    eta <- par[1] # eta in R                   
    phi <- 2 * stats::plogis(eta) # phi in (0, 2)   
    s   <- exp(par[2]) # s > 0              
    g_th <- pcf_cscp_theoretical(
      r,
      lambda = lambda,
      phi = phi,
      s = s,
      df = df
    )
  }

  sum(abs(g_obs^k - g_th^k)^p)
}

fit_cscp_mc <- function(X,
                        start = list(phi = 0.5, s = 0.05),
                        df = 1,
                        bw = 0.2,
                        correction = "isotropic",
                        divisor = "a",
                        zerocor = "JonesFoster",
                        r_min = 0.02,
                        r_max = NULL,
                        k = 0.25,
                        p = 2) {
  stopifnot(inherits(X, "ppp"))
  stopifnot(is.list(start), !is.null(start$phi), !is.null(start$s))
  stopifnot(start$phi > 0, start$phi < 2, start$s > 0)

  lambda_hat <- spatstat.geom::intensity(X)

  pcf_obs <- estimate_pcf_for_mc(
    X,
    bw = bw,
    correction = correction,
    divisor = divisor,
    zerocor = zerocor,
    r_min = r_min,
    r_max = r_max
  )

  # Inverse of phi = 2*plogis(eta)
  eta_start <- stats::qlogis(start$phi / 2)

  opt <- stats::optim(
    par = c(eta_start, log(start$s)),
    fn = mc_pcf_objective,
    r = pcf_obs$r,
    g_obs = pcf_obs$g,
    model = "cscp",
    lambda = lambda_hat,
    df = df,
    k = k,
    p = p,
    method = "L-BFGS-B",
    lower = c(-20, log(1e-8)),
    upper = c(20, log(10))
  )

  pars <- c(
    lambda = lambda_hat,
    phi = 2 * stats::plogis(opt$par[1]),
    s = exp(opt$par[2])
  )

  g_fit <- pcf_cscp_theoretical(
    pcf_obs$r,
    lambda = pars["lambda"],
    phi = pars["phi"],
    s = pars["s"],
    df = df
  )

  structure(
    list(
      model = "CSCP",
      df = df,
      par = pars,
      objective = opt$value,
      convergence = opt$convergence,
      optim = opt,
      eta_hat = opt$par[1],
      r = pcf_obs$r,
      g_obs = pcf_obs$g,
      g_fit = g_fit,
      pcf_obs = pcf_obs
    ),
    class = "mc_pcf_fit"
  )
}

```

**Basic structure**

The fitting routine proceeds as follows:

1. Compute an empirical pair correlation function using estimate_pcf_for_mc().
2. Estimate the intensity $\lambda$ from the observed point pattern using spatstat.geom::intensity(X).
3. Convert the user-supplied starting value for $\phi$ into the unconstrained parameter $\eta$.
4. Minimize the minimum contrast objective using optim().
5. Back-transform the optimized parameters to the original scale and return both 
the fitted parameters and fitted PCF values.

Note $\lambda$ is *not* optimized within the minimum contrast step.

The objective function is:

$$
\sum_i |\hat{g}(r_i)^k - g(r_i; \theta)^k|^p,
$$

where $\hat{g}(r_i)$ is the empirical PCF, $g(r_i; \theta)$ is the theoretical 
PCF under the model, and $k$ and $p$ are tuning constants.

**Reparameterization from $\phi$ to $\eta$**

In the case of the CSCP, the parameter space of $\phi$ is $[0, 2]$, 
and of $s$ is $[0, \infty)$.

To enforce $\phi \in [0, 2]$, we optimize over $\eta$, where $\phi = 2\cdot \text{plogis}(\eta)$.

To enforce $s \in [0, \infty)$, we optimize over $\log(s)$.

**Default values**

I chose the following as default values, with my justification being they either
produced fairly decent fits in my limited experience, or because they were 
suggested by my GOATs (supervisors).

| Argument      | Default        | Role                              |
|:--------------|:---------------|:----------------------------------|
| `bw`          | `0.2`          | Bandwidth for empirical PCF estimation |
| `correction`  | `"isotropic"`  | Edge correction used in `pcf()`   |
| `divisor`     | `"a"`          | Denominator choice for `pcf()`    |
| `zerocor`     | `"JonesFoster"`| Correction near the origin        |
| `r_min`       | `0.02`         | Lower cutoff for fitting range    | 
| `r_max`       | `NULL`         | Upper cutoff for fitting range    |
| `k`           | `0.25`         | Power transform in minimum contrast |
| `p`           | `2`            | Loss power                        |
| `start$phi`   | `0.5`          | Starting value for clustering strength |
| `start$s`     | `0.05`         | Starting value for range/scale    |

I also understand the argument for using whatever `spatstat` uses as its defaults.

**Criticsms**

- It's a pain in the ass to plot the fitted vs empirical right now. Would be nice 
to be able to streamline this.

- Also, The way I have it setup right now is not very flexible if we want to consider
different models in the future. In particular, a better design pattern might have
been to have users pass there own objective function, rather than the way I have
`mc_pcf_objective()` used internally 
(`spatstat` has the users pass it to `mincontrast()`).

- I also think `spatstat` fits minimum contrast by using the K-function by default? 
Maybe this would be better.

- I don't really know the theory behind the $k$ and $p$ tuning parameters.

## Comparison to `spatstat` implementation

```{r message = FALSE, warning = FALSE, echo = FALSE}
set.seed(1)

fit_cscp_mc_spatstat <- function(X,
                                 startpar = c(phi = 0.5, s = 0.05),
                                 df = 1,
                                 bw = 0.2,
                                 correction = "isotropic",
                                 divisor = "a",
                                 zerocor = "JonesFoster",
                                 r_min = 0.0001,
                                 r_max = NULL,
                                 k = 0.25) {
  stopifnot(inherits(X, "ppp"))

  lambda_hat <- spatstat.geom::intensity(X)

  obs <- spatstat.explore::pcf(
    X,
    bw = bw,
    correction = correction,
    divisor = divisor,
    zerocor = zerocor
  )

  y_name <- switch(
    correction,
    isotropic = "iso",
    translate = "trans",
    Ripley = "iso",
    "iso"
  )

  if (!y_name %in% names(obs)) {
    y_name <- intersect(c("iso", "trans", "un"), names(obs))[1]
  }

  keep <- is.finite(obs$r) & is.finite(obs[[y_name]]) & obs$r > r_min
  if (!is.null(r_max)) keep <- keep & obs$r <= r_max

  obs_sub <- obs[keep, ]

  # Theoretical PCF for mincontrast
  theo_fun <- function(par, r) {
    pcf_cscp_theoretical(
      r = r,
      lambda = lambda_hat,
      phi = par["phi"],
      s = par["s"],
      df = df
    )
  }

  fit <- spatstat.model::mincontrast(
    obs = obs_sub,
    theoretical = theo_fun,
    startpar = startpar,
    ctrl = list(parscale = c(phi = 0.25, s = 0.05))
  )

  pars <- c(
    lambda = lambda_hat,
    phi = fit$estimate["phi"],
    s = fit$estimate["s"]
  )

  list(
    model = "CSCP",
    par = pars,
    fit = fit,
    obs = obs_sub,
    correction_used = y_name
  )
}

scenarios <- list(
  list(lambda = 300, phi = 1, s = 0.05),
  list(lambda = 300, phi = 2, s = 0.03),
  list(lambda = 300, phi = 1, s = 0.08),
  list(lambda = 300, phi = 2, s = 0.08)
)

old_par <- par(no.readonly = TRUE)
par(mfrow = c(2, 2), mar = c(3.5, 3.5, 2.2, 0.8), mgp = c(2.1, 0.7, 0))

r_plot <- seq(0, 1, by = 0.001)

for (sc in scenarios) {
  # Simulate one dataset
  X <- sim_cscp(lambda = sc$lambda, phi = sc$phi, s = sc$s)$pp

  # Fit using your implementation
  fit_own <- fit_cscp_mc(
    X,
    start = list(phi = 0.5, s = 0.05),
    bw = 0.01,
    r_min = 0.01
  )

  # Fit using spatstat::mincontrast()
  fit_spat <- fit_cscp_mc_spatstat(
    X,
    startpar = c(phi = 0.5, s = 0.05),
    bw = 0.01,
    r_min = 0.01
  )

  # Empirical PCF
  emp <- pcf(
    X,
    correction = "isotropic",
    divisor = "a",
    zerocor = "JonesFoster",
    bw = 0.01
  )

  # Curves
  g_true <- pcf_cscp_theoretical(
    r_plot,
    lambda = sc$lambda,
    phi = sc$phi,
    s = sc$s
  )

  g_own <- pcf_cscp_theoretical(
    r_plot,
    lambda = fit_own$par[["lambda"]],
    phi = fit_own$par[["phi"]],
    s = fit_own$par[["s"]]
  )

  g_spat <- pcf_cscp_theoretical(
    r_plot,
    lambda = fit_spat$par[["lambda"]],
    phi = fit_spat$fit$par[["phi"]],
    s = fit_spat$fit$par[["s"]]
  )

  ymax <- max(
    emp$iso[is.finite(emp$iso)],
    g_true, g_own, g_spat,
    na.rm = TRUE
  )

  plot(
    emp,
    . ~ r,
    ylim = c(0.8, ymax),
    legend = FALSE,
    main = bquote(phi == .(sc$phi) * "," ~~ s == .(sc$s)),
    xlab = "r",
    ylab = "g(r)"
  )

  lines(r_plot, g_true, lwd = 2, col = "forestgreen")
  lines(r_plot, g_own,  lwd = 2, col = "red")
  lines(r_plot, g_spat, lwd = 2, col = "blue", lty = 2)
}

legend(
  "topright",
  legend = c("Empirical", "True", "Own MC", "spatstat MC"),
  col = c("black", "forestgreen", "red", "blue"),
  lwd = c(1, 2, 2, 2),
  lty = c(1, 1, 1, 2),
  bty = "n"
)

par(old_par)
```

THANK frick they are (almost) identical. I can rest in peace.

## Todo

0. Add inline & roxygen comments.

1. I hate how annoying it is to plot the fitted PCF vs observed. Will be 
implementing a class + summary / plotting methods. 

2. It would be nice if we could calculate the empirical PCF using `spatstat`'s
PCF function, and then just pass that PCF object to the minimum contrast fitting
function we have. Makes comparing different fits easier, and feels more reliable.

3. I have no idea above uncertainty quantification for MC fitting. Will look into
this.

4. I believe the reparameterization of $\phi$ to $\eta$ does not allow estimates 
of $\phi \in \{0, 1\}$. Is this a problem?
